---
title: "cov_visualizations"
output: html_document
date: '2023-12-05'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r, echo=FALSE}
library(ggplot2)
library(ggridges)
library(viridis)
library(plyr)
library(tidyr)
library(reshape2)
library(dplyr)
library(cowplot)
library(ggpubr)
library(purrr)
```

## Basic Functions

```{r}
three_to_single <- function(three_letter) {
  aa_map <- list(
    ALA = "A",
    ARG = "R",
    ASN = "N",
    ASP = "D",
    CYS = "C",
    GLN = "Q",
    GLU = "E",
    GLY = "G",
    HIS = "H",
    ILE = "I",
    LEU = "L",
    LYS = "K",
    MET = "M",
    PHE = "F",
    PRO = "P",
    SER = "S",
    THR = "T",
    TRP = "W",
    TYR = "Y",
    VAL = "V"
    # You can add more amino acids if needed
  )
  
  return(aa_map[[three_letter]])
}

aa_alphabet = c("A","R","N","D","C",
                "Q","E","G","H","I",
                "L","K","M","F","P",
                "S","T","W","Y","V")

get_seqs = function(file_name){
  fasta_file <- file_name
  
  # Read the lines of the FASTA file
  fasta_lines <- readLines(fasta_file)
  
  # Extract the sequence information
  sequence_info <- grep("^>", fasta_lines, value = TRUE)
sequence_name <- gsub(">", "", sequence_info)
sequence <- sapply(seq_along(sequence_info), function(i) {
  start <- grep(sequence_info[i], fasta_lines)[1] + 1
  end <- ifelse(i == length(sequence_info), length(fasta_lines), grep(sequence_info[i + 1], fasta_lines)[1] - 1)
  paste(fasta_lines[start:end], collapse = "")
}, simplify = TRUE)

# Create a named vector with one entry
fasta_data <- setNames(sequence, sequence_name)
  return(fasta_data)
}

zscore = function(df = c()){(df-mean(df))/sd(df)}

combine_df_Starr <- function(focal_df) {
  setwd("/Users/beallma/OneDrive - Emory University/Documents/PIPP/SARS_NLP")
  fname <- "semantic_change_loc_data_DMS.csv"
  
  df_sem_Starr <- read.delim(file = fname, header = TRUE, sep = ",", quote = "\"",
                             dec = ".", fill = TRUE, comment.char = "#")
  df_sem_Starr <- df_sem_Starr %>%
    select(ID, STARR_DMS, STARR_ESC)
  
  focal_df <- focal_df %>% 
    left_join(df_sem_Starr, by = c('Mutations' = 'ID'))
  
  focal_df <- focal_df %>%
    mutate(STARR_scan = case_when(STARR_DMS == "True" ~ ""))
  
  focal_df <- focal_df %>%
    mutate(STARR_scan = case_when(STARR_ESC == "True" ~ "escape",
                                  STARR_ESC == "False" ~ "non-escape",
                                  TRUE ~ "not tested"))
  return(focal_df)
}
```

#SARS-CoV-2 experimental data

These data will define antibody escape mutations

```{r}
# STARR DMS dataset
# Tyler N. Starr et al., Prospective mapping of viral mutations that escape antibodies used to treat COVID-19. Science 371,850-854(2021). DOI:10.1126/science.abf9302
#setwd("/Users/beallma/OneDrive - Emory University/Documents/PIPP/SARS_NLP")
setwd("/Users/beallma/Documents/GitHub/NLP_viral_escape/data/cov")
cov_seq = get_seqs("cov2_spike_wt.fasta")[[1]]

setwd("/Users/beallma/Documents/GitHub/NLP_viral_escape/data/cov")

fname = "starr_dms/Starr_REGN_and_LY-CoV016_raw_data.csv"
starr_escape_fracs = read.delim(file = fname, header = TRUE, sep=",",quote = "\"",
                 dec = ".", fill = TRUE, comment.char = "#")
starr_escape_class = starr_escape_fracs %>%
  mutate(
      Starr_esc = case_when(mut_escape >= 0.5 ~ "escape",
  mut_escape < 0.5 ~ "non-escape")
  ) %>%
  mutate(ID = paste(wildtype,site, mutation, sep = ""))

#lets only use the mutations tested under CB6 (LY-CoV016) conditions
#this is ab used in the escape predictions for FoldX and PyRosetta
starr_escape_class = starr_escape_class %>% 
  filter(condition == "LY-CoV016") %>%
  mutate(mut = paste(site, mutation, sep = "")) %>%
  select(ID, site, mut, Starr_esc) %>%
  mutate(wt = paste0(substring(ID, 2, 4), substring(ID, 1, 1)))

#note that they did not test all sites in Starr. Just the RBD sites 331 to 531
#some of the sites tested were inviable so did not have escape measures. Let's figure out which ones
RBD_sites_tested = c(331:531)
all_RBD_IDs = expand.grid(RBD_sites_tested = RBD_sites_tested, aa_alphabet = aa_alphabet)
all_RBD_IDs = data.frame(mutations = 
                           apply(all_RBD_IDs, 1, function(x) paste(x, collapse = "")))
all_wt_mut = unique(starr_escape_class$wt)
wt_dict = setNames(as.list(starr_escape_class$wt), as.character(starr_escape_class$site))
wt_dict = setNames(as.list(starr_escape_class$wt), as.character(starr_escape_class$site))
cov_seq_vec = unlist(strsplit(cov_seq, NULL))
all_site_idx = c(1:length(cov_seq_vec))
all_mutations = expand.grid(mutations = aa_alphabet, idx = all_site_idx)
all_mutations = all_mutations %>% 
  mutate(wt_bool = ifelse(cov_seq_vec[idx] == mutations, TRUE, FALSE))%>%
  mutate(mutations = paste0(idx, mutations))  
wt_mutations = all_mutations %>% 
  filter(wt_bool)

starr_escape_viable = left_join(all_RBD_IDs, 
                                starr_escape_class, 
                                by = c("mutations" = "mut")) %>%
  filter(!mutations %in% all_wt_mut) %>%
  mutate(site = as.integer(substring(mutations, 1,3))) %>%
  mutate(wt = sapply(site, function(x) wt_mutations$mutations[x])) %>%
  mutate(ID = paste0(substring(wt, 4, 4), mutations)) %>%
  mutate(Starr_esc = ifelse(is.na(Starr_esc), "not viable", Starr_esc))
#Now we have a dataframe that describes all of the sites that were tested in
#the Starr DMS experiments. They can confer escape or not if they are viable,
#otherwise they are listed as not viable

# now let us include mutations that were not tested at all
all_sites_escape_viable = left_join(subset(all_mutations, !wt_bool),
                                starr_escape_viable, 
                                by = "mutations") %>% 
  mutate(site = as.integer(ifelse(is.na(ID), 
                                  substr(mutations, 1,nchar(mutations)-1), site))) %>%
  mutate(wt = ifelse(is.na(wt), 
                     sapply(site, function(x) wt_mutations$mutations[x]),
                     wt)) %>%
  mutate(ID = ifelse(is.na(ID),
                     paste0(substr(wt, nchar(wt), nchar(wt)), mutations),
                     ID)) %>%
  mutate(Starr_esc = ifelse(is.na(Starr_esc), "not tested", Starr_esc)) %>%
  select(-idx, -wt_bool) 

all_sites_escape_viable
#mutate(wt = as.character(
    #sapply(site, function(x) unlist(wt_dict[[as.character(x)]])[1]
    #       )))

# Baum escape mutations
non_escape = c("Q321L", "V341I", "A348T", "N354D", 
               "S359N", "V367F", "K378R", "R408I", 
               "Q409E", "A435S", "K458R", "I472V", 
               "G476S", "V483A", "Y508H", "H519P")
escape = c("K417E", "K444Q", "V445A", "N450D", 
           "Y453F", "L455F", "E484K", "G485D", 
           "F486V", "F490L", "F490S","Q493K")

all_sites_escape_viable = all_sites_escape_viable %>%
  mutate(Baum_esc = 
           case_when(ID %in% escape ~ "escape",
                     ID %in% non_escape ~ "non-escape",
                     TRUE ~ "not tested"))

subset(all_sites_escape_viable, site == 490)
subset(all_sites_escape_viable, Baum_esc == "non-escape")

```

#Hie Bespoke Model
Read in and process the Hie pretrained bespoke model data

```{r}
#fname = "cov/Baum_pretrain_screen_all_sem_gram.txt"
#fname = "/Users/beallma/Documents/GitHub/NLP_viral_escape/language_models/results/cov/cov_mean_embed_Hie.log"
fname = "/Users/beallma/Documents/GitHub/NLP_viral_escape/language_models/results/cov/cov_mean_embed_Hie_081124.log"

df_Hie_Baum = read.delim(file = fname, header = TRUE, sep="",quote = "\"",
                 dec = ".", fill = TRUE, comment.char = "#")
df_Hie_Baum = df_Hie_Baum %>% dplyr::rename("Semantic.change_Hie" = "Semantic.change",
               "Grammaticality_Hie_logit" = "Grammaticality") %>%
  # mutate_at(c(
  #     'Grammaticality_Hie_logit'
  #   ), rank) %>% mutate_at(c(
  #     'Semantic.change_Hie'
  #   ), rank)%>% 
  # rename("Grammaticality_Hie_logit_rank" = Grammaticality_Hie_logit,
  #        "Semantic.change_Hie_rank" = Semantic.change_Hie
  #        ) %>%
  # mutate(
  #     Grammaticality_Hie_logit_rank = max(Grammaticality_Hie_logit_rank) - Grammaticality_Hie_logit_rank + 1
  # ) %>%
  select(-Mutations)
    
```

Read in and process the 
##ESM-2 data

```{r}
#fname = "cov/Baum_pretrain_screen_ESM2_logit_odds.txt"
fname = "/Users/beallma/Documents/GitHub/NLP_viral_escape/language_models/results/cov/cov_mean_embed_ESM2_081324.log"
df_ESM2_Baum = read.delim(file = fname, header = TRUE, sep="",quote = "\"",
                 dec = ".", fill = TRUE, comment.char = "#")
df_ESM2_Baum = df_ESM2_Baum %>% dplyr::rename("Semantic.change_ESM2" = "Semantic.change",
                "Grammaticality_ESM2_unmasked" = "Grammaticality") %>% 
  # mutate_at(c(
  #     'Grammaticality_ESM2_logit'
  #   ), rank) %>% mutate_at(c(
  #     'Semantic.change_ESM2'
  #   ), rank)%>% 
  # rename("Grammaticality_ESM2_logit_rank" = Grammaticality_ESM2_logit,
  #        "Semantic.change_ESM2_rank" = Semantic.change_ESM2
  #        ) %>%
  # mutate(
  #     Grammaticality_ESM2_logit_rank = max(Grammaticality_ESM2_logit_rank) - Grammaticality_ESM2_logit_rank + 1
  # ) %>%
  select(-Mutations, -Sequence)
#high grammar means more like the original sequence, so we want to invert^ this to match semantic change

```

```{r}
RBD_sites_tested = c(331:531)
all_RBD_IDs = expand.grid(RBD_sites_tested = RBD_sites_tested, aa_alphabet = aa_alphabet)
all_RBD_IDs = data.frame(mutations = 
                           apply(all_RBD_IDs, 1, function(x) paste(x, collapse = "")))
all_wt_mut = unique(starr_escape_class$wt)
wt_dict = setNames(as.list(starr_escape_class$wt), as.character(starr_escape_class$site))
wt_dict = setNames(as.list(starr_escape_class$wt), as.character(starr_escape_class$site))
cov_seq_vec = unlist(strsplit(cov_seq, NULL))
all_site_idx = c(1:length(cov_seq_vec))
all_mutations = expand.grid(mutations = aa_alphabet, idx = all_site_idx)
all_mutations = all_mutations %>% 
  mutate(wt_bool = ifelse(cov_seq_vec[idx] == mutations, TRUE, FALSE))%>%
  mutate(mutations = paste0(idx, mutations))  
wt_mutations = all_mutations %>% 
  filter(wt_bool)
```

Read in and process the 
##ESM-2 masked data

```{r}
fname = "cov/ESM_probabilities_cov_wt.fasta.csv"
df_ESM2_cov = read.delim(file = fname, header = TRUE, sep=",",quote = "\"",
                 dec = ".", fill = TRUE, comment.char = "#")

df_ESM2_cov <- df_ESM2_cov %>%
  pivot_longer(cols = -X,
               names_to = "Amino_Acid",
               values_to = "prob_ESM2_masked")

wt_mutations = wt_mutations %>% 
  mutate(AA = paste0(substr(mutations, nchar(mutations), nchar(mutations)), idx))

df_ESM2_cov <- df_ESM2_cov %>%
  mutate(partial_id = paste0(X, Amino_Acid))

df_ESM2_cov <- df_ESM2_cov %>%
  left_join(wt_mutations, by = 
               c("X" = "idx")) %>%
  select(-wt_bool) %>%
  mutate(
    ID = paste0(AA, Amino_Acid),
    mutant = case_when(substr(ID, nchar(ID), nchar(ID)) == substring(ID, 1, 1) ~ "wt",
                   TRUE ~ "mutant")) %>%
  select(-X, -Amino_Acid, -partial_id, -AA, -mutations)
df_ESM2_cov
```

Read in and process the 
##MutComputeX data

```{r}
fname = "cov/cov2-spike-wt_EnsResNet_0222.csv"
df_mutcomp = read.delim(file = fname, header = TRUE, sep=",",quote = "\"",
                 dec = ".", fill = TRUE, comment.char = "#")
names(df_mutcomp)
head(df_mutcomp)
df_mutcomp <- df_mutcomp %>%
  rename(prAA = "mut_pred",
         pred_prob = "mut_pred_prob") %>%
  # rename(mut_pred = "prAA",
  #        mut_pred_prob = "pred_prob") %>%
  pivot_longer(cols = starts_with("pr"),
               names_to = "Amino_Acid",
               values_to = "prob_MutCompX") %>%
  select(chain_id, pos, wtAA, mut_pred, Amino_Acid, prob_MutCompX) %>%
  mutate(
    Amino_Acid = gsub("pr", "", Amino_Acid),
    Amino_Acid = sapply(Amino_Acid, three_to_single),
    wtAA = sapply(wtAA, three_to_single),
    mut_pred = sapply(mut_pred, three_to_single),
    Mutations = paste(wtAA, pos, Amino_Acid, sep = "")
  ) %>% 
  filter(chain_id == "A") %>%
  select(Mutations,prob_MutCompX)
#high probability means more like the original sequence, so we want to invert^ this to match semantic change
df_mutcomp
```

Read in and process the 
##MutRank data

```{r}
fname = "cov/cov2_mutrank_0524_uncertainty.csv"
df_mutrank = read.delim(file = fname, header = TRUE, sep=",",quote = "\"",
                 dec = ".", fill = TRUE, comment.char = "#")
names(df_mutrank)
head(df_mutrank)
df_mutrank <- df_mutrank %>%
  # rename(prAA = "mut_pred",
  #        pred_prob = "mut_pred_prob") %>%
  # rename(mut_pred = "pred_AA",
  #        mut_pred_prob = "pred_prob") %>%
  rename(pred_AA = "mut_pred",
         pred_prob = "mut_pred_prob") %>%
  pivot_longer(cols = starts_with("mean"),
               names_to = "Amino_Acid",
               values_to = "prob_MutRank") %>%
  select(chain_id, res_seq_num, wt_AA, mut_pred, Amino_Acid, prob_MutRank) %>%
  mutate(
    Amino_Acid = gsub("mean_", "", Amino_Acid),
    Amino_Acid = sapply(Amino_Acid, three_to_single),
    wt_AA = sapply(wt_AA, three_to_single),
    mut_pred = sapply(mut_pred, three_to_single),
    Mutations = paste(wt_AA, res_seq_num, Amino_Acid, sep = "")
  ) %>% 
  filter(chain_id == "A") %>%
  select(Mutations,prob_MutRank)
#high probability means more like the original sequence, so we want to invert^ this to match semantic change
df_mutrank
```

Read in and process the 
##StabilityOracle data

```{r}
#fname = "cov/cov2-spike-wt_stability_oracle_0423_uncertainty.csv"
fname = "cov/cov2_stability-oracle_0524_uncertainty.csv"

df_stab_ora = read.delim(file = fname, header = TRUE, sep=",",quote = "\"",
                 dec = ".", fill = TRUE, comment.char = "#")

df_stab_ora <- df_stab_ora %>%
  pivot_longer(cols = starts_with("mean_"),
               names_to = "Amino_Acid",
               values_to = "ddG_Stab_Ora") %>%
  select(chain_id, res_seq_num, wt_AA, pred_AA, Amino_Acid, ddG_Stab_Ora) %>%
  mutate(
    Amino_Acid = gsub("mean_", "", Amino_Acid),
    Amino_Acid = sapply(Amino_Acid, three_to_single),
    wt_AA = sapply(wt_AA, three_to_single),
    pred_AA = sapply(pred_AA, three_to_single),
    Mutations = paste(wt_AA, res_seq_num, Amino_Acid, sep = "")
  ) %>% 
  filter(chain_id == "A") %>%
 select(Mutations,ddG_Stab_Ora)
df_stab_ora
```

## EVEscape

```{r}
setwd("/Users/beallma/Documents/GitHub/NLP_viral_escape/data/cov/evescape")
#fname = "Wuhan(A)_single_mutation_predictions.csv"
fname = "full_Starr_evescape.csv"

df_eves = read.delim(file = fname, header = TRUE, sep=",",quote = "\"",
                 dec = ".", fill = TRUE, comment.char = "#") %>%
  arrange(i) %>%
  mutate(ID = paste0(wt, i, mut)) %>%
  select(ID, evescape, fitness_eve) %>%
  rename("fitness_eve" = "eve")
  #select(mutations, EVEscape)
df_eves 
```

```{r}
ggplot(df_eves, aes(x = eve, y = (evescape))) +
  #geom_line() +
  geom_point() +
  labs(x = "EVE", y = "EVEscape", title = "") + 
  theme_classic()
tes = cor.test(df_eves$eve, df_eves$evescape, method="pearson")
tes
```

## Combine all datastreams into one dataframe

```{r}

all_sites_escape_viable # contains the Starr DMS scan dataset. All sites in the RBD were tested
                        # we annotate sites that are not viable as those that did not have an escape
                        # measurement within the RBD loci. 
                        # escape mutations are those with a measured escape fraction above 0.5
                        # All other sites are "not tested"
df_Hie_Baum # contains the grammaticality and semantic change measures from the Hie model
           # large gramm -> closer to wt -> large rank
           # small sem change -> close to wt -> supposed less likely escape -> smaller rank
df_stab_ora # ddG from stability Oracle. More negative -> more stable -> smaller rank; more positive -> less stable -> larger rank
df_mutcomp # large probability -> more stable -> larger rank
df_mutrank # large log odds --> more stable --> larger rank
df_ESM2_cov # probability from masked ESM model -> larger probability -> more grammatical -> larger rank
df_ESM2_Baum # semantic change and grammar using the ESM model in an unmasked framework

comb_data = df_Hie_Baum  %>% 
  left_join(all_sites_escape_viable, by = c('ID' = 'ID')) %>%
  left_join(df_ESM2_cov, by = c('ID' = 'ID')) %>%
  left_join(df_ESM2_Baum, by = c('ID' = 'ID')) %>%
  left_join(df_stab_ora, by = c('ID' = 'Mutations')) %>%
  left_join(df_mutcomp, by = c('ID' = 'Mutations')) %>%
  left_join(df_mutrank, by = c('ID' = 'Mutations')) %>%
  left_join(df_eves, by = c("ID" = "ID")) %>%
  replace_na(list(Starr_esc = "not tested", Baum_esc = "not tested"))

my_measures = colnames(comb_data %>% 
                         select(-c(ID, mutations, site, 
                                Starr_esc, wt, Baum_esc,
                                mutant, Sequence))
                       )
my_measures
comb_data

rev_rank_measures = c("ddG_Stab_Ora")

comb_data_norm = comb_data %>% 
  mutate_at(my_measures, function(x) rank(x, na.last = "keep")) %>%
  mutate_at(rev_rank_measures, ~max(rank(., na.last = "keep"), na.rm=T) + 1 - rank(.,na.last = "keep") ) %>%
  mutate_at(my_measures, function(x) x/max(x, na.rm = TRUE))

comb_data_norm

mut_by_site = c("ddG_Stab_Ora", "prob_MutCompX", "prob_ESM2_masked", "prob_MutRank")

comb_data_mut_by_site = comb_data %>%
  group_by(site) %>%
  mutate_at(mut_by_site,
            function(x) rank(x, na.last = "keep")) %>%
  mutate_at(mut_by_site, function(x) rank(x, na.last = "keep")) %>%
  select(ID, mutant, site, ddG_Stab_Ora, prob_MutCompX, prob_ESM2_masked, evescape)
  


```

```{r}
to_save = comb_data %>% select(ID, Sequence, Starr_esc)

write.csv(to_save, "cov_sequence_escape.csv", row.names = FALSE, sep = ",")
```

# Pie Chart
```{r}
library(ggforce)
count_comb_cov = comb_data %>% group_by(Starr_esc) %>%
  summarize(
    count = n()
  )


mut_order = c("escape", "non-escape", 
              "not viable","not tested")
pie_data_cov <- count_comb_cov %>%
  arrange(match(Starr_esc, mut_order)) %>% # sort so pie slices end up sorted
  mutate(
    end_angle = 2*pi*cumsum(count)/sum(count),   # ending angle for each pie slice
    start_angle = lag(end_angle, default = 0),   # starting angle for each pie slice
    mid_angle = 0.5*(start_angle + end_angle),   # middle of each pie slice, for text labels
    # horizontal and vertical justifications for outer labels
    hjust = ifelse(mid_angle > pi, 1, 0),
    vjust = ifelse(mid_angle < pi/2 | mid_angle > 3*pi/2, 0, 1)
  )
pie_data_cov
pie_data_cov$hjust[1] = 1 # to get the labels to not overlap since escape makes up such a small fraction of the circle
pie_data_cov$vjust[4] = 1
pie_data_cov

cov_esc_pie = ggplot(pie_data_cov) +
    aes(
      x0 = 0, y0 = 0,
      r0 = 0.4, r = 1,
      amount = count,
      start = start_angle, end = end_angle,
      fill = Starr_esc
    ) +
  geom_arc_bar(alpha = 0.7) +
  geom_text(
    aes(x = 1.1 * sin(mid_angle), 
        y = 1.1 * cos(mid_angle), 
        label = count,
        hjust = hjust,
        vjust = vjust)#vjust)
  ) +
  scale_fill_manual(values = c("escape" = "#5D3A9B",
                                 "non-escape" = "#E66100", 
                                 "not viable" = "#AF596E",
                                 "not tested" = "#C5C5C5"),
                    name = "spike\nmutation type",
                    breaks = mut_order) +
  coord_fixed() +
  xlim(-1.0, 1.0) +
  ylim(-1.1, 1.4) +
  theme_void() +
  theme(
    strip.text = element_blank(),
    strip.background = element_blank(),
    legend.key=element_rect(color="white")
  ) + 
  geom_segment(aes(
      x = 1 * sin(mid_angle), 
      xend = 1.1 * sin(mid_angle),
      y = 1 * cos(mid_angle), 
      yend = 1.1 * cos(mid_angle)
      #y = as.numeric(factor(Measure)) + y_delta, yend = as.numeric(factor(Measure)) - y_delta #midpoint +/- y_delta
    ))
cov_esc_pie

fname = paste0('cov_Starr_DMS_piechart_', Sys.Date(), '.png')
ggsave(
  file = fname, 
  plot = cov_esc_pie,
  width = 4, height = 3.5, units = "in"
)

mut_order = c("escape", "non-escape", 
              "not viable","not tested")
#layer_data(plot = cov_esc_pie)
```

#Make a Long Dataframe and partition by viability

```{r}

comb_data_long = comb_data_norm %>% 
  select(-c(wt, mutations, site)) %>%
  pivot_longer(cols = my_measures,
               names_to = "Measure",
               values_to = "rank") %>%
  mutate(Measure = 
           case_when(
             Measure == "ddG_Stab_Ora" ~ "Stability Oracle",
             Measure == "prob_MutCompX" ~ "MutComputeX",
             Measure == "Grammaticality_ESM2_unmasked" ~ "Grammaticality (ESM2, unmasked)",
             Measure == "Semantic.change_ESM2" ~ "Semantic change (ESM2)",
             Measure == "Semantic.change_Hie" ~ "Semantic change (Hie)",
             Measure == "Grammaticality_Hie_logit" ~ "Grammaticality (Hie)",
             Measure == "prob_ESM2_masked" ~ "Grammaticality (ESM2, masked)",
             Measure == "prob_MutRank" ~ "MutRank",
             Measure == "evescape" ~ "EVEscape",
             Measure == "eve" ~ "EVE"
             )
         )
my_measures_rename = colnames(comb_data_long %>% 
                         select(-c(ID, 
                                Starr_esc, Baum_esc,
                                rank)))

my_comparisons_gramm <- list( c("not tested", "viable"), 
                          c("not tested", "not viable"), 
                          c("viable", "not viable") )

comb_data_long_gramm = comb_data_long %>% 
  mutate(viability = case_when(Starr_esc == "not tested" ~ "not tested",
                               Starr_esc == "non-escape" ~ "viable",
                               Starr_esc == "escape" ~ "viable",
                               Starr_esc == "not viable" ~ "not viable"))
comb_data_long_gramm
```

##Viability Barplot

```{r}
nested_comb_data_long_gramm = comb_data_long_gramm |>
  pivot_wider(names_from = "viability", values_from = "rank") |>
  select(-c(mutant, Starr_esc, Baum_esc, Sequence))

comb_data_long_gramm %>% select(-c(ID, mutant, Starr_esc, Baum_esc, rank)) %>% group_by_all() %>%
    summarize(
        count = n()
    ) %>%
  pivot_wider(names_from = "viability", values_from = "count")

x_order = rev(c("Grammaticality (Hie)",
            "Grammaticality (ESM2, unmasked)",
            "Grammaticality (ESM2, masked)", 
            "MutComputeX",
            "MutRank",
            "Stability Oracle",
            "EVEscape",
            "EVE",
            "Semantic change (Hie)",
            "Semantic change (ESM2)"
         ))
#remove where the measure has no values (all NAs because the site+mutation was not tested with that metric)

#nested_comb_data_long_gramm = nested_comb_data_long_gramm[rowSums(is.na(nested_comb_data_long_gramm)) < 3,]
nested_comb_data_long_gramm_temp = nested_comb_data_long_gramm %>%
   subset(Measure != "FoldX") %>% subset(Measure != "PyRosetta") %>% nest(data = -Measure)

#nested_comb_data_long_gramm = nested_comb_data_long_gramm %>% mutate(factor(Measure, levels = x_order))

data_stats <- nested_comb_data_long_gramm %>% select(-ID) |>
  nest(data = -Measure) |> subset(Measure != "FoldX") %>% subset(Measure != "PyRosetta") %>%
  mutate(
    via_non_via = map(data, ~broom::tidy(wilcox.test(.x$viable, .x$`not viable`))),
    via_not_tes = map(data, ~broom::tidy(wilcox.test(.x$viable, .x$`not tested`))),
    non_via_not_tes = map(data, ~broom::tidy(wilcox.test(.x$`not viable`, .x$`not tested`))),
    Measure = factor(Measure, levels = x_order)
  ) |>
  pivot_longer(
    cols = c(via_non_via, via_not_tes, non_via_not_tes),
    names_to = "test",
    values_to = "stats"
  ) |>
  mutate(
    column1 = case_when(
      test == "via_non_via" ~ "viable",
      test == "via_not_tes" ~ "viable",
      test == "non_via_not_tes" ~ "not tested"
    ),
    column2 = case_when(
      test == "via_non_via" ~ "not viable",
      test == "via_not_tes" ~ "not tested",
      test == "non_via_not_tes" ~ "not viable"
    )
  ) |>
  unnest_wider(stats) |>
  select(Measure, column1, column2, p.value)

y_delta = .25 # distance between adjacent bloxplots (.8/number of boxplots)
x_delta = .075 # distance between line segments
data_comparisons <- data_stats |>
  mutate(
    # apply overall false-discovery rate correction
    p.value = p.adjust(p.value, method = "fdr"),
    # set stars
    label = case_when(
      p.value < 0.001 ~ "***",
      p.value < 0.01 ~ "**",
      p.value < 0.05 ~ "*",
      TRUE ~ "NS"
    ),
    # offset to correct for stars not being vertically centered
    label_off = case_when(
      label == "NS" ~ 0,
      TRUE ~ -.04,
    ),
    # calculate x and y offsets for line segments
    yoff1 = case_when(
      column1 == "viable" ~ y_delta,
      column1 == "not tested" ~ -y_delta,
      TRUE ~ 0
    ),
    yoff2 = case_when(
      column2 == "viable" ~ y_delta,
      column2 == "not tested" ~ -y_delta,
      TRUE ~ 0
    ),
    xoff = case_when(
      column1 == "viable" & column2 == "not viable" ~ 0,
      column1 == "not tested" & column2 == "not viable" ~ x_delta,
      TRUE ~ 2 * x_delta
    )
  )

custom_order_via = c("viable", "not viable", "not tested")

#data_comparisons = data_comparisons %>% arrange(factor(Measure, levels = x_order))

my_plot = nested_comb_data_long_gramm %>% 
  select(-ID) |> subset(Measure != "FoldX") %>% subset(Measure != "PyRosetta") %>%
  pivot_longer(-Measure, names_to = "type", values_to = "value") %>% drop_na(value) |>
  ggplot(aes(x = value, y = factor(Measure, levels = x_order), fill = type)) +
  #ggplot(aes(x = value, y = Measure, fill = type)) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_manual(values = c("viable" = "#399c52",
                                 "not viable" = '#AF596E', #"#cd6a83", 
                                 "not tested" = "#C5C5C5"),
                      breaks = custom_order_via,
                      name = "spike mutation type"
                    ) +
  geom_segment( # line segments for significance
    data = data_comparisons,
    aes(
      x = 1.0 + xoff + .03, 
      xend = 1.0 + xoff + .03,
      y = as.numeric(factor(Measure, levels = x_order)) + yoff1, 
      yend = as.numeric(factor(Measure, levels = x_order)) + yoff2
      #y = as.numeric(factor(Measure)) + y_delta, yend = as.numeric(factor(Measure)) - y_delta #midpoint +/- y_delta
    ),
    inherit.aes = FALSE
  ) +
  geom_text(
    data = data_comparisons,
    aes(
      x = 1 + xoff +.04,
      y = as.numeric(factor(Measure, levels = x_order)) + (yoff1 + yoff2) / 2 + label_off,
      label = label,
    ),
    size = 10 / .pt, # 10 point font size
    vjust = 0.5,
    hjust = 0,
    inherit.aes = FALSE
  ) +
  xlim(-.04, 1.25) +
  scale_x_continuous(name = "relative rank", breaks =c(0,.5,1)) +
  scale_y_discrete(labels = c("Grammaticality (ESM2, unmasked)" = "Grammaticality\n(ESM2, unmasked)",
                              "Grammaticality (ESM2, masked)" = "Grammaticality\n(ESM2, masked)",
                              "Grammaticality (Hie)" = "Grammaticality\n(Hie)")) +
  theme_classic() + ylab("") + theme(legend.position = "top")
my_plot

fname = paste0('cov_all_measure_boxplot_viability_', Sys.Date(), '.png')
ggsave(
  file = fname, 
  plot = my_plot,
  width = 6.5, height = 5.5, units = "in"
)
layer_data(plot = my_plot) #offset (y) is .1875
# Work on moving the text to the right of the line instead of on top. Remove the white box from geom_tile
```

```{r}
custom_order_esc = c("escape", "non-escape", "not viable")
x_order = rev(c("Grammaticality (Hie)",
            "Grammaticality (ESM2, unmasked)",
            "Grammaticality (ESM2, masked)", 
            "MutComputeX",
            "MutRank",
            "Stability Oracle",
            "EVEscape",
            "EVE",
            "Semantic change (Hie)",
            "Semantic change (ESM2)"
         ))

nested_comb_data_long_esc = comb_data_long %>% 
  subset(Starr_esc != "not tested") %>% 
  subset(Measure != "FoldX") %>% 
  subset(Measure != "PyRosetta") |>
  pivot_wider(names_from = "Starr_esc", values_from = "rank") |>
  select(-c(mutant, ID, Baum_esc, Sequence))  %>% 
  arrange(factor(Measure, levels = x_order))

# comb_data_long_esc %>% select(-c(ID, mutant, Starr_esc, Baum_esc, rank)) %>% group_by_all() %>%
#     summarize(
#         count = n()
#     ) %>%
#   pivot_wider(names_from = "viability", values_from = "count")


nested_comb_data_long_esc_temp = nested_comb_data_long_esc %>% nest(data = -Measure)

data_stats_esc <- nested_comb_data_long_esc |>
  nest(data = -Measure) |>
  mutate(
    esc_not_esc = map(data, ~broom::tidy(wilcox.test(.x$escape, .x$`non-escape`))),
    esc_non_via = map(data, ~broom::tidy(wilcox.test(.x$escape, .x$`not viable`))),
    not_esc_non_via = map(data, ~broom::tidy(wilcox.test(.x$`non-escape`, .x$`not viable`)))
  ) |>
  pivot_longer(
    cols = c(esc_not_esc, esc_non_via, not_esc_non_via),
    names_to = "test",
    values_to = "stats"
  ) |>
  mutate(
    column1 = case_when(
      test == "esc_not_esc" ~ "escape",
      test == "esc_non_via" ~ "escape",
      test == "not_esc_non_via" ~ "not viable"
    ),
    column2 = case_when(
      test == "esc_not_esc" ~ "non-escape",
      test == "esc_non_via" ~ "not viable",
      test == "not_esc_non_via" ~ "non-escape"
    )
  )

data_stats_esc = data_stats_esc |>
  unnest_wider(stats) |>
  select(Measure, column1, column2, p.value)

y_delta = .25 # distance between adjacent bloxplots (.8/number of boxplots)
x_delta = .185 #.075 # distance between line segments
data_comparisons_esc <- data_stats_esc |>
  mutate(
    # apply overall false-discovery rate correction
    p.value = p.adjust(p.value, method = "fdr"),
    # set stars
    label = case_when(
      p.value < 0.001 ~ "***",
      p.value < 0.01 ~ "**",
      p.value < 0.05 ~ "*",
      TRUE ~ "NS"
    ),
    # offset to correct for stars not being vertically centered
    label_off = case_when(
      label == "NS" ~ 0,
      TRUE ~ -.04,
    ),
    # calculate x and y offsets for line segments
    yoff1 = case_when(
      column1 == "escape" ~ y_delta,
      column1 == "not viable" ~ -y_delta,
      TRUE ~ 0
    ),
    yoff2 = case_when(
      column2 == "escape" ~ y_delta,
      column2 == "not viable" ~ -y_delta,
      TRUE ~ 0
    ),
    xoff = case_when(
      column1 == "escape" & column2 == "non-escape" ~ 0,
      column1 == "not viable" & column2 == "non-escape" ~ x_delta,
      TRUE ~ 2 * x_delta
    )
  )

#data_comparisons = data_comparisons %>% arrange(factor(Measure, levels = x_order))

my_plot = nested_comb_data_long_esc %>% 
  subset(Measure != "FoldX") %>% subset(Measure != "PyRosetta") %>%
  pivot_longer(-Measure, names_to = "type", values_to = "value") %>% drop_na(value) |>
  ggplot(aes(x = value, y = factor(Measure, levels = x_order), fill = factor(type, rev(custom_order_esc)))) +
  #ggplot(aes(x = value, y = Measure, fill = type)) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_manual(values = c("escape" = "#5D3A9B",
                                 "non-escape" = "#E66100", 
                                 "not viable" = "#AF596E"),
                      breaks = custom_order_esc,
                      name = "spike mutation type"
                    ) +
  geom_segment( # line segments for significance
    data = data_comparisons_esc,
    aes(
      x = 1.0 + xoff + .03, 
      xend = 1.0 + xoff + .03,
      y = as.numeric(factor(Measure, levels = x_order)) + yoff1, 
      yend = as.numeric(factor(Measure, levels = x_order)) + yoff2
      #y = as.numeric(factor(Measure)) + y_delta, yend = as.numeric(factor(Measure)) - y_delta #midpoint +/- y_delta
    ),
    inherit.aes = FALSE
  ) +
  geom_text(
    data = data_comparisons_esc,
    aes(
      x = 1 + xoff +.04,
      y = as.numeric(factor(Measure, levels = x_order)) + (yoff1 + yoff2) / 2 + label_off,
      label = label,
    ),
    size = 10 / .pt, # 10 point font size
    vjust = 0.5,
    hjust = 0,
    inherit.aes = FALSE
  ) +
  xlim(-.02, 1.8) +
  scale_x_continuous(name = "relative rank", breaks =c(0,.5,1)) +
  scale_y_discrete(labels = c("Grammaticality (ESM2, unmasked)" = "Grammaticality\n(ESM2, unmasked)",
                              "Grammaticality (ESM2, masked)" = "Grammaticality\n(ESM2, masked)",
                              "Grammaticality (Hie)" = "Grammaticality\n(Hie)")) +
  theme_classic() + ylab("") + theme(legend.position = "top")
my_plot

fname = paste0('cov_all_measure_boxplot_escape_', Sys.Date(), '.png')
ggsave(
  file = fname, 
  plot = my_plot,
  width = 6.5, height = 5.5, units = "in"
)
#layer_data(plot = my_plot) #offset (y) is .1875

```

# 

```{r}
comb_data_long = comb_data_norm %>% 
  select(-c(wt, mutations, site, Sequence)) %>%
  pivot_longer(cols = my_measures,
               names_to = "Measure",
               values_to = "rank") %>%
  mutate(Measure = 
           case_when(
             Measure == "ddG_Stab_Ora" ~ "Stability Oracle",
             Measure == "prob_MutCompX" ~ "MutComputeX",
             Measure == "Grammaticality_ESM2_unmasked" ~ "Grammaticality (ESM2, unmasked)",
             Measure == "Semantic.change_ESM2" ~ "Semantic change (ESM2)",
             Measure == "Semantic.change_Hie" ~ "Semantic change (Hie)",
             Measure == "Grammaticality_Hie_logit" ~ "Grammaticality (Hie)",
             Measure == "PyRosetta_ddG_fold" ~ "PyRosetta",
             Measure == "FoldX_ddG_fold" ~ "FoldX",
             Measure == "prob_ESM2_masked" ~ "Grammaticality (ESM2, masked)",
             Measure == "prob_MutRank" ~ "MutRank",
             Measure == "evescape" ~ "EVEscape",
             Measure == "eve" ~ "EVE"
             )
         )
my_measures_rename = colnames(comb_data_long %>% 
                         select(-c(ID, 
                                Starr_esc, Baum_esc,
                                rank)))

my_comparisons_gramm <- list( c("not tested", "viable"), 
                          c("not tested", "not viable"), 
                          c("viable", "not viable") )

comb_data_long_gramm = comb_data_long %>% 
  mutate(viability = case_when(Starr_esc == "not tested" ~ "not tested",
                               Starr_esc == "non-escape" ~ "viable",
                               Starr_esc == "escape" ~ "viable",
                               Starr_esc == "not viable" ~ "not viable"))
comb_data_norm
comb_data_long_gramm
```

# Scatter

```{r}
 custom_esc_order = (c("escape","non-escape", "not viable","not tested"))
comb_data_scatter = comb_data %>% 
  mutate(viability = case_when(Starr_esc == "not tested" ~ "not tested",
                               Starr_esc == "non-escape" ~ "viable",
                               Starr_esc == "escape" ~ "viable",
                               Starr_esc == "not viable" ~ "not viable"),
         Starr_esc = factor(Starr_esc, levels = rev(custom_esc_order))) %>%
  select(viability, Starr_esc, ID, Semantic.change_Hie, Grammaticality_Hie_logit) %>%
  arrange(Starr_esc)

p_cov_scatter_ungrouped= ggplot() + #ggplot(subset(comb_data, Starr_esc != "not viable")) +
  geom_point( data =comb_data_scatter, 
              aes(x = Grammaticality_Hie_logit, y = Semantic.change_Hie, 
                 color = Starr_esc),
             alpha = 0.5, size = 1) +
  geom_point(data = subset(comb_data, Baum_esc == "escape"), 
             aes(x = Grammaticality_Hie_logit, y = Semantic.change_Hie),
             shape = 4, color = "black", size = 1.5, stroke = .75, alpha = .7)  +
  geom_point(data = subset(comb_data, Baum_esc == "non-escape"), 
             aes(x = Grammaticality_Hie_logit, y = Semantic.change_Hie),
             shape = 16, color = "black", size = 1.5, stroke = .75, alpha = .7) +
  scale_color_manual(values = c("escape" = "#5D3A9B",
                                  "non-escape" = "#E66100", 
                                  "not viable" = "#AF596E",
                                  "not tested" = "#C5C5C5"))  +
  theme_bw() + theme(legend.position = "none",#"top",
                     #aspect.ratio = 1,
                     legend.title = element_blank()
                     ) + 
  ylab("Semantic Change") +
  xlab("Grammaticality")
p_cov_scatter_ungrouped
p_cov_scatter_marginal = ggMarginal(p_cov_scatter_ungrouped, 
                                    groupColour = TRUE, groupFill = TRUE,
                                    alpha = 0.0)

fname = paste0('cov_hie_scatter_escape_BaumX_', Sys.Date(), '.png')
ggsave(
  file = fname, 
  plot = p_cov_scatter_marginal,
  width = 3.5, height = 3.5, units = "in",
  dpi = 600
)

# fname = paste0('cov_hie_scatter_escape_nonescape_BaumX_', Sys.Date(), '.png')
# ggsave(
#   file = fname, 
#   plot = p_cov_scatter + 
#   geom_point(data = subset(comb_data, Baum_esc == "non-escape"), 
#              aes(x = Grammaticality_Hie_logit, y = Semantic.change_Hie),
#                  shape = 16, color = "black", size = 1.5, stroke = .75, alpha = .5) 
#   ,
#   width = 2.5, height = 3.2, units = "in",
#   dpi = 600
# )
# 
# p_cov_scatter + 
#   geom_point(data = subset(comb_data, Baum_esc == "non-escape"), 
#              aes(x = Grammaticality_Hie_logit, y = Semantic.change_Hie),
#                  shape = 16, color = "black", size = 1.5, stroke = .75, alpha = .5) +
#   theme(legend.position = "bottom")
#   
# 
# fname = paste0('cov_Starr_DMS_piechart_', Sys.Date(), '.png')
# ggsave(
#   file = fname, 
#   plot = cov_esc_pie + theme(legend.position = "none"),
#   width = 3, height = 3.2, units = "in",
#   dpi = 600
# )
  
```

#Pie + Scatter

```{r}
library(cowplot)
setwd("/Users/beallma/OneDrive - Emory University/Documents/PIPP/SARS_NLP/aggregate_data_vizualization")
#renv::init()
```

```{r}
#mywd = "/Users/beallma/OneDrive - Emory University/Documents/PIPP/SARS_NLP/aggregate_data_vizualization"
setwd("/Users/beallma/OneDrive - Emory University/Documents/PIPP/SARS_NLP/aggregate_data_vizualization")
p_cov_scatter_marginal
cov_esc_pie

prow <- plot_grid(
  p_cov_scatter_marginal,
  cov_esc_pie + theme(legend.position="none") + theme(plot.margin = unit(c(-.35,-.35,-.35,-.35),"cm")),
  align = 'vh',
  labels = c("A", "B"),
  hjust = -1,
  nrow = 1,
  rel_widths = c(1, 1)
)
prow

legend_b <- get_legend(
  cov_esc_pie + 
    guides(color = guide_legend(nrow = 1)) +
    theme(legend.position = "bottom")
)
scatter_pie = plot_grid(prow, legend_b, 
          ncol = 1,
          rel_heights = c(1, .15))
scatter_pie
fname = paste0('cov_scatter_escape_BaumX_piechart_', Sys.Date(), '.png')
ggsave(
  file = fname,
  plot = scatter_pie + theme(plot.background = element_rect(fill = "white", linetype = "blank"),
                             panel.border = element_blank()),
  width = 5.3, height = 3.8, units = "in"
)

```

# Escape Barplot
```{r}
custom_order_esc = c("escape", "non-escape", "not viable")
x_order = rev(c("Grammaticality (Hie)",
            "Grammaticality (ESM2, unmasked)",
            "Grammaticality (ESM2, masked)", 
            "MutComputeX",
            "MutRank",
            "Stability Oracle",
            "EVE"
         ))

data_comparisons_esc_gramm = data_comparisons_esc %>%
  subset(!(Measure %in% c("Semantic change (Hie)", 
                          "Semantic change (ESM2)",
                          "EVEscape",
                          "PyRosetta",
                          "FoldX"))) %>% arrange(factor(Measure, levels = (x_order)))
data_comparisons_esc_sem = data_comparisons_esc %>%
  subset(Measure %in% c("Semantic change (Hie)", 
                          "Semantic change (ESM2)",
                        "EVEscape"))

nested_comb_data_long_esc_gramm = nested_comb_data_long_esc %>%
  subset(!(Measure %in% c("Semantic change (Hie)", 
                          "Semantic change (ESM2)",
                          "EVEscape",
                          "PyRosetta",
                          "FoldX"))) %>% arrange(factor(Measure, levels = (x_order)))
nested_comb_data_long_esc_sem = nested_comb_data_long_esc %>%
  subset(Measure %in% c("Semantic change (Hie)", 
                          "Semantic change (ESM2)",
                        "EVEscape"))

esc_gramm_plot = nested_comb_data_long_esc_gramm %>%
  pivot_longer(-Measure, names_to = "type", values_to = "value") %>% drop_na(value) |>
  ggplot(aes(x = value, y = factor(Measure, level = x_order), fill = factor(type, rev(custom_order_esc)))) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_manual(values = c("escape" = "#5D3A9B",
                                 "non-escape" = "#E66100", 
                                 "not viable" = "#AF596E"),
                      breaks = custom_order_esc,
                      name = "spike\nmutation type"
                    ) +
  geom_segment( # line segments for significance
    data = data_comparisons_esc_gramm,
    aes(
      x = 1.0 + xoff + .03, 
      xend = 1.0 + xoff + .03,
      y = as.numeric(factor(Measure, level = x_order)) + yoff1, 
      yend = as.numeric(factor(Measure, level = x_order)) + yoff2
      #y = as.numeric(factor(Measure)) + y_delta, yend = as.numeric(factor(Measure)) - y_delta #midpoint +/- y_delta
    ),
    inherit.aes = FALSE
  ) +
  geom_text(
    data = data_comparisons_esc_gramm,
    aes(
      x = 1 + xoff +.04,
      y = as.numeric(factor(Measure, level = x_order)) + (yoff1 + yoff2) / 2 + label_off,
      label = label,
    ),
    size = 10 / .pt, # 10 point font size
    vjust = 0.5,
    hjust = 0,
    inherit.aes = FALSE
  ) +
  xlim(-.02, 1.55) +
  scale_x_continuous(name = "relative rank", breaks =c(0,.5,1)) +
  scale_y_discrete(labels = c("Grammaticality (ESM2, unmasked)" = "Grammaticality\n(ESM2, unmasked)",
                              "Grammaticality (ESM2, masked)" = "Grammaticality\n(ESM2, masked)",
                              "Grammaticality (Hie)" = "Grammaticality\n(Hie)")) +
  theme_classic() + ylab("") + theme(legend.position = "top")
esc_gramm_plot

esc_sem_plot = nested_comb_data_long_esc_sem %>%
  pivot_longer(-Measure, names_to = "type", values_to = "value") %>% drop_na(value) |>
  ggplot(aes(x = value, y = factor(Measure), fill = factor(type, rev(custom_order_esc)))) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_manual(values = c("escape" = "#5D3A9B",
                                 "non-escape" = "#E66100", 
                                 "not viable" = "#AF596E"),
                      breaks = custom_order_esc,
                      name = "spike mutation type"
                    ) +
  geom_segment( # line segments for significance
    data = data_comparisons_esc_sem,
    aes(
      x = 1.0 + xoff + .03, 
      xend = 1.0 + xoff + .03,
      y = as.numeric(factor(Measure)) + yoff1, 
      yend = as.numeric(factor(Measure)) + yoff2
    ),
    inherit.aes = FALSE
  ) +
  geom_text(
    data = data_comparisons_esc_sem,
    aes(
      x = 1 + xoff +.04,
      y = as.numeric(factor(Measure)) + (yoff1 + yoff2) / 2 + label_off,
      label = label,
    ),
    size = 10 / .pt, # 10 point font size
    vjust = 0.5,
    hjust = 0,
    inherit.aes = FALSE
  ) +
  xlim(-.04, 1.55) +
  scale_x_continuous(name = "relative rank", breaks =c(0,.5,1)) +
  scale_y_discrete(labels = c("Semantic change (ESM2)" = "Semantic change\n(ESM2)",
                              "Semantic change (Hie)" = "Semantic change\n(Hie)")
                   ) +
  theme_classic() + ylab("") + theme(legend.position = "none")
esc_sem_plot
```

```{r}
esc_sem_plot_new_labs = esc_sem_plot + 
  theme(legend.position="none")+ # plot.margin = margin(r = 0, l = 0, t= 0, b = 0.1, unit = "inches"))
  scale_y_discrete(labels = c("Semantic change (ESM2)" = "Semantic\nchange\n(ESM2)",
                              "Semantic change (Hie)" = "Semantic\nchange\n(Hie)")) + 
  scale_x_continuous(name = "relative rank", breaks =c(0,.5,1), limits = c(NA, 1.55))
esc_sem_plot_new_labs

esc_gramm_plot_new = esc_gramm_plot + 
  theme(legend.position="none", plot.margin = margin(l = 0,t = 0.1, b = 0.1, r = 0.1, unit = "inches"))+#unit(c(0, 0, 0, 0), "inches"))  + 
  #xlim(-.02,2.5) +
  scale_x_continuous(name = "relative rank", breaks =c(0,.5,1), limits = c(NA, 1.55))
esc_gramm_plot_new

legend_b <- get_legend(
  esc_gramm_plot_new + theme(legend.position="right",
                             legend.key.width = unit(.6, "in"),
                             legend.key.height = unit(.1, "in")) +
    guides(color = guide_legend(nrow = 3,
                                title = "spike\nmutation type"))
)

aligned <- align_plots(esc_gramm_plot_new+ xlab("relative rank"), 
                       esc_sem_plot_new_labs+ xlab("relative rank"),
                       align = 'h', axis = 'l')

sem_col <- plot_grid(
  aligned[[2]],#via_sem_plot_new_labs+ xlab("relative rank"),
  legend_b,
  NULL,
  labels = c("", "", ""),
  nrow = 3,
  rel_heights = c(1, .5, .35)
)
sem_col

gram_col = plot_grid(
  aligned[[1]] ,
  NULL,
  labels = c("", ""),
  nrow = 2,
  rel_heights = c(1, .05)
)

escape_all_measures = plot_grid(
  sem_col,#aligned[[1]], 
  #NULL,
  gram_col, 
  ncol = 2,
  align = 'h',
  labels = c("A", "B"),
  label_x = 0.05, label_y = 0.95,
  hjust = -0.5, vjust = -0.5,
  rel_widths = c(.75, 1))

escape_all_measures

fname = paste0('cov_boxplots_escape_', Sys.Date(), '.png')
ggsave(
  file = fname, 
  plot = escape_all_measures + theme(plot.background = element_rect(fill = "white", linetype = "blank"),
                             panel.border = element_blank()),
  width = 6.5, height = 5.3, units = "in"
)
```
##Viability Barplot
```{r}
nested_comb_data_long_gramm = comb_data_long_gramm |>
  pivot_wider(names_from = "viability", values_from = "rank") |>
  select(-c(mutant, Starr_esc, Baum_esc)) %>%
  subset(!(Measure %in% c(
                          "PyRosetta",
                          "FoldX", NA)))

comb_data_long_gramm %>% select(-c(ID, mutant, Starr_esc, Baum_esc, rank)) %>% group_by_all() %>%
    summarize(
        count = n()
    ) %>%
  pivot_wider(names_from = "viability", values_from = "count")

custom_order_via = c("viable", "not viable", "not tested")

x_order = rev(c("Grammaticality (Hie)",
            "Grammaticality (ESM2, unmasked)",
            "Grammaticality (ESM2, masked)", 
            "MutComputeX",
            "MutRank",
            "Stability Oracle",
            "EVE"
         ))

data_stats_gramm <- nested_comb_data_long_gramm %>% select(-ID) |>
  nest(data = -Measure) %>%
  mutate(
    via_non_via = map(data, ~broom::tidy(wilcox.test(.x$viable, .x$`not viable`))),
    via_not_tes = map(data, ~broom::tidy(wilcox.test(.x$viable, .x$`not tested`))),
    non_via_not_tes = map(data, ~broom::tidy(wilcox.test(.x$`not viable`, .x$`not tested`)))
  ) |>
  pivot_longer(
    cols = c(via_non_via, via_not_tes, non_via_not_tes),
    names_to = "test",
    values_to = "stats"
  ) |>
  mutate(
    column1 = case_when(
      test == "via_non_via" ~ "viable",
      test == "via_not_tes" ~ "viable",
      test == "non_via_not_tes" ~ "not tested"
    ),
    column2 = case_when(
      test == "via_non_via" ~ "not viable",
      test == "via_not_tes" ~ "not tested",
      test == "non_via_not_tes" ~ "not viable"
    )
  ) |>
  unnest_wider(stats) |>
  select(Measure, column1, column2, p.value)

y_delta = .25 # distance between adjacent bloxplots (.8/number of boxplots)
x_delta = .185 # distance between line segments
data_comparisons_gramm <- data_stats_gramm |>
  mutate(
    # apply overall false-discovery rate correction
    p.value = p.adjust(p.value, method = "fdr"),
    # set stars
    label = case_when(
      p.value < 0.001 ~ "***",
      p.value < 0.01 ~ "**",
      p.value < 0.05 ~ "*",
      TRUE ~ "NS"
    ),
    # offset to correct for stars not being vertically centered
    label_off = case_when(
      label == "NS" ~ 0,
      TRUE ~ -.04,
    ),
    # calculate x and y offsets for line segments
    yoff1 = case_when(
      column1 == "viable" ~ y_delta,
      column1 == "not tested" ~ -y_delta,
      TRUE ~ 0
    ),
    yoff2 = case_when(
      column2 == "viable" ~ y_delta,
      column2 == "not tested" ~ -y_delta,
      TRUE ~ 0
    ),
    xoff = case_when(
      column1 == "viable" & column2 == "not viable" ~ 0,
      column1 == "not tested" & column2 == "not viable" ~ x_delta,
      TRUE ~ 2 * x_delta
    )
  )

data_comparisons_via_gramm = data_comparisons_gramm %>% 
  subset(!(Measure %in% c("Semantic change (Hie)", 
                          "Semantic change (ESM2)",
                          "EVEscape",
                          "PyRosetta",
                          "FoldX"))) %>% arrange(factor(Measure, levels = (x_order)))
data_comparisons_via_sem = data_comparisons_gramm %>%
  subset(Measure %in% c("Semantic change (Hie)", 
                          "Semantic change (ESM2)",
                        "EVEscape"))

nested_comb_data_long_via_gramm = nested_comb_data_long_gramm %>%
  subset(!(Measure %in% c("Semantic change (Hie)", 
                          "Semantic change (ESM2)",
                          "EVEscape",
                          "PyRosetta",
                          "FoldX"))) %>% arrange(factor(Measure, levels = (x_order)))
nested_comb_data_long_via_sem = nested_comb_data_long_gramm %>%
  subset(Measure %in% c("Semantic change (Hie)", 
                          "Semantic change (ESM2)",
                        "EVEscape"))

via_gramm_plot = nested_comb_data_long_gramm %>% select(-ID, -Sequence) %>% 
  subset(!(Measure %in% c("Semantic change (Hie)", 
                          "Semantic change (ESM2)",
                          "EVEscape",
                          "PyRosetta",
                          "FoldX"))) %>% arrange(factor(Measure, levels = (x_order))) %>%
  pivot_longer(-Measure, names_to = "type", values_to = "value") %>% drop_na(value) |>
  ggplot(aes(x = value, y = factor(Measure, level = x_order), fill = factor(type, rev(custom_order_via)))) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_manual(values = c("viable" = "#399c52",
                                 "not viable" = '#AF596E', #"#cd6a83", 
                                 "not tested" = "#C5C5C5"),
                      breaks = custom_order_via,
                      name = "spike\nmutation type"
                    ) +
  geom_segment( # line segments for significance
    data = data_comparisons_via_gramm,
    aes(
      x = 1.0 + xoff + .03, 
      xend = 1.0 + xoff + .03,
      y = as.numeric(factor(Measure, level = x_order)) + yoff1, 
      yend = as.numeric(factor(Measure, level = x_order)) + yoff2
      #y = as.numeric(factor(Measure)) + y_delta, yend = as.numeric(factor(Measure)) - y_delta #midpoint +/- y_delta
    ),
    inherit.aes = FALSE
  ) +
  geom_text(
    data = data_comparisons_via_gramm,
    aes(
      x = 1 + xoff +.04,
      y = as.numeric(factor(Measure, level = x_order)) + (yoff1 + yoff2) / 2 + label_off,
      label = label,
    ),
    size = 10 / .pt, # 10 point font size
    vjust = 0.5,
    hjust = 0,
    inherit.aes = FALSE
  ) +
  xlim(-.02, 1.55) +
  scale_x_continuous(name = "relative rank", breaks =c(0,.5,1)) +
  scale_y_discrete(labels = c("Grammaticality (ESM2, unmasked)" = "Grammaticality\n(ESM2, unmasked)",
                              "Grammaticality (ESM2, masked)" = "Grammaticality\n(ESM2, masked)",
                              "Grammaticality (Hie)" = "Grammaticality\n(Hie)")) +
  theme_classic() + ylab("") + theme(legend.position = "top")
via_gramm_plot

via_sem_plot = nested_comb_data_long_via_sem %>% select(-ID, -Sequence) %>% 
  subset(Measure %in% c("Semantic change (Hie)", 
                          "Semantic change (ESM2)",
                        "EVEscape"
                        )) %>% 
  pivot_longer(-Measure, names_to = "type", values_to = "value") %>% drop_na(value) |>
  ggplot(aes(x = value, y = factor(Measure), fill = factor(type, rev(custom_order_via)))) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_manual(values = c("viable" = "#399c52",
                                 "not viable" = '#AF596E', #"#cd6a83", 
                                 "not tested" = "#C5C5C5"),
                      breaks = custom_order_via,
                      name = "spike mutation type"
                    ) +
  geom_segment( # line segments for significance
    data = data_comparisons_via_sem,
    aes(
      x = 1.0 + xoff + .03, 
      xend = 1.0 + xoff + .03,
      y = as.numeric(factor(Measure)) + yoff1, 
      yend = as.numeric(factor(Measure)) + yoff2
    ),
    inherit.aes = FALSE
  ) +
  geom_text(
    data = data_comparisons_via_sem,
    aes(
      x = 1 + xoff +.04,
      y = as.numeric(factor(Measure)) + (yoff1 + yoff2) / 2 + label_off,
      label = label,
    ),
    size = 10 / .pt, # 10 point font size
    vjust = 0.5,
    hjust = 0,
    inherit.aes = FALSE
  ) +
  xlim(-.04, 1.55) +
  scale_x_continuous(name = "relative rank", breaks =c(0,.5,1)) +
  scale_y_discrete(labels = c("Semantic change (ESM2)" = "Semantic\nchange\n(ESM2)",
                              "Semantic change (Hie)" = "Semantic\nchange\n(Hie)")
                   ) +
  theme_classic() + ylab("") + theme(legend.position = "none")
via_sem_plot

```

```{r}
via_sem_plot_new_labs = via_sem_plot + 
  theme(legend.position="none")+ # plot.margin = margin(r = 0, l = 0, t= 0, b = 0.1, unit = "inches"))
  scale_y_discrete(labels = c("Semantic change (ESM2)" = "Semantic\nchange\n(ESM2)",
                              "Semantic change (Hie)" = "Semantic\nchange\n(Hie)")) + 
  scale_x_continuous(name = "relative rank", breaks =c(0,.5,1), limits = c(NA, 1.55))
via_sem_plot_new_labs

via_gramm_plot_new = via_gramm_plot + 
  theme(legend.position="none", plot.margin = margin(l = 0,t = 0.1, b = 0.1, r = 0.1, unit = "inches"))+#unit(c(0, 0, 0, 0), "inches"))  + 
  #xlim(-.02,2.5) +
  scale_x_continuous(name = "relative rank", breaks =c(0,.5,1), limits = c(NA, 1.55))
via_gramm_plot_new

legend_b <- get_legend(
  via_gramm_plot_new + theme(legend.position="right",
                             legend.key.width = unit(.6, "in"),
                             legend.key.height = unit(.1, "in")) +
    guides(color = guide_legend(nrow = 3,
                                title = "spike\nmutation type"))
)

aligned <- align_plots(via_gramm_plot_new+ xlab("relative rank"), 
                       via_sem_plot_new_labs+ xlab("relative rank"), 
                       align = 'h', axis = 'l')

sem_col <- plot_grid(
  aligned[[2]],#via_sem_plot_new_labs+ xlab("relative rank"),
  legend_b,
  NULL,
  labels = c("", "", ""),
  nrow = 3,
  rel_heights = c(1, .5, .35)
)
sem_col

gram_col = plot_grid(
  aligned[[1]] ,
  NULL,
  labels = c("", ""),
  nrow = 2,
  rel_heights = c(1, .05)
)

via_all_measures = plot_grid(
  gram_col,#aligned[[1]], 
  #NULL,
  sem_col, 
  ncol = 2,
  align = 'h',
  labels = c("A", "B"),
  label_x = 0.05, label_y = 0.95,
  hjust = -0.5, vjust = -0.5,
  rel_widths = c(1, .75))

via_all_measures

fname = paste0('cov_boxplots_viability_', Sys.Date(), '.png')
ggsave(
  file = fname, 
  plot = via_all_measures + theme(plot.background = element_rect(fill = "white", linetype = "blank"),
                             panel.border = element_blank()),
  width = 6.5, height = 5.3, units = "in"
)
```

##Playground
```{r}
custom_order_esc = c("escape", "non-escape", "not viable")
x_order = rev(c("Grammaticality (Hie)",
            "Grammaticality (ESM2, unmasked)",
            "Grammaticality (ESM2, masked)", 
            "MutComputeX",
            "MutRank",
            "Stability Oracle"
         ))

data_comparisons_esc_gramm = data_comparisons_esc %>%
  subset(!(Measure %in% c("Semantic change (Hie)", 
                          "Semantic change (ESM2)",
                          "PyRosetta",
                          "FoldX"))) %>% arrange(factor(Measure, levels = (x_order)))
data_comparisons_esc_sem = data_comparisons_esc %>%
  subset(Measure %in% c("Semantic change (Hie)", 
                          "Semantic change (ESM2)"))

nested_comb_data_long_esc_gramm = nested_comb_data_long_esc %>%
  subset(!(Measure %in% c("Semantic change (Hie)", 
                          "Semantic change (ESM2)",
                          "PyRosetta",
                          "FoldX"))) %>% arrange(factor(Measure, levels = (x_order)))
nested_comb_data_long_esc_sem = nested_comb_data_long_esc %>%
  subset(Measure %in% c("Semantic change (Hie)", 
                          "Semantic change (ESM2)"))

esc_gramm_plot = nested_comb_data_long_esc_gramm %>% subset(Measure %in% c("Grammaticality (Hie)",
            "Grammaticality (ESM2, unmasked)",
            "Grammaticality (ESM2, masked)"
         )) %>%
  pivot_longer(-Measure, names_to = "type", values_to = "value") %>% drop_na(value) |>
  ggplot(aes(x = value, y = factor(Measure, level = x_order), fill = factor(type, rev(custom_order_esc)))) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_manual(values = c("escape" = "#5D3A9B",
                                 "non-escape" = "#E66100", 
                                 "not viable" = "#AF596E"),
                      breaks = custom_order_esc,
                      name = "spike\nmutation type"
                    ) +
  geom_segment( # line segments for significance
    data = data_comparisons_esc_gramm,
    aes(
      x = 1.0 + xoff + .03, 
      xend = 1.0 + xoff + .03,
      y = as.numeric(factor(Measure, level = x_order)) + yoff1, 
      yend = as.numeric(factor(Measure, level = x_order)) + yoff2
      #y = as.numeric(factor(Measure)) + y_delta, yend = as.numeric(factor(Measure)) - y_delta #midpoint +/- y_delta
    ),
    inherit.aes = FALSE
  ) +
  geom_text(
    data = data_comparisons_esc_gramm,
    aes(
      x = 1 + xoff +.04,
      y = as.numeric(factor(Measure, level = x_order)) + (yoff1 + yoff2) / 2 + label_off,
      label = label,
    ),
    size = 10 / .pt, # 10 point font size
    vjust = 0.5,
    hjust = 0,
    inherit.aes = FALSE
  ) +
  xlim(-.02, 1.55) +
  scale_x_continuous(name = "relative rank", breaks =c(0,.5,1)) +
  scale_y_discrete(labels = c("Grammaticality (ESM2, unmasked)" = "Grammaticality\n(ESM2, unmasked)",
                              "Grammaticality (ESM2, masked)" = "Grammaticality\n(ESM2, masked)",
                              "Grammaticality (Hie)" = "Grammaticality\n(Hie)")) +
  theme_classic() + ylab("") + theme(legend.position = "top")
esc_gramm_plot

esc_sem_plot = nested_comb_data_long_esc_sem %>%
  pivot_longer(-Measure, names_to = "type", values_to = "value") %>% drop_na(value) |>
  ggplot(aes(x = value, y = factor(Measure), fill = factor(type, rev(custom_order_esc)))) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_manual(values = c("escape" = "#5D3A9B",
                                 "non-escape" = "#E66100", 
                                 "not viable" = "#AF596E"),
                      breaks = custom_order_esc,
                      name = "spike mutation type"
                    ) +
  geom_segment( # line segments for significance
    data = data_comparisons_esc_sem,
    aes(
      x = 1.0 + xoff + .03, 
      xend = 1.0 + xoff + .03,
      y = as.numeric(factor(Measure)) + yoff1, 
      yend = as.numeric(factor(Measure)) + yoff2
    ),
    inherit.aes = FALSE
  ) +
  geom_text(
    data = data_comparisons_esc_sem,
    aes(
      x = 1 + xoff +.04,
      y = as.numeric(factor(Measure)) + (yoff1 + yoff2) / 2 + label_off,
      label = label,
    ),
    size = 10 / .pt, # 10 point font size
    vjust = 0.5,
    hjust = 0,
    inherit.aes = FALSE
  ) +
  xlim(-.04, 1.55) +
  scale_x_continuous(name = "relative rank", breaks =c(0,.5,1)) +
  scale_y_discrete(labels = c("Semantic change (ESM2)" = "Semantic change\n(ESM2)",
                              "Semantic change (Hie)" = "Semantic change\n(Hie)")
                   ) +
  theme_classic() + ylab("") + theme(legend.position = "none")
esc_sem_plot
```

##ESM-2 Concat
These data were made by creating Embeddings for the S1, S2, and N-terminus domains of the spike protein and then concatenating them together. The same was done for the reference sequence.

```{r}
#fname = "cov/Baum_pretrain_screen_ESM2_logit_odds.txt"
fname = "/Users/beallma/Documents/GitHub/NLP_viral_escape/language_models/results/cov/cov_concat_embed_base_ESM2_082724.log"
df_ESM2_conc_sub = read.delim(file = fname, header = TRUE, sep="",quote = "\"",
                 dec = ".", fill = TRUE, comment.char = "#")
df_ESM2_conc_sub = df_ESM2_conc_sub %>% dplyr::rename("Semantic.change_ESM2_concat" = "Semantic.change",
                "Grammaticality_ESM2_unmasked_concat" = "Grammaticality") %>% 
  # mutate_at(c(
  #     'Grammaticality_ESM2_logit'
  #   ), rank) %>% mutate_at(c(
  #     'Semantic.change_ESM2'
  #   ), rank)%>% 
  # rename("Grammaticality_ESM2_logit_rank" = Grammaticality_ESM2_logit,
  #        "Semantic.change_ESM2_rank" = Semantic.change_ESM2
  #        ) %>%
  # mutate(
  #     Grammaticality_ESM2_logit_rank = max(Grammaticality_ESM2_logit_rank) - Grammaticality_ESM2_logit_rank + 1
  # ) %>%
  select(-Mutations, -Sequence)
#high grammar means more like the original sequence, so we want to invert^ this to match semantic change
df_ESM2_conc_sub
df_ESM2_Baum
```
```{r}
esm2_concat_compare = df_ESM2_Baum %>% left_join(df_ESM2_conc_sub, by = "ID") %>% 
  left_join(all_sites_escape_viable, by = c('ID' = 'ID')) %>% 
  mutate(viability = case_when(Starr_esc == "not tested" ~ "not tested",
                               Starr_esc == "non-escape" ~ "viable",
                               Starr_esc == "escape" ~ "viable",
                               Starr_esc == "not viable" ~ "not viable"),
         Starr_esc = factor(Starr_esc, levels = rev(custom_esc_order)))
esm2_concat_compare 

concat_scatter= ggplot() + #ggplot(subset(comb_data, Starr_esc != "not viable")) +
  geom_point( data =esm2_concat_compare %>% filter(Starr_esc == "not tested"), 
              aes(x = Grammaticality_ESM2_unmasked_concat, y = Semantic.change_ESM2_concat, 
                 color = Starr_esc),
             alpha = 0.5, size = 1.5) +
  geom_point( data =esm2_concat_compare %>% filter(Starr_esc == "not viable"), 
              aes(x = Grammaticality_ESM2_unmasked_concat, y = Semantic.change_ESM2_concat, 
                 color = Starr_esc),
             alpha = 0.5, size = 1.5) +
  geom_point( data =esm2_concat_compare %>% filter(Starr_esc == "non-escape"), 
              aes(x = Grammaticality_ESM2_unmasked_concat, y = Semantic.change_ESM2_concat, 
                 color = Starr_esc),
             alpha = 0.5, size = 1.5) +
  geom_point( data =esm2_concat_compare %>% filter(Starr_esc == "escape"), 
              aes(x = Grammaticality_ESM2_unmasked_concat, y = Semantic.change_ESM2_concat, 
                 color = Starr_esc),
             alpha = 0.5, size = 1.5) +
  scale_color_manual(values = c("escape" = "#5D3A9B",
                                 "non-escape" = "#E66100", 
                                 "not viable" = "#AF596E",
                                 "not tested" = "#C5C5C5")) +
  geom_point(data = subset(esm2_concat_compare, Baum_esc == "escape"), 
             aes(x = Grammaticality_ESM2_unmasked_concat, y = Semantic.change_ESM2_concat),
                 shape = 4, color = "black", size = 1.5, stroke = .75, alpha = .7)  +
  theme_bw() + theme(legend.position = "none",#"top",
                     #aspect.ratio = 1,
                     legend.title = element_blank()
                     ) + 
  ylab("Semantic Change") +
  xlab("Grammaticality")# + title("Concat. S1, S2, N-term Embeds")
concat_scatter

sc_vs_sc_concat= ggplot() + #ggplot(subset(comb_data, Starr_esc != "not viable")) +
  geom_point( data =esm2_concat_compare %>% filter(Starr_esc == "not tested"), 
              aes(x = Semantic.change_ESM2, y = Semantic.change_ESM2_concat, 
                 color = Starr_esc),
             alpha = 0.5, size = 1.5) +
  geom_point( data =esm2_concat_compare %>% filter(Starr_esc == "not viable"), 
              aes(x = Semantic.change_ESM2, y = Semantic.change_ESM2_concat, 
                 color = Starr_esc),
             alpha = 0.5, size = 1.5) +
  geom_point( data =esm2_concat_compare %>% filter(Starr_esc == "non-escape"), 
              aes(x = Semantic.change_ESM2, y = Semantic.change_ESM2_concat, 
                 color = Starr_esc),
             alpha = 0.5, size = 1.5) +
  geom_point( data =esm2_concat_compare %>% filter(Starr_esc == "escape"), 
              aes(x = Semantic.change_ESM2, y = Semantic.change_ESM2_concat, 
                 color = Starr_esc),
             alpha = 0.5, size = 1.5) +
  scale_color_manual(values = c("escape" = "#5D3A9B",
                                 "non-escape" = "#E66100", 
                                 "not viable" = "#AF596E",
                                 "not tested" = "#C5C5C5")) +
  geom_abline(intercept = 0, slope = 1) + 
  #geom_point(data = subset(esm2_concat_compare, Baum_esc == "escape"), 
  #           aes(x = Grammaticality_ESM2_unmasked_concat, y = Semantic.change_ESM2_concat),
  #               shape = 4, color = "black", size = 1.5, stroke = .75, alpha = .7)  +
  theme_bw() + theme(legend.position = "none",#"top",
                     #aspect.ratio = 1,
                     legend.title = element_blank()
                     ) + 
  ylab("Semantic Change (concat)") +
  xlab("Semantic Change (full seq)")# + title("Concat. S1, S2, N-term Embeds")
sc_vs_sc_concat
```
```{r}
custom_order_esc = c("escape", "non-escape", "not viable")
x_order = rev(c("Grammaticality (Hie)",
            "Grammaticality (ESM2, unmasked)",
            "Grammaticality (ESM2, masked)", 
            "MutComputeX",
            "MutRank",
            "Stability Oracle",
            "Semantic change (Hie)",
            "Semantic change (ESM2)"
         ))

nested_comb_data_long_esc = comb_data_long %>% 
  subset(Starr_esc != "not tested") %>% 
  subset(Measure != "FoldX") %>% 
  subset(Measure != "PyRosetta") |>
  pivot_wider(names_from = "Starr_esc", values_from = "rank") |>
  select(-c(mutant, ID, Baum_esc))  %>% 
  arrange(factor(Measure, levels = x_order))

# comb_data_long_esc %>% select(-c(ID, mutant, Starr_esc, Baum_esc, rank)) %>% group_by_all() %>%
#     summarize(
#         count = n()
#     ) %>%
#   pivot_wider(names_from = "viability", values_from = "count")


nested_comb_data_long_esc_temp = nested_comb_data_long_esc %>% nest(data = -Measure)

data_stats_esc <- nested_comb_data_long_esc |>
  nest(data = -Measure) |>
  mutate(
    esc_not_esc = map(data, ~broom::tidy(wilcox.test(.x$escape, .x$`non-escape`))),
    esc_non_via = map(data, ~broom::tidy(wilcox.test(.x$escape, .x$`not viable`))),
    not_esc_non_via = map(data, ~broom::tidy(wilcox.test(.x$`non-escape`, .x$`not viable`)))
  ) |>
  pivot_longer(
    cols = c(esc_not_esc, esc_non_via, not_esc_non_via),
    names_to = "test",
    values_to = "stats"
  ) |>
  mutate(
    column1 = case_when(
      test == "esc_not_esc" ~ "escape",
      test == "esc_non_via" ~ "escape",
      test == "not_esc_non_via" ~ "not viable"
    ),
    column2 = case_when(
      test == "esc_not_esc" ~ "non-escape",
      test == "esc_non_via" ~ "not viable",
      test == "not_esc_non_via" ~ "non-escape"
    )
  )

data_stats_esc = data_stats_esc |>
  unnest_wider(stats) |>
  select(Measure, column1, column2, p.value)

y_delta = .25 # distance between adjacent bloxplots (.8/number of boxplots)
x_delta = .185 #.075 # distance between line segments
data_comparisons_esc <- data_stats_esc |>
  mutate(
    # apply overall false-discovery rate correction
    p.value = p.adjust(p.value, method = "fdr"),
    # set stars
    label = case_when(
      p.value < 0.001 ~ "***",
      p.value < 0.01 ~ "**",
      p.value < 0.05 ~ "*",
      TRUE ~ "NS"
    ),
    # offset to correct for stars not being vertically centered
    label_off = case_when(
      label == "NS" ~ 0,
      TRUE ~ -.04,
    ),
    # calculate x and y offsets for line segments
    yoff1 = case_when(
      column1 == "escape" ~ y_delta,
      column1 == "not viable" ~ -y_delta,
      TRUE ~ 0
    ),
    yoff2 = case_when(
      column2 == "escape" ~ y_delta,
      column2 == "not viable" ~ -y_delta,
      TRUE ~ 0
    ),
    xoff = case_when(
      column1 == "escape" & column2 == "non-escape" ~ 0,
      column1 == "not viable" & column2 == "non-escape" ~ x_delta,
      TRUE ~ 2 * x_delta
    )
  )

#data_comparisons = data_comparisons %>% arrange(factor(Measure, levels = x_order))

my_plot = nested_comb_data_long_esc %>% 
  subset(Measure != "FoldX") %>% subset(Measure != "PyRosetta") %>%
  pivot_longer(-Measure, names_to = "type", values_to = "value") %>% drop_na(value) |>
  ggplot(aes(x = value, y = factor(Measure, levels = x_order), fill = factor(type, rev(custom_order_esc)))) +
  #ggplot(aes(x = value, y = Measure, fill = type)) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_manual(values = c("escape" = "#5D3A9B",
                                 "non-escape" = "#E66100", 
                                 "not viable" = "#AF596E"),
                      breaks = custom_order_esc,
                      name = "spike mutation type"
                    ) +
  geom_segment( # line segments for significance
    data = data_comparisons_esc,
    aes(
      x = 1.0 + xoff + .03, 
      xend = 1.0 + xoff + .03,
      y = as.numeric(factor(Measure, levels = x_order)) + yoff1, 
      yend = as.numeric(factor(Measure, levels = x_order)) + yoff2
      #y = as.numeric(factor(Measure)) + y_delta, yend = as.numeric(factor(Measure)) - y_delta #midpoint +/- y_delta
    ),
    inherit.aes = FALSE
  ) +
  geom_text(
    data = data_comparisons_esc,
    aes(
      x = 1 + xoff +.04,
      y = as.numeric(factor(Measure, levels = x_order)) + (yoff1 + yoff2) / 2 + label_off,
      label = label,
    ),
    size = 10 / .pt, # 10 point font size
    vjust = 0.5,
    hjust = 0,
    inherit.aes = FALSE
  ) +
  xlim(-.02, 1.8) +
  scale_x_continuous(name = "relative rank", breaks =c(0,.5,1)) +
  scale_y_discrete(labels = c("Grammaticality (ESM2, unmasked)" = "Grammaticality\n(ESM2, unmasked)",
                              "Grammaticality (ESM2, masked)" = "Grammaticality\n(ESM2, masked)",
                              "Grammaticality (Hie)" = "Grammaticality\n(Hie)")) +
  theme_classic() + ylab("") + theme(legend.position = "top")
my_plot

fname = paste0('cov_all_measure_boxplot_escape_', Sys.Date(), '.png')
ggsave(
  file = fname, 
  plot = my_plot,
  width = 6.5, height = 5.5, units = "in"
)
#layer_data(plot = my_plot) #offset (y) is .1875

```

```{r}
custom_order_esc = c("escape", "non-escape", "not viable")
x_order = rev(c("Grammaticality (Hie)",
            "Grammaticality (ESM2, unmasked)",
            "Grammaticality (ESM2, masked)", 
            "MutComputeX",
            "MutRank",
            "Stability Oracle"
         ))

data_comparisons_esc_gramm = data_comparisons_esc %>%
  subset(!(Measure %in% c("Semantic change (Hie)", 
                          "Semantic change (ESM2)",
                          "PyRosetta",
                          "FoldX"))) %>% arrange(factor(Measure, levels = (x_order)))
data_comparisons_esc_sem = data_comparisons_esc %>%
  subset(Measure %in% c("Semantic change (Hie)", 
                          "Semantic change (ESM2)"))

nested_comb_data_long_esc_gramm = nested_comb_data_long_esc %>%
  subset(!(Measure %in% c("Semantic change (Hie)", 
                          "Semantic change (ESM2)",
                          "PyRosetta",
                          "FoldX"))) %>% arrange(factor(Measure, levels = (x_order)))
nested_comb_data_long_esc_sem = nested_comb_data_long_esc %>%
  subset(Measure %in% c("Semantic change (Hie)", 
                          "Semantic change (ESM2)"))

esc_gramm_plot = nested_comb_data_long_esc_gramm %>%
  pivot_longer(-Measure, names_to = "type", values_to = "value") %>% drop_na(value) |>
  ggplot(aes(x = value, y = factor(Measure, level = x_order), fill = factor(type, rev(custom_order_esc)))) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_manual(values = c("escape" = "#5D3A9B",
                                 "non-escape" = "#E66100", 
                                 "not viable" = "#AF596E"),
                      breaks = custom_order_esc,
                      name = "spike\nmutation type"
                    ) +
  geom_segment( # line segments for significance
    data = data_comparisons_esc_gramm,
    aes(
      x = 1.0 + xoff + .03, 
      xend = 1.0 + xoff + .03,
      y = as.numeric(factor(Measure, level = x_order)) + yoff1, 
      yend = as.numeric(factor(Measure, level = x_order)) + yoff2
      #y = as.numeric(factor(Measure)) + y_delta, yend = as.numeric(factor(Measure)) - y_delta #midpoint +/- y_delta
    ),
    inherit.aes = FALSE
  ) +
  geom_text(
    data = data_comparisons_esc_gramm,
    aes(
      x = 1 + xoff +.04,
      y = as.numeric(factor(Measure, level = x_order)) + (yoff1 + yoff2) / 2 + label_off,
      label = label,
    ),
    size = 10 / .pt, # 10 point font size
    vjust = 0.5,
    hjust = 0,
    inherit.aes = FALSE
  ) +
  xlim(-.02, 1.55) +
  scale_x_continuous(name = "relative rank", breaks =c(0,.5,1)) +
  scale_y_discrete(labels = c("Grammaticality (ESM2, unmasked)" = "Grammaticality\n(ESM2, unmasked)",
                              "Grammaticality (ESM2, masked)" = "Grammaticality\n(ESM2, masked)",
                              "Grammaticality (Hie)" = "Grammaticality\n(Hie)")) +
  theme_classic() + ylab("") + theme(legend.position = "top")
esc_gramm_plot

esc_sem_plot = nested_comb_data_long_esc_sem %>%
  pivot_longer(-Measure, names_to = "type", values_to = "value") %>% drop_na(value) |>
  ggplot(aes(x = value, y = factor(Measure), fill = factor(type, rev(custom_order_esc)))) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_manual(values = c("escape" = "#5D3A9B",
                                 "non-escape" = "#E66100", 
                                 "not viable" = "#AF596E"),
                      breaks = custom_order_esc,
                      name = "spike mutation type"
                    ) +
  geom_segment( # line segments for significance
    data = data_comparisons_esc_sem,
    aes(
      x = 1.0 + xoff + .03, 
      xend = 1.0 + xoff + .03,
      y = as.numeric(factor(Measure)) + yoff1, 
      yend = as.numeric(factor(Measure)) + yoff2
    ),
    inherit.aes = FALSE
  ) +
  geom_text(
    data = data_comparisons_esc_sem,
    aes(
      x = 1 + xoff +.04,
      y = as.numeric(factor(Measure)) + (yoff1 + yoff2) / 2 + label_off,
      label = label,
    ),
    size = 10 / .pt, # 10 point font size
    vjust = 0.5,
    hjust = 0,
    inherit.aes = FALSE
  ) +
  xlim(-.04, 1.55) +
  scale_x_continuous(name = "relative rank", breaks =c(0,.5,1)) +
  scale_y_discrete(labels = c("Semantic change (ESM2)" = "Semantic change\n(ESM2)",
                              "Semantic change (Hie)" = "Semantic change\n(Hie)")
                   ) +
  theme_classic() + ylab("") + theme(legend.position = "none")
esc_sem_plot
```
