---
title: "dadonaite_Delta_viz.Rmd"
output: html_document
date: '2024-11-12'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE}
library(ggplot2)
library(ggridges)
library(viridis)
library(plyr)
library(tidyr)
library(reshape2)
library(dplyr)
library(cowplot)
library(ggpubr)
library(purrr)

# Load required library
library(Biostrings)

# Function to read a FASTA file and convert the protein sequence to a dataframe
convert_protein_fasta_to_dataframe <- function(fasta_file) {
  # Read the FASTA file
  fasta <- readAAStringSet(fasta_file)
  
  # Extract the sequence as a string
  sequence <- as.character(fasta[[1]])
  
  # Split the sequence into individual amino acids
  amino_acids <- strsplit(sequence, split = "")[[1]]
  
  # Convert the amino acids into a dataframe
  df <- data.frame(position = 1:length(amino_acids), amino_acid = amino_acids, stringsAsFactors = FALSE)
  
  return(df)
}

# Example usage
# Replace "your_protein_sequence.fasta" with the path to your FASTA file
fasta_file <- "/Users/beallma/Documents/GitHub/NLP_viral_escape/data/cov/dadonaite/Delta/Delta.fasta"
wt_mutations <- convert_protein_fasta_to_dataframe(fasta_file)
wt_mutations = wt_mutations
# View the dataframe
print(wt_mutations)

# Load the required library
library(Biostrings)

# Define the function
create_locus_map <- function(fasta_file, reference_sequence_name) {
  # Read the FASTA file
  alignment <- readAAStringSet(fasta_file)
  
  # Ensure the reference sequence exists
  if (!reference_sequence_name %in% names(alignment)) {
    stop("Reference sequence not found in the FASTA file.")
  }
  
  # Extract sequences
  ref_seq <- alignment[[reference_sequence_name]]
  other_seq_name <- setdiff(names(alignment), reference_sequence_name)
  
  if (length(other_seq_name) != 1) {
    stop("The alignment must contain exactly two sequences.")
  }
  
  other_seq <- alignment[[other_seq_name]]
  
  # Initialize variables for mapping
  ref_pos <- 0
  other_pos <- 0
  result <- data.frame(ReferenceAA = character(), ReferenceLocus = integer(),
                       OtherLocus = integer(), stringsAsFactors = FALSE)
  
  # Iterate through both sequences
  for (i in seq_along(ref_seq)) {
    ref_aa <- as.character(ref_seq[i])
    other_aa <- as.character(other_seq[i])
    
    # Update loci if the character is not a gap
    if (ref_aa != "-") {
      ref_pos <- ref_pos + 1
    }
    if (other_aa != "-") {
      other_pos <- other_pos + 1
    }
    
    # Store mapping if the reference has a valid amino acid
    if (ref_aa != "-") {
      result <- rbind(result, data.frame(ReferenceAA = ref_aa,
                                         ReferenceLocus = ref_pos,
                                         OtherLocus = ifelse(other_aa != "-", other_pos, NA)))
    }
  }
  
  return(result)
}
fasta_file <- "/Users/beallma/Documents/GitHub/NLP_viral_escape/data/cov/dadonaite/Delta/Delta_alignment.fasta"
my_locus_map = create_locus_map(fasta_file = fasta_file, 
                                reference_sequence_name = "Delta_Dadonaite")
my_locus_map = my_locus_map %>%
  mutate(Dado_AA_Delta_loc = paste0(ReferenceAA, OtherLocus),
         Dado_AA_Dado_loc = paste0(ReferenceAA, ReferenceLocus))
my_locus_map
```
## Basic Functions

```{r}
three_to_single <- function(three_letter) {
  aa_map <- list(
    ALA = "A",
    ARG = "R",
    ASN = "N",
    ASP = "D",
    CYS = "C",
    GLN = "Q",
    GLU = "E",
    GLY = "G",
    HIS = "H",
    ILE = "I",
    LEU = "L",
    LYS = "K",
    MET = "M",
    PHE = "F",
    PRO = "P",
    SER = "S",
    THR = "T",
    TRP = "W",
    TYR = "Y",
    VAL = "V"
    # You can add more amino acids if needed
  )
  
  return(aa_map[[three_letter]])
}

aa_alphabet = c("A","R","N","D","C",
                "Q","E","G","H","I",
                "L","K","M","F","P",
                "S","T","W","Y","V")

get_seqs = function(file_name){
  fasta_file <- file_name
  
  # Read the lines of the FASTA file
  fasta_lines <- readLines(fasta_file)
  
  # Extract the sequence information
  sequence_info <- grep("^>", fasta_lines, value = TRUE)
sequence_name <- gsub(">", "", sequence_info)
sequence <- sapply(seq_along(sequence_info), function(i) {
  start <- grep(sequence_info[i], fasta_lines)[1] + 1
  end <- ifelse(i == length(sequence_info), length(fasta_lines), grep(sequence_info[i + 1], fasta_lines)[1] - 1)
  paste(fasta_lines[start:end], collapse = "")
}, simplify = TRUE)

# Create a named vector with one entry
fasta_data <- setNames(sequence, sequence_name)
  return(fasta_data)
}

zscore = function(df = c()){(df-mean(df))/sd(df)}
```

# Escape Scores

```{r}
setwd("/Users/beallma/Documents/GitHub/NLP_viral_escape/data/")

fname = "cov/dadonaite/Delta/267C_avg.csv"
df_267C = read.delim(file = fname, header = TRUE, sep=",",quote = "\"",
                 dec = ".", fill = TRUE, comment.char = "#") %>%
  mutate(antibody = "267C")

fname = "cov/dadonaite/Delta/279C_avg.csv"
df_279C = read.delim(file = fname, header = TRUE, sep=",",quote = "\"",
                 dec = ".", fill = TRUE, comment.char = "#") %>%
  mutate(antibody = "279C")

fname = "cov/dadonaite/Delta/REGN10933_avg.csv"
df_REGN10933 = read.delim(file = fname, header = TRUE, sep=",",quote = "\"",
                 dec = ".", fill = TRUE, comment.char = "#") %>%
  mutate(antibody = "REGN10933")

names(df_267C)
head(df_267C)

all_esc_scores = df_267C %>% 
  full_join(df_279C) %>% 
  full_join(df_REGN10933) %>%
  select(c(epitope, site, wildtype, mutant, mutation, escape_median, escape_std, antibody)) %>%
  mutate(escape = case_when(
    (is.na(escape_std)) & (escape_median > 0) ~ "escape",
    (is.na(escape_std)) & (escape_median < 0) ~ "non-escape",
    (abs(escape_median) - (2*escape_std) > 0) & (escape_median > 0)   ~ "escape",
    (abs(escape_median) - (2*escape_std) > 0) & (escape_median < 0)   ~ "non-escape",
    TRUE ~ "non-escape"
  )) %>%
  nest(data = -antibody) %>%
  mutate(data = map(data, ~ .x %>% mutate(ranked_score = rank(escape_median))),
         plot = map2(data, antibody, ~ ggplot(.x, aes(x = ranked_score, y = escape_median)) +
                  geom_point() +
                  labs(title = paste("Antibody:", .y, 
                                     nrow(.x %>% subset(escape == "escape")),
                                     "escape mutations"), 
                       x = "Rank", y = "Escape Median"))
         )

head(all_esc_scores)
tail(all_esc_scores)
all_esc_scores$plot
esc_scores = df_267C %>% 
  full_join(df_279C) %>% 
  full_join(df_REGN10933) %>%
  dplyr::rename("ID" = "mutation") %>%
  mutate(escape = case_when(
    (is.na(escape_std)) & (escape_median > 0) ~ "escape",
    (is.na(escape_std)) & (escape_median < 0) ~ "non-escape",
    (abs(escape_median) - (2*escape_std) > 0) & (escape_median > 0)   ~ "escape",
    (abs(escape_median) - (2*escape_std) > 0) & (escape_median < 0)   ~ "non-escape",
    TRUE ~ "non-escape"
  ))
esc_scores

esc_overall = esc_scores %>% 
  subset((mutant != "*") | (mutant != "-")) %>%
  select(ID, antibody, escape) %>%
  group_by(ID) %>%
  summarise(
    escape_status = if_else(any(escape == "escape"), "escape", "non-escape"),
    antibodies = paste(unique(antibody), collapse = ", ")
  ) %>%
  ungroup()
esc_overall
```

# Functional Scores
```{r}

Delta_thresh = -1.46 # https://dms-vep.org/SARS-CoV-2_Delta_spike_DMS_REGN10933/267C_escape_plot.html
setwd("/Users/beallma/Documents/GitHub/NLP_viral_escape/data/")
fname = "cov/dadonaite/Delta/muteffects_observed.csv"
func_scores = read.delim(file = fname, header = TRUE, sep=",",quote = "\"",
                 dec = ".", fill = TRUE, comment.char = "#")%>%
  mutate(ID = paste0(wildtype, sequential_site, mutant),
         wt = case_when(
           is.na(times_seen) ~ TRUE,
           TRUE ~ FALSE
         ),
         viable = case_when(
           (effect - (2*effect_std) >= Delta_thresh) & (effect >= Delta_thresh)   ~ "viable",
           (effect + (2*effect_std) >= Delta_thresh) & (effect <= Delta_thresh)  ~ "viable",
           TRUE ~ "not viable"
         )) %>% 
  select(c(-Lib.1.effect, -Lib.2.effect, -Lib.3.effect, -Lib.4.effect, -phenotype, -n_libraries, -times_seen))
func_scores
```

```{r}
# now let us include mutations that were not tested at all
all_sites_escape_viable = 
  func_scores %>% subset((mutant != "*") || mutant != "-") %>%
  select(c("effect", "ID", "wt","reference_site", "viable", "sequential_site")) %>%
  left_join(esc_overall, by = "ID") %>% 
  mutate(site = as.integer(ifelse(is.na(ID), 
                                  substr(ID, 1,nchar(ID)-1), sequential_site))) %>%
  mutate(wt = ifelse(is.na(wt), 
                     sapply(site, function(x) wt_mutations$mutations[x]),
                     wt)) %>%
  mutate(ID = ifelse(is.na(ID),
                     paste0(substr(wt, nchar(wt), nchar(wt)), mutations),
                     ID)) %>%
  mutate(dadonaite_esc = ifelse(is.na(escape_status), "not tested", escape_status)) %>%
  select(-escape_status)
all_sites_escape_viable
all_sites_escape_viable %>% subset(is.na(viable)) #empty
all_sites_escape_viable %>% subset((viable == "not viable") & dadonaite_esc == "escape") # 144
all_sites_escape_viable %>% subset((viable == "not viable") & (dadonaite_esc == "non-escape")) # 565


```

# Hie Bespoke Model
Read in and process the Hie pretrained bespoke model data

```{r}
fname = "/Users/beallma/OneDrive - Emory University/Documents/PIPP/SARS_NLP/aggregate_data_vizualization/cov/cov_Delta_screen_wc03.log"

df_Hie_dado = read.delim(file = fname, header = TRUE, sep="",quote = "\"",
                 dec = ".", fill = TRUE, comment.char = "#")
df_Hie_dado = df_Hie_dado %>% dplyr::rename("Semantic.change_Hie" = "Semantic.change",
               "Grammaticality_Hie_logit" = "Grammaticality") %>%
  select(-Mutations, -Sequence)
```


#ESM-2 data

```{r}
fname = "/Users/beallma/OneDrive - Emory University/Documents/PIPP/SARS_NLP/aggregate_data_vizualization/cov/cov_Delta_ESM2.log"

df_ESM2_dado = read.delim(file = fname, header = TRUE, sep="",quote = "\"",
                 dec = ".", fill = TRUE, comment.char = "#")
df_ESM2_dado = df_ESM2_dado %>% dplyr::rename("Semantic.change_ESM2" = "Semantic.change",
                "Grammaticality_ESM2_unmasked" = "Grammaticality") %>% 
  select(-Mutations, -Sequence)
#high grammar means more like the original sequence, so we want to invert^ this to match semantic change

```

Read in and process the 
#ESM-2 masked data

```{r}
setwd("/Users/beallma/Documents/GitHub/NLP_viral_escape/language_models/results")

fname = "cov/logits_Delta_ESM2_650M_masked.csv"
df_ESM2_dado_masked = read.delim(file = fname, header = TRUE, sep=",",quote = "\"",
                 dec = ".", fill = TRUE, comment.char = "#")

df_ESM2_dado_masked <- df_ESM2_dado_masked %>%
  pivot_longer(cols = -X,
               names_to = "Amino_Acid",
               values_to = "prob_ESM2_masked")

wt_mutations = wt_mutations %>% 
  mutate(AA = paste0(amino_acid, position))

df_ESM2_dado_masked <- df_ESM2_dado_masked %>%
  mutate(partial_id = paste0(X, Amino_Acid))

df_ESM2_dado_masked <- df_ESM2_dado_masked %>%
  left_join(wt_mutations, by = 
               c("X" = "position")) %>%
  mutate(
    ID = paste0(AA, Amino_Acid),
    mutant = case_when(substr(ID, nchar(ID), nchar(ID)) == substring(ID, 1, 1) ~ "wt",
                   TRUE ~ "mutant")) %>%
  select(-X, -Amino_Acid, -partial_id, -AA)
df_ESM2_dado_masked
```

Read in and process the 
#MutComputeX data

```{r}
setwd("/Users/beallma/Documents/GitHub/NLP_viral_escape/structural_models/results")

fname = "cov/delta-af2-trunc_mutcomputex.csv"
df_mutcomp = read.delim(file = fname, header = TRUE, sep=",",quote = "\"",
                 dec = ".", fill = TRUE, comment.char = "#")
names(df_mutcomp)
head(df_mutcomp)
df_mutcomp <- df_mutcomp %>%
  rename(prAA = "mut_pred",
         pred_prob = "mut_pred_prob") %>%
  pivot_longer(cols = starts_with("pr"),
               names_to = "Amino_Acid",
               values_to = "prob_MutCompX") %>%
  select(chain_id, pos, wtAA, mut_pred, Amino_Acid, prob_MutCompX) %>%
  mutate(
    Amino_Acid = gsub("pr", "", Amino_Acid),
    Amino_Acid = sapply(Amino_Acid, three_to_single),
    wtAA = sapply(wtAA, three_to_single),
    mut_pred = sapply(mut_pred, three_to_single),
    Mutations = paste(wtAA, pos, Amino_Acid, sep = "")
  ) %>% 
  filter(chain_id == "A") %>%
  select(Mutations,prob_MutCompX)
#high probability means more like the original sequence, so we want to invert^ this to match semantic change
df_mutcomp
```

Read in and process the 
#MutRank data

```{r}
setwd("/Users/beallma/Documents/GitHub/NLP_viral_escape/structural_models/results")

fname = "cov/delta-af2-trunc_mutrank.csv"
df_mutrank = read.delim(file = fname, header = TRUE, sep=",",quote = "\"",
                 dec = ".", fill = TRUE, comment.char = "#")
names(df_mutrank)
head(df_mutrank)
df_mutrank <- df_mutrank %>%
  rename(pred_AA = "mut_pred",
         pred_prob = "mut_pred_prob") %>%
  pivot_longer(cols = starts_with("mean"),
               names_to = "Amino_Acid",
               values_to = "prob_MutRank") %>%
  select(chain_id, res_seq_num, wt_AA, mut_pred, Amino_Acid, prob_MutRank) %>%
  mutate(
    Amino_Acid = gsub("mean_", "", Amino_Acid),
    Amino_Acid = sapply(Amino_Acid, three_to_single),
    wt_AA = sapply(wt_AA, three_to_single),
    mut_pred = sapply(mut_pred, three_to_single),
    Mutations = paste(wt_AA, res_seq_num, Amino_Acid, sep = "")
  ) %>% 
  filter(chain_id == "A") %>%
  select(Mutations,prob_MutRank)
#high probability means more like the original sequence, so we want to invert^ this to match semantic change
df_mutrank
```

Read in and process the 
#StabilityOracle data

```{r}
setwd("/Users/beallma/Documents/GitHub/NLP_viral_escape/structural_models/results")
fname = "cov/delta-af2-trunc_stability-oracle.csv"

df_stab_ora = read.delim(file = fname, header = TRUE, sep=",",quote = "\"",
                 dec = ".", fill = TRUE, comment.char = "#")

df_stab_ora <- df_stab_ora %>%
  pivot_longer(cols = starts_with("mean_"),
               names_to = "Amino_Acid",
               values_to = "ddG_Stab_Ora") %>%
  select(chain_id, res_seq_num, wt_AA, pred_AA, Amino_Acid, ddG_Stab_Ora) %>%
  mutate(
    Amino_Acid = gsub("mean_", "", Amino_Acid),
    Amino_Acid = sapply(Amino_Acid, three_to_single),
    wt_AA = sapply(wt_AA, three_to_single),
    pred_AA = sapply(pred_AA, three_to_single),
    Mutations = paste(wt_AA, res_seq_num, Amino_Acid, sep = "")
  ) %>% 
  filter(chain_id == "A") %>%
 select(Mutations,ddG_Stab_Ora)
df_stab_ora
```

# EVEscape

```{r}
setwd("/Users/beallma/Documents/GitHub/NLP_viral_escape/data/cov/evescape")
#fname = "Wuhan(A)_single_mutation_predictions.csv"
#fname = "full_Delta_evescape.csv"
fname = "Delta(AY.4.2)_single_mutation_predictions.csv"

df_eves = read.delim(file = fname, header = TRUE, sep=",",quote = "\"",
                 dec = ".", fill = TRUE, comment.char = "#")

result_string <- character(max(df_eves$pos))

# Place the letters in the appropriate positions
result_string[df_eves$pos] <- df_eves$wt

# Combine the characters into a single string
final_string <- paste0(result_string, collapse = "")

# Print the result
print(final_string)

df_eves = df_eves %>%
  #arrange(i) %>%
  mutate(ID = paste0(wt, pos, mut),
         wt_Delta = paste0(wt,pos)) %>%
  select(ID, EVEscape, fitness_evol_indices, wt_Delta,mut) %>%
  rename("fitness_evol_indices" = "eve") %>%
  rename("EVEscape" = "evescape") %>%
  left_join(my_locus_map, by = c("wt_Delta" = "Dado_AA_Delta_loc")) %>%
  mutate(Dado_loc_Delt_mut = paste0(Dado_AA_Dado_loc, mut)) %>%
  select(c(ID, wt_Delta, Dado_loc_Delt_mut, eve, evescape))
  
  #select(mutations, EVEscape)
df_eves 
```

# Pie Chart
```{r}
library(ggforce)
comb_esc_via = all_sites_escape_viable %>%
  full_join(df_Hie_dado, by ="ID") %>% # to include all possible mutations
  replace_na(list(dadonaite_esc = "not tested", viable = "not tested")) %>% 
  mutate(dadonaite_esc = ifelse((viable == "not viable") & 
                                  (dadonaite_esc == "not tested"), 
                                "not viable", dadonaite_esc),
         dadonaite_esc = ifelse((viable == "viable") & 
                                  (dadonaite_esc == "not tested"), 
                                "viable", dadonaite_esc)
         )
count_comb_cov = comb_esc_via %>% group_by(dadonaite_esc) %>%
  summarize(
    count = n()
  )


mut_order = c("escape", "non-escape", "viable", 
              "not viable","not tested")
pie_data_cov <- count_comb_cov %>%
  arrange(match(dadonaite_esc, mut_order)) %>% # sort so pie slices end up sorted
  mutate(
    end_angle = 2*pi*cumsum(count)/sum(count),   # ending angle for each pie slice
    start_angle = lag(end_angle, default = 0),   # starting angle for each pie slice
    mid_angle = 0.5*(start_angle + end_angle),   # middle of each pie slice, for text labels
    # horizontal and vertical justifications for outer labels
    hjust = ifelse(mid_angle > pi, 1, 0),
    vjust = ifelse(mid_angle < pi/2 | mid_angle > 3*pi/2, 0, 1)
  )
pie_data_cov

pie_data_cov$hjust[1] = 1
pie_data_cov$vjust[4] = 1
pie_data_cov$vjust[3] = 0
pie_data_cov$mid_angle[3] = 1.1645
pie_data_cov$mid_angle[5] = (2*pi) - pie_data_cov$mid_angle[4] -.3
pie_data_cov$vjust[5] = 1


pie_data_cov

cov_esc_pie = ggplot(pie_data_cov) +
    aes(
      x0 = 0, y0 = 0,
      r0 = 0.4, r = 1,
      amount = count,
      start = start_angle, end = end_angle,
      fill = dadonaite_esc
    ) +
  geom_arc_bar(alpha = 0.7) +
  geom_text(
    aes(x = 1.1 * sin(mid_angle), 
        y = 1.1 * cos(mid_angle), 
        label = count,
        hjust = hjust,
        vjust = vjust)#vjust)
  ) +
  scale_fill_manual(values = c("escape" = "#5D3A9B",
                                 "non-escape" = "#E66100",
                                 "viable" = "#399c52",
                                 "not viable" = "#AF596E",
                                 "not tested" = "#C5C5C5"),
                    name = "spike\nmutation type",
                    breaks = mut_order) +
  coord_fixed() +
  xlim(-1, 1.3) +
  ylim(-1.1, 1.4) +
  theme_void() +
  theme(
    strip.text = element_blank(),
    strip.background = element_blank(),
    legend.key=element_rect(color="white"),
    legend.position = "none"
  ) + 
  geom_segment(aes(
      x = 1 * sin(mid_angle), 
      xend = 1.1 * sin(mid_angle),
      y = 1 * cos(mid_angle), 
      yend = 1.1 * cos(mid_angle)
    )) + 
  annotate("text", x = 0, y = 0, label = "Delta", size = 8, fontface = "bold")
cov_esc_pie

fname = paste0('cov_Dadonaite_Delta_DMS_noindel_piechart_', Sys.Date(), '.png')
ggsave(
  file = fname, 
  plot = cov_esc_pie,
  width = 4, height = 3.5, units = "in"
)

mut_order = c("escape", "non-escape", "viable", 
              "not viable","not tested")
#layer_data(plot = cov_esc_pie)
```

# Scatter w/ Marginal
```{r}
library("ggExtra")

comb_esc_via = comb_esc_via %>% 
  mutate(dadonaite_esc = 
           factor(dadonaite_esc, levels = rev(mut_order))) %>%
  arrange(dadonaite_esc)

#comb_esc_via$dadonaite_esc <- factor(comb_esc_via$dadonaite_esc, levels = (mut_order))
p_cov_scatter= ggplot() + #ggplot(subset(comb_data, Starr_esc != "not viable")) +
  geom_point( data =comb_esc_via %>% filter(dadonaite_esc == "not tested"), 
              aes(x = Grammaticality_Hie_logit, y = Semantic.change_Hie, 
                 color = dadonaite_esc),
             alpha = 0.5, size = 1.5) +
  geom_point( data =comb_esc_via %>% filter(dadonaite_esc == "not viable"), 
              aes(x = Grammaticality_Hie_logit, y = Semantic.change_Hie, 
                 color = dadonaite_esc),
             alpha = 0.5, size = 1.5) +
  geom_point( data =comb_esc_via %>% filter(dadonaite_esc == "non-escape"), 
              aes(x = Grammaticality_Hie_logit, y = Semantic.change_Hie, 
                 color = dadonaite_esc),
             alpha = 0.5, size = 1.5) +
  geom_point( data =comb_esc_via %>% filter(dadonaite_esc == "escape"), 
              aes(x = Grammaticality_Hie_logit, y = Semantic.change_Hie, 
                 color = dadonaite_esc),
             alpha = 0.5, size = 1.5) +
  scale_color_manual(values = c("escape" = "#5D3A9B",
                                 "non-escape" = "#E66100", 
                                 "not viable" = "#AF596E",
                                 "not tested" = "#C5C5C5")) +
  # geom_point(data = subset(comb_data, Baum_esc == "escape"), 
  #            aes(x = Grammaticality_Hie_logit, y = Semantic.change_Hie),
  #                shape = 4, color = "black", size = 1.5, stroke = .75, alpha = .7)  +
  theme_bw() + theme(legend.position = "none",#"top",
                     #aspect.ratio = 1,
                     legend.title = element_blank()
                     ) + 
  ylab("Semantic Change") +
  xlab("Grammaticality")

p_cov_scatter_ungrouped= ggplot() + #ggplot(subset(comb_data, Starr_esc != "not viable")) +
  geom_point( data =comb_esc_via %>% subset(dadonaite_esc != "viable"), 
              aes(x = Grammaticality_Hie_logit, y = Semantic.change_Hie, 
                 color = dadonaite_esc),
             alpha = 0.5, size = 1) +
  scale_color_manual(values = c("escape" = "#5D3A9B",
                                 "non-escape" = "#E66100", 
                                 "not viable" = "#AF596E",
                                 "not tested" = "#C5C5C5")) +
  # geom_point(data = subset(comb_data, Baum_esc == "escape"), 
  #            aes(x = Grammaticality_Hie_logit, y = Semantic.change_Hie),
  #                shape = 4, color = "black", size = 1.5, stroke = .75, alpha = .7)  +
  theme_bw() + theme(legend.position = "none",#"top",
                     #aspect.ratio = 1,
                     legend.title = element_blank()
                     ) + 
  ylab("Semantic Change") +
  xlab("Grammaticality")
p_cov_scatter_ungrouped

p_cov_scatter_ungrouped = ggMarginal(p_cov_scatter_ungrouped, 
                                     groupColour = TRUE, groupFill = TRUE,
                                     alpha = 0.0)
fname = paste0('cov_hie_scatter_escape_Delta_', Sys.Date(), '.png')
ggsave(
  file = fname, 
  plot = p_cov_scatter_ungrouped,
  width = 3.5, height = 3.5, units = "in"
)
```

#Pie with Scatter
```{r}
library(cowplot)
setwd("/Users/beallma/OneDrive - Emory University/Documents/PIPP/SARS_NLP/aggregate_data_vizualization")
#renv::init()
```

```{r}
#mywd = "/Users/beallma/OneDrive - Emory University/Documents/PIPP/SARS_NLP/aggregate_data_vizualization"
setwd("/Users/beallma/OneDrive - Emory University/Documents/PIPP/SARS_NLP/aggregate_data_vizualization")
cov_esc_pie
p_cov_scatter_ungrouped

prow <- plot_grid(
  p_cov_scatter_ungrouped, cov_esc_pie + theme(legend.position="none"),
  align = 'h',
  labels = c("A", "B"),
  hjust = -1,
  nrow = 1,
  rel_widths = c(1, 1)
)
prow

legend_b <- get_legend(
  cov_esc_pie + 
    guides(color = guide_legend(nrow = 1)) +
    theme(legend.position = "bottom")
)
scatter_pie = plot_grid(prow, legend_b, 
          ncol = 1,
          rel_heights = c(1, .15))
scatter_pie
fname = paste0('Delta_scatter_escape_piechart_', Sys.Date(), '.png')
ggsave(
  file = fname,
  plot = scatter_pie + theme(plot.background = element_rect(fill = "white", linetype = "blank"),
                             panel.border = element_blank()),
  width = 5.3, height = 3.8, units = "in"
)
```

# Combine all datastreams into one dataframe

```{r}

all_sites_escape_viable 
 # contains the grammaticality and semantic change measures from the Hie model
           # large gramm -> closer to wt -> large rank
           # small sem change -> close to wt -> supposed less likely escape -> smaller rank
df_stab_ora # ddG from stability Oracle. More negative -> more stable -> smaller rank; more positive -> less stable -> larger rank
df_mutcomp # large probability -> more stable -> larger rank
df_mutrank # large log odds --> more stable --> larger rank
df_ESM2_dado # probability from masked ESM model -> larger probability -> more grammatical -> larger rank
df_ESM2_dado_masked # semantic change and grammar using the ESM model in an unmasked framework

comb_data = df_Hie_dado  %>% 
  left_join(all_sites_escape_viable, by = c('ID' = 'ID')) %>%
  left_join(df_ESM2_dado, by = c('ID' = 'ID')) %>%
  left_join(df_ESM2_dado_masked, by = c('ID' = 'ID')) %>%
  left_join(df_stab_ora, by = c('ID' = 'Mutations')) %>%
  left_join(df_mutcomp, by = c('ID' = 'Mutations')) %>%
  left_join(df_mutrank, by = c('ID' = 'Mutations')) %>%
  left_join(df_eves, by = c("ID" = "Dado_loc_Delt_mut")) %>%
  replace_na(list(dadonaite_esc = "not tested", viable = "not tested"))

my_measures = colnames(comb_data %>% 
                         select(-c(ID, reference_site, antibodies,
                                wt,sequential_site, effect,
                                mutant, dadonaite_esc, viable, amino_acid, site,
                                wt_Delta, ID.y, site))
                       )
my_measures
comb_data

rev_rank_measures = c("ddG_Stab_Ora")

comb_data_norm = comb_data %>% 
  mutate_at(my_measures, function(x) rank(x, na.last = "keep")) %>%
  mutate_at(rev_rank_measures, ~max(rank(., na.last = "keep"), na.rm=T) + 1 - rank(.,na.last = "keep") ) %>%
  mutate_at(my_measures, function(x) x/max(x, na.rm = TRUE))

comb_data_norm

mut_by_site = c("ddG_Stab_Ora", "prob_MutCompX", "prob_ESM2_masked", "prob_MutRank")

comb_data_mut_by_site = comb_data %>%
  group_by(site) %>%
  mutate_at(mut_by_site,
            function(x) rank(x, na.last = "keep")) %>%
  mutate_at(mut_by_site, function(x) rank(x, na.last = "keep")) %>%
  select(ID, mutant, site, ddG_Stab_Ora, prob_MutCompX, prob_ESM2_masked, evescape, eve)
  
comb_data_long = comb_data_norm %>% 
  select(-effect, -mutant, -wt, -reference_site, -site, 
         -sequential_site, -antibodies, -amino_acid, -ID.y, -wt_Delta) %>%
  pivot_longer(cols = my_measures ,
               names_to = "Measure",
               values_to = "rank") %>%
  mutate(Measure = 
           case_when(
             Measure == "ddG_Stab_Ora" ~ "Stability Oracle",
             Measure == "prob_MutCompX" ~ "MutComputeX",
             Measure == "prob_MutRank" ~ "MutRank",
             Measure == "prob_ESM2_masked" ~ "Grammaticality (ESM2, masked)",
             Measure == "Semantic.change_Hie" ~ "Semantic change (Hie)",
             Measure == "Grammaticality_Hie_logit" ~ "Grammaticality (Hie)",
             Measure == "Semantic.change_ESM2" ~ "Semantic change (ESM2)",
             Measure == "Grammaticality_ESM2_unmasked" ~ "Grammaticality (ESM2, unmasked)",
             Measure == "evescape" ~ "EVEscape",
             Measure == "eve" ~ "EVE"
             )
         )
comb_data_long

```
# Boxplots viability

Now I want to recreate the cov bar charts. The issue is choosing which measure of viability to use

Let's start with the coloring the plot by viability

prepare the data

```{r}
comb_data_norm
comb_data_long

comb_data_long_gramm = comb_data_long

x_order = rev(c("Grammaticality (Hie)",
            "Grammaticality (ESM2, unmasked)",
            "Grammaticality (ESM2, masked)", 
            "MutComputeX",
            "MutRank",
            "Stability Oracle",
            "EVE",
            "EVEscape",
            "Semantic change (Hie)",
            "Semantic change (ESM2)"
         ))

nested_comb_data_long_esc_eff_tol = comb_data_long %>% 
  subset(viable != "not tested") %>% 
  pivot_wider(names_from = "viable", values_from = "rank") |>
  arrange(factor(Measure, levels = x_order))

fname = paste0('Delta_boxplots_viability_tested_', Sys.Date(), '.png')

nested_comb_data_long_esc_temp = nested_comb_data_long_esc_eff_tol %>% nest(data = -Measure)

nested_comb_data_long_esc_eff_tol
```


```{r}

data_stats_gramm <- nested_comb_data_long_esc_eff_tol |>
  nest(data = -Measure) %>%
  mutate(
    via_non_via = map(data, ~broom::tidy(wilcox.test(.x$viable, .x$`not viable`)))
    
  ) |>
  pivot_longer(
    cols = c(via_non_via),
    names_to = "test",
    values_to = "stats"
  ) |>
  mutate(
    column1 = case_when(
      test == "via_non_via" ~ "viable"
    ),
    column2 = case_when(
      test == "via_non_via" ~ "not viable"
    )
  ) |>
  unnest_wider(stats) |>
  select(Measure, column1, column2, p.value)

y_delta = 0#.25 # distance between adjacent bloxplots (.8/number of boxplots)
x_delta = 0#.185 # distance between line segments
data_comparisons_gramm <- data_stats_gramm |>
  mutate(
    # apply overall false-discovery rate correction
    p.value = p.adjust(p.value, method = "fdr"),
    # set stars
    label = case_when(
      p.value < 0.001 ~ "***",
      p.value < 0.01 ~ "**",
      p.value < 0.05 ~ "*",
      TRUE ~ "NS"
    ),
    # offset to correct for stars not being vertically centered
    label_off = case_when(
      label == "NS" ~ 0,
      TRUE ~ -.04,
    ),
    # calculate x and y offsets for line segments
    yoff1 = case_when(
      column1 == "viable" ~ y_delta,
      column1 == "not viable" ~ -y_delta,
      TRUE ~ 0
    ),
    yoff2 = case_when(
      column2 == "viable" ~ y_delta,
      column2 == "not viable" ~ -y_delta,
      TRUE ~ 0
    ),
    xoff = case_when(
      column1 == "viable" & column2 == "not viable" ~ 0,
      column1 == "not viable" & column2 == "viable" ~ x_delta,
      TRUE ~ 2 * x_delta
    )
  )

data_comparisons_via_gramm = data_comparisons_gramm %>% 
  subset(!(Measure %in% c("Semantic change (Hie)", 
                          "Semantic change (ESM2)",
                          "EVEscape",
                          "PyRosetta",
                          "FoldX"))) %>% arrange(factor(Measure, levels = (x_order)))
data_comparisons_via_sem = data_comparisons_gramm %>%
  subset(Measure %in% c("Semantic change (Hie)", 
                          "Semantic change (ESM2)",
                        "EVEscape"))

nested_comb_data_long_via_gramm = nested_comb_data_long_esc_temp %>%
  subset(!(Measure %in% c("Semantic change (Hie)", 
                          "Semantic change (ESM2)",
                          "EVEscape",
                          "PyRosetta",
                          "FoldX"))) %>% arrange(factor(Measure, levels = (x_order)))
nested_comb_data_long_via_sem = nested_comb_data_long_esc_temp %>%
  subset(Measure %in% c("Semantic change (Hie)", 
                          "Semantic change (ESM2)",
                        "EVEscape"))

custom_order_via = c("viable", "not viable")

x_order_gramm = rev(c("Grammaticality (Hie)",
            "Grammaticality (ESM2, unmasked)",
            "Grammaticality (ESM2, masked)", 
            "MutComputeX",
            "MutRank",
            "Stability Oracle",
            "EVE"
         ))

nested_comb_data_long_esc_temp
nested_comb_data_long_esc_eff_tol

via_gramm_plot = nested_comb_data_long_esc_eff_tol %>% select(c(Measure, viable, `not viable`)) %>%#select(-c(ID_HXB2,ID, any_ab_ESC, res_seq_num, ESC_via_thresh0005)) %>%
  subset(!(Measure %in% c("Semantic change (Hie)", 
                          "Semantic change (ESM2)",
                          "EVEscape",
                          "PyRosetta",
                          "FoldX"))) %>% 
  arrange(factor(Measure, levels = (x_order_gramm))) %>%
  pivot_longer(-Measure, names_to = "type", values_to = "value") %>% 
  drop_na(value) |>
  ggplot(aes(x = value, y = factor(Measure, level = x_order_gramm), fill = factor(type, rev(custom_order_via)))) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_manual(values = c("viable" = "#399c52",
                                 "not viable" = '#AF596E' #"#cd6a83", 
                                 ),
                      breaks = custom_order_via,
                      name = "spike\nmutation type"
                    ) +
  geom_segment( # line segments for significance
    data = data_comparisons_via_gramm,
    aes(
      x = 1.0 + xoff + .03, 
      xend = 1.0 + xoff + .03,
      y = as.numeric(factor(Measure, level = x_order_gramm)) + yoff1, 
      yend = as.numeric(factor(Measure, level = x_order_gramm)) + yoff2
      #y = as.numeric(factor(Measure)) + y_delta, yend = as.numeric(factor(Measure)) - y_delta #midpoint +/- y_delta
    ),
    inherit.aes = FALSE
  ) +
  geom_text(
    data = data_comparisons_via_gramm,
    aes(
      x = 1 + xoff +.04,
      y = as.numeric(factor(Measure, level = x_order_gramm)) + (yoff1 + yoff2) / 2 + label_off,
      label = label,
    ),
    size = 10 / .pt, # 10 point font size
    vjust = 0.5,
    hjust = 0,
    inherit.aes = FALSE
  ) +
  xlim(-.02, 1.55) +
  scale_x_continuous(name = "relative rank", breaks =c(0,.5,1)) +
  scale_y_discrete(labels = c("Grammaticality (ESM2, unmasked)" = "Grammaticality\n(ESM2, unmasked)",
                              "Grammaticality (ESM2, masked)" = "Grammaticality\n(ESM2, masked)",
                              "Grammaticality (Hie)" = "Grammaticality\n(Hie)")) +
  # scale_y_discrete(labels = c("Grammaticality (ESM2)" = "Grammaticality\n(ESM2, unmasked)",
  #                             "Grammaticality (Hie)" = "Grammaticality\n(Hie)")) +
  theme_classic() + ylab("") + theme(legend.position = "top")
via_gramm_plot
data_stats_gramm
data_comparisons_via_gramm
layer_data(via_gramm_plot)
```

Semantic change subplot for viability

```{r}
x_order_sem = rev(c("Semantic change (Hie)", "Semantic change (ESM2)", "EVEscape"))

via_sem_plot = nested_comb_data_long_esc_eff_tol %>% select(c(Measure, viable, `not viable`)) %>% 
  subset(Measure %in% c("Semantic change (Hie)", 
                          "Semantic change (ESM2)",
                        "EVEscape"
                        )) %>% 
  pivot_longer(-Measure, names_to = "type", values_to = "value") %>% drop_na(value) |> 
  arrange(factor(Measure, levels = (x_order_sem))) %>%
  ggplot(aes(x = value, y = factor(Measure, level = x_order_sem), fill = factor(type, rev(custom_order_via)))) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_manual(values = c("viable" = "#399c52",
                                 "not viable" = '#AF596E', #"#cd6a83", 
                                 "not tested" = "#C5C5C5"),
                      breaks = custom_order_via,
                      name = "spike\nmutation type"
                    ) +
  geom_segment( # line segments for significance
    data = data_comparisons_via_sem,
    aes(
      x = 1.0 + xoff + .03, 
      xend = 1.0 + xoff + .03,
      y = as.numeric(factor(Measure, level = x_order_sem)) + yoff1, 
      yend = as.numeric(factor(Measure, level = x_order_sem)) + yoff2
    ),
    inherit.aes = FALSE
  ) +
  geom_text(
    data = data_comparisons_via_sem,
    aes(
      x = 1 + xoff +.04,
      y = as.numeric(factor(Measure, level = x_order_sem)) + (yoff1 + yoff2) / 2 + label_off,
      label = label,
    ),
    size = 10 / .pt, # 10 point font size
    vjust = 0.5,
    hjust = 0,
    inherit.aes = FALSE
  ) +
  xlim(-.04, 1.55) +
  scale_x_continuous(name = "relative rank", breaks =c(0,.5,1)) +
  scale_y_discrete(labels = c("Semantic change (ESM2)" = "Semantic\nchange\n(ESM2)",
                              "Semantic change (Hie)" = "Semantic\nchange\n(Hie)")
                   ) +
  theme_classic() + ylab("") + theme(legend.position = "none")
via_sem_plot
```


```{r}
via_sem_plot_new_labs = via_sem_plot + 
  theme(legend.position="none")+ # plot.margin = margin(r = 0, l = 0, t= 0, b = 0.1, unit = "inches"))
  scale_y_discrete(labels = c("Semantic change (ESM2)" = "Semantic\nchange\n(ESM2)",
                              "Semantic change (Hie)" = "Semantic\nchange\n(Hie)")) + 
  scale_x_continuous(name = "relative rank", breaks =c(0,.5,1), limits = c(NA, 1.2))
via_sem_plot_new_labs

via_gramm_plot_new = via_gramm_plot + 
  theme(legend.position="none", plot.margin = margin(l = 0,t = 0.1, b = 0.1, r = 0.1, unit = "inches"))+#unit(c(0, 0, 0, 0), "inches"))  + 
  #xlim(-.02,2.5) +
  scale_x_continuous(name = "relative rank", breaks =c(0,.5,1), limits = c(NA, 1.2))
via_gramm_plot_new

legend_b <- get_legend(
  via_gramm_plot_new + theme(legend.position="right",
                             legend.key.width = unit(.6, "in"),
                             legend.key.height = unit(.1, "in")) +
    guides(color = guide_legend(nrow = 3,
                                title = "spike\nmutation type"))
)

aligned <- align_plots(via_gramm_plot_new+ xlab("relative rank"), 
                       via_sem_plot_new_labs+ xlab("relative rank"), 
                       align = 'h', axis = 'l')

sem_col <- plot_grid(
  aligned[[2]],#via_sem_plot_new_labs+ xlab("relative rank"),
  legend_b,
  NULL,
  labels = c("", "", ""),
  nrow = 3,
  rel_heights = c(1, .5, .35)
)
sem_col

gram_col = plot_grid(
  aligned[[1]] ,
  NULL,
  labels = c("", ""),
  nrow = 2,
  rel_heights = c(1, .05)
)

via_all_measures = plot_grid(
  gram_col,#aligned[[1]], 
  #NULL,
  sem_col, 
  ncol = 2,
  align = 'h',
  labels = c("A", "B"),
  label_x = 0.05, label_y = 0.95,
  hjust = -0.5, vjust = -0.5,
  rel_widths = c(1, .75))

via_all_measures

setwd("/Users/beallma/OneDrive - Emory University/Documents/PIPP/SARS_NLP/aggregate_data_vizualization")
fname = paste0('Delta_boxplots_viability_', Sys.Date(), '.png')
ggsave(
  file = fname, 
  plot = via_all_measures + theme(plot.background = element_rect(fill = "white", linetype = "blank"),
                             panel.border = element_blank()),
  width = 6.5, height = 5.3, units = "in"
)
```

# Boxplots Escape

```{r}
nested_comb_data_long_esc_eff_tol = comb_data_long %>% 
  subset(dadonaite_esc != "not tested") %>% 
  pivot_wider(names_from = "dadonaite_esc", values_from = "rank") |>
  arrange(factor(Measure, levels = x_order))

nested_comb_data_long_esc_temp = nested_comb_data_long_esc_eff_tol %>% nest(data = -Measure)
nested_comb_data_long_esc_eff_tol
```


```{r}

data_stats_gramm <- nested_comb_data_long_esc_eff_tol |>
  nest(data = -Measure) %>%
  mutate(
    via_non_via = map(data, ~broom::tidy(wilcox.test(.x$escape, .x$`non-escape`)))
    
  ) |>
  pivot_longer(
    cols = c(via_non_via),
    names_to = "test",
    values_to = "stats"
  ) |>
  mutate(
    column1 = case_when(
      test == "via_non_via" ~ "escape"
    ),
    column2 = case_when(
      test == "via_non_via" ~ "non-escape"
    )
  ) |>
  unnest_wider(stats) |>
  select(Measure, column1, column2, p.value)

y_delta = 0#.25 # distance between adjacent bloxplots (.8/number of boxplots)
x_delta = 0#.185 # distance between line segments
data_comparisons_gramm <- data_stats_gramm |>
  mutate(
    # apply overall false-discovery rate correction
    p.value = p.adjust(p.value, method = "fdr"),
    # set stars
    label = case_when(
      p.value < 0.001 ~ "***",
      p.value < 0.01 ~ "**",
      p.value < 0.05 ~ "*",
      TRUE ~ "NS"
    ),
    # offset to correct for stars not being vertically centered
    label_off = case_when(
      label == "NS" ~ 0,
      TRUE ~ -.04,
    ),
    # calculate x and y offsets for line segments
    yoff1 = case_when(
      column1 == "viable" ~ y_delta,
      column1 == "not viable" ~ -y_delta,
      TRUE ~ 0
    ),
    yoff2 = case_when(
      column2 == "viable" ~ y_delta,
      column2 == "not viable" ~ -y_delta,
      TRUE ~ 0
    ),
    xoff = case_when(
      column1 == "escape" & column2 == "non-escape" ~ 0,
      column1 == "non-escape" & column2 == "escape" ~ x_delta,
      TRUE ~ 2 * x_delta
    )
  )

data_comparisons_via_gramm = data_comparisons_gramm %>% 
  subset(!(Measure %in% c("Semantic change (Hie)", 
                          "Semantic change (ESM2)",
                          "EVEscape",
                          "PyRosetta",
                          "FoldX"))) %>% arrange(factor(Measure, levels = (x_order)))
data_comparisons_via_sem = data_comparisons_gramm %>%
  subset(Measure %in% c("Semantic change (Hie)", 
                          "Semantic change (ESM2)",
         "EVEscape"))

nested_comb_data_long_via_gramm = nested_comb_data_long_esc_temp %>%
  subset(!(Measure %in% c("Semantic change (Hie)", 
                          "Semantic change (ESM2)",
                          "PyRosetta",
                          "FoldX"))) %>% arrange(factor(Measure, levels = (x_order)))
nested_comb_data_long_via_sem = nested_comb_data_long_esc_temp %>%
  subset(Measure %in% c("Semantic change (Hie)", 
                          "Semantic change (ESM2)"))

custom_order_via = c("escape", "non-escape")

x_order_gramm = rev(c("Grammaticality (Hie)",
            "Grammaticality (ESM2, unmasked)",
            "Grammaticality (ESM2, masked)", 
            "MutComputeX",
            "MutRank",
            "Stability Oracle",
            "EVE"
         ))

nested_comb_data_long_esc_temp
nested_comb_data_long_esc_eff_tol

via_gramm_plot = nested_comb_data_long_esc_eff_tol %>% select(c(Measure, escape, `non-escape`)) %>%#select(-c(ID_HXB2,ID, any_ab_ESC, res_seq_num, ESC_via_thresh0005)) %>%
  subset(!(Measure %in% c("Semantic change (Hie)", 
                          "Semantic change (ESM2)",
                          "EVEscape",
                          "PyRosetta",
                          "FoldX"))) %>% 
  arrange(factor(Measure, levels = (x_order_gramm))) %>%
  pivot_longer(-Measure, names_to = "type", values_to = "value") %>% 
  drop_na(value) |>
  ggplot(aes(x = value, y = factor(Measure, level = x_order_gramm), fill = factor(type, rev(custom_order_via)))) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_manual(values = c("escape" = "#5D3A9B",
                                 "non-escape" = "#E66100" 
                                 ),
                      breaks = custom_order_via,
                      name = "spike\nmutation type"
                    ) +
  geom_segment( # line segments for significance
    data = data_comparisons_via_gramm,
    aes(
      x = 1.0 + xoff + .03, 
      xend = 1.0 + xoff + .03,
      y = as.numeric(factor(Measure, level = x_order_gramm)) + yoff1, 
      yend = as.numeric(factor(Measure, level = x_order_gramm)) + yoff2
      #y = as.numeric(factor(Measure)) + y_delta, yend = as.numeric(factor(Measure)) - y_delta #midpoint +/- y_delta
    ),
    inherit.aes = FALSE
  ) +
  geom_text(
    data = data_comparisons_via_gramm,
    aes(
      x = 1 + xoff +.04,
      y = as.numeric(factor(Measure, level = x_order_gramm)) + (yoff1 + yoff2) / 2 + label_off,
      label = label,
    ),
    size = 10 / .pt, # 10 point font size
    vjust = 0.5,
    hjust = 0,
    inherit.aes = FALSE
  ) +
  xlim(-.02, 1.55) +
  scale_x_continuous(name = "relative rank", breaks =c(0,.5,1)) +
  scale_y_discrete(labels = c("Grammaticality (ESM2, unmasked)" = "Grammaticality\n(ESM2, unmasked)",
                              "Grammaticality (ESM2, masked)" = "Grammaticality\n(ESM2, masked)",
                              "Grammaticality (Hie)" = "Grammaticality\n(Hie)")) +
  # scale_y_discrete(labels = c("Grammaticality (ESM2)" = "Grammaticality\n(ESM2, unmasked)",
  #                             "Grammaticality (Hie)" = "Grammaticality\n(Hie)")) +
  theme_classic() + ylab("") + theme(legend.position = "top")
via_gramm_plot
data_stats_gramm
data_comparisons_via_gramm
layer_data(via_gramm_plot)
```

Semantic change subplot for escape

```{r}
x_order_sem = rev(c("Semantic change (Hie)", "Semantic change (ESM2)", "EVEscape"))

via_sem_plot = nested_comb_data_long_esc_eff_tol %>% select(c(Measure, escape, `non-escape`)) %>% 
  subset(Measure %in% c("Semantic change (Hie)", 
                          "Semantic change (ESM2)",
                        "EVEscape"
                        )) %>% 
  pivot_longer(-Measure, names_to = "type", values_to = "value") %>% drop_na(value) |> 
  arrange(factor(Measure, levels = (x_order_sem))) %>%
  ggplot(aes(x = value, y = factor(Measure, level = x_order_sem), fill = factor(type, rev(custom_order_via)))) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_manual(values = c("escape" = "#5D3A9B",
                                 "non-escape" = "#E66100"),
                      breaks = custom_order_via,
                      name = "HA mutation type"
                    ) +
  geom_segment( # line segments for significance
    data = data_comparisons_via_sem,
    aes(
      x = 1.0 + xoff + .03, 
      xend = 1.0 + xoff + .03,
      y = as.numeric(factor(Measure, level = x_order_sem)) + yoff1, 
      yend = as.numeric(factor(Measure, level = x_order_sem)) + yoff2
    ),
    inherit.aes = FALSE
  ) +
  geom_text(
    data = data_comparisons_via_sem,
    aes(
      x = 1 + xoff +.04,
      y = as.numeric(factor(Measure, level = x_order_sem)) + (yoff1 + yoff2) / 2 + label_off,
      label = label,
    ),
    size = 10 / .pt, # 10 point font size
    vjust = 0.5,
    hjust = 0,
    inherit.aes = FALSE
  ) +
  xlim(-.04, 1.55) +
  scale_x_continuous(name = "relative rank", breaks =c(0,.5,1)) +
  scale_y_discrete(labels = c("Semantic change (ESM2)" = "Semantic\nchange\n(ESM2)",
                              "Semantic change (Hie)" = "Semantic\nchange\n(Hie)")
                   ) +
  theme_classic() + ylab("") + theme(legend.position = "none")
via_sem_plot
```

```{r}
via_sem_plot_new_labs = via_sem_plot + 
  theme(legend.position="none")+ # plot.margin = margin(r = 0, l = 0, t= 0, b = 0.1, unit = "inches"))
  scale_y_discrete(labels = c("Semantic change (ESM2)" = "Semantic\nchange\n(ESM2)",
                              "Semantic change (Hie)" = "Semantic\nchange\n(Hie)")) + 
  scale_x_continuous(name = "relative rank", breaks =c(0,.5,1), limits = c(NA, 1.2))
via_sem_plot_new_labs

via_gramm_plot_new = via_gramm_plot + 
  theme(legend.position="none", plot.margin = margin(l = 0,t = 0.1, b = 0.1, r = 0.1, unit = "inches"))+#unit(c(0, 0, 0, 0), "inches"))  + 
  #xlim(-.02,2.5) +
  scale_x_continuous(name = "relative rank", breaks =c(0,.5,1), limits = c(NA, 1.2))
via_gramm_plot_new

legend_b <- get_legend(
  via_gramm_plot_new + theme(legend.position="right",
                             legend.key.width = unit(.6, "in"),
                             legend.key.height = unit(.1, "in")) +
    guides(color = guide_legend(nrow = 3,
                                title = "HA\nmutation type"))
)

aligned <- align_plots(via_gramm_plot_new+ xlab("relative rank"), 
                       via_sem_plot_new_labs+ xlab("relative rank"), 
                       align = 'h', axis = 'l')

sem_col <- plot_grid(
  aligned[[2]],#via_sem_plot_new_labs+ xlab("relative rank"),
  legend_b,
  NULL,
  labels = c("", "", ""),
  nrow = 3,
  rel_heights = c(1, .5, .35)
)
sem_col

gram_col = plot_grid(
  aligned[[1]] ,
  NULL,
  labels = c("", ""),
  nrow = 2,
  rel_heights = c(1, .05)
)

via_all_measures = plot_grid(
  sem_col,#aligned[[1]], 
  #NULL,
  gram_col, 
  ncol = 2,
  align = 'h',
  labels = c("A", "B"),
  label_x = 0.05, label_y = 0.95,
  hjust = -0.5, vjust = -0.5,
  rel_widths = c(.75, 1))

via_all_measures

setwd("/Users/beallma/OneDrive - Emory University/Documents/PIPP/SARS_NLP/aggregate_data_vizualization")
fname = paste0('Delta_boxplots_escape_', Sys.Date(), '.png')
ggsave(
  file = fname, 
  plot = via_all_measures + theme(plot.background = element_rect(fill = "white", linetype = "blank"),
                             panel.border = element_blank()),
  width = 6.5, height = 5.3, units = "in"
)
```