---
title: "hiv_visualizations"
output: html_document
date: '2024-04-08'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(ggplot2)
library(ggridges)
library(viridis)
library(plyr)
library(tidyr)
library(reshape2)
library(dplyr)
library(cowplot)
library(ggpubr)
library(Biostrings)

library(purrr)

three_to_single <- function(three_letter) {
  aa_map <- list(
    ALA = "A",
    ARG = "R",
    ASN = "N",
    ASP = "D",
    CYS = "C",
    GLN = "Q",
    GLU = "E",
    GLY = "G",
    HIS = "H",
    ILE = "I",
    LEU = "L",
    LYS = "K",
    MET = "M",
    PHE = "F",
    PRO = "P",
    SER = "S",
    THR = "T",
    TRP = "W",
    TYR = "Y",
    VAL = "V"
    # You can add more amino acids if needed
  )
  
  return(aa_map[[three_letter]])
}

aa_alphabet = c("A","R","N","D","C",
                "Q","E","G","H","I",
                "L","K","M","F","P",
                "S","T","W","Y","V")

get_seqs = function(file_name){
  fasta_file <- file_name
  
  # Read the lines of the FASTA file
  fasta_lines <- readLines(fasta_file)
  
  # Extract the sequence information
  sequence_info <- grep("^>", fasta_lines, value = TRUE)
sequence_name <- gsub(">", "", sequence_info)
sequence <- sapply(seq_along(sequence_info), function(i) {
  start <- grep(sequence_info[i], fasta_lines)[1] + 1
  end <- ifelse(i == length(sequence_info), length(fasta_lines), grep(sequence_info[i + 1], fasta_lines)[1] - 1)
  paste(fasta_lines[start:end], collapse = "")
}, simplify = TRUE)

# Create a named vector with one entry
fasta_data <- setNames(sequence, sequence_name)
  return(fasta_data)
}
```

#HIV Experimental Data

```{r}
setwd("/Users/beallma/OneDrive - Emory University/Documents/PIPP/SARS_NLP/aggregate_data_vizualization")

hiv_seq = get_seqs("hiv/BG505_env_translated.fasta")[[1]]
threshold = 0.11 # as defined by Hie

# fname = "hiv/Haddox_supp/results/HXB2_numbering/BG505_to_HXB2.csv"
fname = "hiv/Dingens_ab_escape/BG505_to_HXB2.csv"
site_map_csv = read.delim(file = fname, header = TRUE, sep=",",quote = "\"",
                 dec = ".", fill = TRUE, comment.char = "#")

ab_PGT121_ = c('332', '334', '327', '415', '330', '325', '408', '406', '428', '323', '443', '419', '326', '471', '380')
ab_VRC01_ = c('279', '326', '369', '197', '209')
ab_101074_ = c('334', '325', '332', '327', '415', '330', '204', '441', '323')
ab_3BNC117_ = c('207', '304', '197', '318', '471', '279', '119', '308', '209', '206', '182', '120', '204', '369', '274')
ab_VRC34_ = c('88', '85', '90', '524', '512', '513', '515', '516', '229', '514', '243', '518', '629')
ab_PG9_ = c('160', '171', '169', '162')
ab_PGT151_ = c('613', '611', '512', '637', '639', '514', '647', '322a', '318', '327', '224', '629', '142')
ab_PGT145_ = c('166', '160', '121', '169', '162', '127', '124', '315', '123', '202', '197', '312', '203')
ab_10E8_ = c('671', '673', '672', '683', '680', '676', '643', '682', '609')

values <- site_map_csv$original
site_map <- setNames(site_map_csv$original, site_map_csv$new)#setNames(as.list(values), site_map_csv$new)

fname = "hiv/Dingens_ab_escape/summary_3BN-1074-pool-meanmutfracsurvive.csv_with_sig_sites_fdr0.01.csv"
pool_3BNC117_101074 = read.delim(file = fname, header = TRUE, sep=",",quote = "\"",
                 dec = ".", fill = TRUE, comment.char = "#")

threshold_pool_3BNC117_101074 = 0.0014 
#threshold_pool_3BNC117_101074 = threshold
pool_3BNC117_101074 = pool_3BNC117_101074 %>% 
  mutate(HXB2_site = site_map[as.character(site)]) %>% 
  arrange(HXB2_site,mutation) %>% relocate("HXB2_site") %>% 
  mutate(ID = paste(wildtype, site, mutation, sep = ""),
         ID_HXB2 = paste(wildtype, HXB2_site, mutation, sep = ""),
         ab_pool_3BNC117_101074 = ifelse(sig == "True", "escape", "non-escape")
         #ab_pool_3BNC117_101074 = ifelse(mutfracsurvive >= threshold_pool_3BNC117_101074, "escape", "non-escape")#,
         # ab_PGT145 = ifelse(site %in% ab_PGT145_, "escape", "non-escape"),
         # ab_PGT121 = ifelse(site %in% ab_PGT121_, "escape", "non-escape"),
         # ab_VRC01 = ifelse(site %in% ab_VRC01_, "escape", "non-escape"),
         # ab_101074 = ifelse(site %in% ab_101074_, "escape", "non-escape"),
         # ab_3BNC117 = ifelse(site %in% ab_3BNC117_, "escape", "non-escape"),
         # ab_VRC34 = ifelse(site %in% ab_VRC34_, "escape", "non-escape"),
         # ab_PG9 = ifelse(site %in% ab_PG9_, "escape", "non-escape"),
         # ab_PGT151 = ifelse(site %in% ab_PGT151_, "escape", "non-escape"),
         # ab_10E8 = ifelse(site %in% ab_10E8_, "escape", "non-escape"),
         # any_ab_ESC = rowSums(across(
         #   names(pool_3BNC117_101074 %>% select(starts_with("ab_"))),
         #   ~.x == "escape")) > 0
         )
subset(pool_3BNC117_101074, ab_pool_3BNC117_101074 == "escape")

## PGT121 ##
fname = "hiv/Dingens_ab_escape/summary_PGT121-meanmutfracsurvive.csv_with_sig_sites_fdr0.01.csv"
data_PGT121 = read.delim(file = fname, header = TRUE, sep=",",quote = "\"",
                 dec = ".", fill = TRUE, comment.char = "#")

threshold_PGT121 = 0.011 # this threshold doesnt yield the same results as reported in their paper. going to just manually take their sites.
data_PGT121 = data_PGT121 %>% 
  mutate(HXB2_site = site_map[as.character(site)]) %>% 
  arrange(HXB2_site,mutation) %>% relocate("HXB2_site") %>% 
  mutate(ID = paste(wildtype, site, mutation, sep = ""),
         ID_HXB2 = paste(wildtype, HXB2_site, mutation, sep = ""),
         ab_PGT121 = ifelse(sig == "True", "escape", "non-escape")
         )
subset(data_PGT121, ab_PGT121 == "escape")

## VRC01 ##
fname = "hiv/Dingens_ab_escape/summary_VRC01-meanmutfracsurvive.csv_with_sig_sites_fdr0.01.csv"
data_VRC01 = read.delim(file = fname, header = TRUE, sep=",",quote = "\"",
                 dec = ".", fill = TRUE, comment.char = "#")

threshold_VRC01 = 0.01 # this threshold doesnt yield the same results as reported in their paper. going to just manually take their sites.
data_VRC01 = data_VRC01 %>% 
  mutate(HXB2_site = site_map[as.character(site)]) %>% 
  arrange(HXB2_site,mutation) %>% relocate("HXB2_site") %>% 
  mutate(ID = paste(wildtype, site, mutation, sep = ""),
         ID_HXB2 = paste(wildtype, HXB2_site, mutation, sep = ""),
         ab_VRC01 = ifelse(sig == "True", "escape", "non-escape")
         )
subset(data_VRC01, ab_VRC01 == "escape")

## 101074 ##
fname = "hiv/Dingens_ab_escape/summary_101074-meanmutfracsurvive.csv_with_sig_sites_fdr0.01.csv"
data_101074 = read.delim(file = fname, header = TRUE, sep=",",quote = "\"",
                 dec = ".", fill = TRUE, comment.char = "#")

threshold_101074 = 0.018 # this threshold doesnt yield the same results as reported in their paper. going to just manually take their sites.
data_101074 = data_101074 %>% 
  mutate(HXB2_site = site_map[as.character(site)]) %>% 
  arrange(HXB2_site,mutation) %>% relocate("HXB2_site") %>% 
  mutate(ID = paste(wildtype, site, mutation, sep = ""),
         ID_HXB2 = paste(wildtype, HXB2_site, mutation, sep = ""),
         ab_101074 = ifelse(sig == "True", "escape", "non-escape")
         )
subset(data_101074, ab_101074 == "escape")

## 3BNC117 ##
fname = "hiv/Dingens_ab_escape/summary_3BNC117-meanmutfracsurvive.csv_with_sig_sites_fdr0.01.csv"
data_3BNC117 = read.delim(file = fname, header = TRUE, sep=",",quote = "\"",
                 dec = ".", fill = TRUE, comment.char = "#")

threshold_3BNC117 = 0.018 # this threshold doesnt yield the same results as reported in their paper. going to just manually take their sites.
data_3BNC117 = data_3BNC117 %>% 
  mutate(HXB2_site = site_map[as.character(site)]) %>% 
  arrange(HXB2_site,mutation) %>% relocate("HXB2_site") %>% 
  mutate(ID = paste(wildtype, site, mutation, sep = ""),
         ID_HXB2 = paste(wildtype, HXB2_site, mutation, sep = ""),
         ab_3BNC117 = ifelse(sig == "True", "escape", "non-escape")
         )
subset(data_3BNC117, ab_3BNC117 == "escape")

## VRC34 ##
fname = "hiv/Dingens_ab_escape/summary_VRC34-meanmutfracsurvive.csv_with_sig_sites_fdr0.01.csv"
data_VRC34 = read.delim(file = fname, header = TRUE, sep=",",quote = "\"",
                 dec = ".", fill = TRUE, comment.char = "#")

threshold_VRC34 = 0.083 # this threshold doesnt yield the same results as reported in their paper. going to just manually take their sites.
data_VRC34 = data_VRC34 %>% 
  mutate(HXB2_site = site_map[as.character(site)]) %>% 
  arrange(HXB2_site,mutation) %>% relocate("HXB2_site") %>% 
  mutate(ID = paste(wildtype, site, mutation, sep = ""),
         ID_HXB2 = paste(wildtype, HXB2_site, mutation, sep = ""),
         ab_VRC34 = ifelse(sig == "True", "escape", "non-escape")
         )
subset(data_VRC34, ab_VRC34 == "escape")

## PG9 ##
fname = "hiv/Dingens_ab_escape/summary_PG9-meanmutfracsurvive.csv_with_sig_sites_fdr0.01.csv"
data_PG9 = read.delim(file = fname, header = TRUE, sep=",",quote = "\"",
                 dec = ".", fill = TRUE, comment.char = "#")

threshold_PG9 = 0.083 # this threshold doesnt yield the same results as reported in their paper. going to just manually take their sites.
data_PG9 = data_PG9 %>% 
  mutate(HXB2_site = site_map[as.character(site)]) %>% 
  arrange(HXB2_site,mutation) %>% relocate("HXB2_site") %>% 
  mutate(ID = paste(wildtype, site, mutation, sep = ""),
         ID_HXB2 = paste(wildtype, HXB2_site, mutation, sep = ""),
         ab_PG9 = ifelse(sig == "True", "escape", "non-escape")
         )
subset(data_PG9, ab_PG9 == "escape")

## PGT151 ##
fname = "hiv/Dingens_ab_escape/summary_PGT151-meanmutfracsurvive.csv_with_sig_sites_fdr0.01.csv"
data_PGT151 = read.delim(file = fname, header = TRUE, sep=",",quote = "\"",
                 dec = ".", fill = TRUE, comment.char = "#")

threshold_PGT151 = 0.043 # this threshold doesnt yield the same results as reported in their paper. going to just manually take their sites.
data_PGT151 = data_PGT151 %>% 
  mutate(HXB2_site = site_map[as.character(site)]) %>% 
  arrange(HXB2_site,mutation) %>% relocate("HXB2_site") %>% 
  mutate(ID = paste(wildtype, site, mutation, sep = ""),
         ID_HXB2 = paste(wildtype, HXB2_site, mutation, sep = ""),
         ab_PGT151 = ifelse(sig == "True", "escape", "non-escape")
         )
subset(data_PGT151, ab_PGT151 == "escape")

## PGT145 ##
fname = "hiv/Dingens_ab_escape/summary_PGT145-meanmutfracsurvive.csv_with_sig_sites_fdr0.01.csv"
data_PGT145 = read.delim(file = fname, header = TRUE, sep=",",quote = "\"",
                 dec = ".", fill = TRUE, comment.char = "#")

threshold_PGT145 = 0.043 # this threshold doesnt yield the same results as reported in their paper. going to just manually take their sites.
data_PGT145 = data_PGT145 %>% 
  mutate(HXB2_site = site_map[as.character(site)]) %>% 
  arrange(HXB2_site,mutation) %>% relocate("HXB2_site") %>% 
  mutate(ID = paste(wildtype, site, mutation, sep = ""),
         ID_HXB2 = paste(wildtype, HXB2_site, mutation, sep = ""),
         ab_PGT145 = ifelse(sig == "True", "escape", "non-escape")
         )
subset(data_PGT145, ab_PGT145 == "escape")

## 10E8 ##
fname = "hiv/Dingens_ab_escape/summary_10E8-meanmutfracsurvive.csv_with_sig_sites_fdr0.01.csv"
data_10E8 = read.delim(file = fname, header = TRUE, sep=",",quote = "\"",
                 dec = ".", fill = TRUE, comment.char = "#")

threshold_10E8 = 0.0038 # this threshold doesnt yield the same results as reported in their paper. going to just manually take their sites.
data_10E8 = data_10E8 %>% 
  mutate(HXB2_site = site_map[as.character(site)]) %>% 
  arrange(HXB2_site,mutation) %>% relocate("HXB2_site") %>% 
  mutate(ID = paste(wildtype, site, mutation, sep = ""),
         ID_HXB2 = paste(wildtype, HXB2_site, mutation, sep = ""),
         ab_10E8 = ifelse(sig == "True", "escape", "non-escape")
         )
subset(data_10E8, ab_10E8 == "escape")

```

## Combine Antibody Data
```{r}
cnames = c("HXB2_site",  "site", "wildtype", "mutation", "ID", "ID_HXB2")
hiv_DMS_comb = data_10E8 %>% 
  full_join(data_PGT145, by = c("HXB2_site",  "site", "wildtype", "mutation", "ID", "ID_HXB2"),
            suffix = c("data_10E8", "data_PGT145")) %>% 
  full_join(data_101074, by = c("HXB2_site",  "site", "wildtype", "mutation", "ID", "ID_HXB2"),
            suffix = c("data_PGT145", "data_101074")) %>% 
  full_join(data_3BNC117, by = c("HXB2_site",  "site", "wildtype", "mutation", "ID", "ID_HXB2"),
            suffix = c("data_101074", "data_3BNC117")) %>% 
  full_join(data_PG9, by = c("HXB2_site",  "site", "wildtype", "mutation", "ID", "ID_HXB2"),
            suffix = c("data_3BNC117", "data_PG9")) %>% 
  full_join(data_PGT121, by = c("HXB2_site",  "site", "wildtype", "mutation", "ID", "ID_HXB2"),
            suffix = c("data_PG9", "data_PGT121")) %>% 
  full_join(data_VRC01, by = c("HXB2_site",  "site", "wildtype", "mutation", "ID", "ID_HXB2"),
            suffix = c("data_PGT121", "data_VRC01")) %>% 
  full_join(data_PGT151, by = c("HXB2_site",  "site", "wildtype", "mutation", "ID", "ID_HXB2"),
            suffix = c("data_VRC01", "data_PGT145")) %>% 
  full_join(data_VRC34, by = c("HXB2_site",  "site", "wildtype", "mutation", "ID", "ID_HXB2"),
            suffix = c("data_PGT145", "data_VRC34")) 
hiv_DMS_comb = hiv_DMS_comb %>%
  mutate(any_ab_ESC = rowSums(
    across(
      names(hiv_DMS_comb %>% select(starts_with("ab_"))),
      ~.x == "escape")) > 0
    ) %>%
  select(-(names(hiv_DMS_comb %>% select(starts_with("ab_"))))) %>%
  select(-(names(hiv_DMS_comb %>% select(starts_with("P"))))) %>%
  select(-(names(hiv_DMS_comb %>% select(starts_with("Q"))))) %>%
  select(-(names(hiv_DMS_comb %>% select(starts_with("sig"))))) %>%
  select(-(names(hiv_DMS_comb %>% select(starts_with("mutfrac")))))
hiv_DMS_comb
```

## Viability Data

```{r}
fname = "hiv/Haddox_supp/results/prefs/rescaled_BG505_avgprefs.csv"
env_prefs_scaled = read.delim(file = fname, header = TRUE, sep=",",quote = "\"",
                 dec = ".", fill = TRUE, comment.char = "")
wt_df = read.delim(file = "hiv/Haddox_supp/results/prefs/BG505_wildtypeoverlay.csv", 
                   header = TRUE, sep=",",quote = "\"", dec = ".", fill = TRUE, comment.char = "")
# ^^ shows us that the sites in the preferences are not in the HXB2 indeces

fname = "hiv/Haddox_supp/results/HXB2_numbering/BG505_to_HXB2.csv"
site_map_csv_haddox = read.delim(file = fname, header = TRUE, sep=",",quote = "\"",
                 dec = ".", fill = TRUE, comment.char = "#")

values <- site_map_csv_haddox$original
site_map_haddox <- setNames(site_map_csv_haddox$original, site_map_csv_haddox$new)#setNames(as.list(values), site_map_csv$new)

env_prefs_scaled_long = env_prefs_scaled %>% 
  full_join(wt_df, by = c("site")) %>%
  mutate(HXB2_site = site_map_haddox[as.character(site)]) %>% 
  pivot_longer(cols = c(-site, -wildtype, -HXB2_site),
               names_to = "Amino_Acid",
               values_to = "pref_scaled") %>%
  arrange(HXB2_site) %>% relocate("HXB2_site") %>%
  mutate(pref_ranked = rank(pref_scaled, na.last = "keep"),
         ID = paste(wildtype, site, Amino_Acid, sep = ""),
         ID_HXB2 = paste(wildtype, HXB2_site, Amino_Acid, sep = ""))

env_prefs_scaled_long
subset(env_prefs_scaled_long, site == "317a")
```

###Viability Threshold

```{r}
# env_prefs_scaled_long %>%
#   ggplot(aes(x =pref_ranked, y = pref_scaled)) +
#   geom_hline(yintercept = quantile(env_prefs_scaled_long$pref_scaled, 
#                                    probs = 0.01, type = 1, names = FALSE),
#              linetype = "dotted")

env_prefs_scaled_long %>%
  ggplot(aes(x =log10(pref_scaled))) +
  geom_histogram(aes(y = ..density..),
                 bins = 100,
                 color="black",
                 fill="white") +
  geom_vline(xintercept = quantile(log10(env_prefs_scaled_long$pref_scaled), 
                                   probs = 0.01, type = 1, names = FALSE),
             linetype = "dotted")

quant = round(quantile(log10(env_prefs_scaled_long$pref_scaled),
                 probs = 0.01, type = 1, names = FALSE),
              digits = 6)

ggplot(env_prefs_scaled_long, aes(x = pref_ranked, y = log10(pref_scaled))) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept = quant,
             linetype = "dotted") +
  geom_text(aes(x=max(pref_ranked), 
                label= paste0(as.character(quant), ", 99% quantile"), 
                y=quant),
            colour="black", angle=0, hjust = 1, vjust = 1.1, check_overlap = TRUE) + 
  geom_hline(yintercept = log10(0.0005),
             linetype = "dotted", color = "blue") +
  geom_text(aes(x=max(pref_ranked), 
                label= paste0(as.character(0.0005), ", stringent error rate"), 
                y=log10(0.0005)),
            colour="blue", angle=0, hjust = 1, vjust = -0.2, check_overlap = TRUE) +
  geom_hline(yintercept = log10(0.0008),
             linetype = "dotted", color = "red") +
  geom_text(aes(x=max(pref_ranked), 
                label= paste0(as.character(0.0008), ", non. synon. post-selection freq"), 
                y=log10(0.0008)),
            colour="red", angle=0, hjust = 1, vjust = -0.2, check_overlap = TRUE) +
  labs(x = "ranked preference", y = "log(preference)", title = "")

plot_ranked_thresholds = ggplot(env_prefs_scaled_long, aes(x = pref_ranked, y = log10(pref_scaled))) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept = -1,
             linetype = "dotted") +
  geom_text(aes(x= 0,#max(pref_ranked), 
                label= paste0(
                  dim(subset(env_prefs_scaled_long, log10(pref_scaled) >= -1))[1], 
                  " mutations above 0.1"), 
                y=-1),
            colour="black", angle=0, hjust = 0, vjust = 1.1, check_overlap = TRUE) + 
  geom_hline(yintercept = -3,
             linetype = "dotted") +
  geom_text(aes(x=max(pref_ranked), 
                label= paste0(
                  dim(subset(env_prefs_scaled_long, log10(pref_scaled) >= -3))[1], 
                  " mutations above 0.001"), 
                y=-3),
            colour="black", angle=0, hjust = 1, vjust = 1.1, check_overlap = TRUE) +
  labs(x = "ranked preference", y = "preference", title = "") + 
  theme_classic()
plot_ranked_thresholds

thresh = .001
env_prefs_scaled_long = env_prefs_scaled_long %>% 
  # mutate(viab_thresh0005 = case_when(pref_scaled > thresh ~ "viable",
  #                                     TRUE ~ "not viable"),
  #        viab_threshquant99 = case_when(log10(pref_scaled) > quant ~ "viable",
  #                                     TRUE ~ "not viable"),
  #        viab_thresh0008 = case_when(pref_scaled > 0.0008 ~ "viable",
  #                                     TRUE ~ "not viable"),
  #        ID = paste0(wildtype,site,Amino_Acid)) %>%
  mutate(viab_thresh001 = case_when(pref_scaled >= .001 ~ "viable",
                                     TRUE ~ "not viable"),
         
         viab_thresh1 = case_when(pref_scaled >= .1 ~ "viable",
                                     TRUE ~ "not viable"),
         ID = paste0(wildtype,site,Amino_Acid)) %>%
  mutate(HXB2_site = site_map_haddox[as.character(site)])
  # group_by(site) %>%
  # mutate(my_shannon_entropy = shannon.entropy(pref_scaled),
  #        my_eff_tolerated = exp(my_shannon_entropy),
  #   rev_rank_by_site = max(rank(pref_scaled)) + 1 - rank(pref_scaled), #lower numbers mean more preferable
  #   my_viab_eff_tol = case_when(rev_rank_by_site <= round(my_eff_tolerated) ~ "viable",
  #                                     TRUE ~ "not viable"))
env_prefs_scaled_long
site_map_haddox
env_prefs_scaled
#subset(env_prefs_scaled_long, is.na(HXB2_site))

# subset(env_prefs_scaled_long, viab_thresh0008 == "not viable")
# subset(env_prefs_scaled_long, viab_thresh0005 == "not viable")
# subset(env_prefs_scaled_long, viab_threshquant99 == "not viable")

fname = paste0('hiv_amino_acid_prefs_', Sys.Date(), '.png')
ggsave(
  file = fname, 
  plot = plot_ranked_thresholds + theme(plot.background = element_rect(fill = "white", linetype = "blank"),panel.border = element_blank()),
  width = 3.5, height = 3, units = "in"
)

```

#Hie bespoke model

```{r}
# fname = "hiv/hiv_BG505_pretrtain_w_covid_script_logit_odds_screen.log"
# df_Hie_hiv = read.delim(file = fname, header = TRUE, sep="",quote = "\"",
#                  dec = ".", fill = TRUE, comment.char = "#")
# df_Hie_hiv = df_Hie_hiv %>% dplyr::rename("Semantic.change_Hie" = "Semantic.change",
#                "Grammaticality_Hie" = "Grammaticality") %>%
#   select(-Mutations, -Sequence)
# df_Hie_hiv   

fname = "hiv/analyze_semantics_hiv_bilstm_512.txt"
df_Hie_hiv = read.delim(file = fname, header = TRUE, sep="\t",quote = "\"",
                 dec = ".", fill = TRUE, comment.char = "#")
df_Hie_hiv = df_Hie_hiv %>% dplyr::rename("Semantic.change_Hie" = "change",
               "Grammaticality_Hie" = "prob") %>%
  mutate(ID = paste0(wt, as.character(pos + 1), mut),
         Grammaticality_Hie = log10(Grammaticality_Hie),
         Semantic.change_Hie = log10(Semantic.change_Hie)
         ) %>%
  select(-is_viable, -is_escape, -pos, -wt, - mut)
 df_Hie_hiv 
```

Read in the 
# ESM-2 Data

```{r}
fname = "hiv/hiv_BG505_pretrain_screen_ESM2_logit_odds.log"
df_ESM2_hiv = read.delim(file = fname, header = TRUE, sep="",quote = "\"",
                 dec = ".", fill = TRUE, comment.char = "#")
df_ESM2_hiv = df_ESM2_hiv %>% dplyr::rename("Semantic.change_ESM2" = "Semantic.change",
               "Grammaticality_ESM2_log" = "Grammaticality") %>%
  mutate(Grammaticality_ESM2_prob = 10^Grammaticality_ESM2_log) %>%
  mutate(site = as.integer(substr(Mutations, 2, nchar(Mutations) - 1))) %>%
  select(-Mutations, -Sequence)

df_ESM2_hiv

comb_data = df_ESM2_hiv
```

Read in the New 
#ESM-2 Grammaticality with Masking Data

```{r}

fname = "hiv/ESM_probabilities_hiv_BG505.fasta.csv"
df_ESM2_hiv_mask = read.delim(file = fname, header = TRUE, sep=",",quote = "\"",
                 dec = ".", fill = TRUE, comment.char = "#")

df_ESM2_hiv_mask <- df_ESM2_hiv_mask %>%
  pivot_longer(cols = -X,
               names_to = "Amino_Acid",
               values_to = "prob_ESM2_masked")

df_Hie_hiv_temp <- df_Hie_hiv %>%
  mutate(first_letter = substr(ID, 1, 1),
         integer_col = as.numeric(substring(ID, 2, nchar(ID) - 1)),
         second_letter = substr(ID, nchar(ID), nchar(ID)),
         partial_id = paste0(integer_col, second_letter))

df_ESM2_hiv_mask <- df_ESM2_hiv_mask %>%
  mutate(partial_id = paste0(X, Amino_Acid))

df_ESM2_hiv_mask <- df_ESM2_hiv_mask %>%
  left_join(df_Hie_hiv_temp, by = 
               c("X" = "integer_col", 
                 "Amino_Acid" = "second_letter", 
                 "partial_id" = "partial_id")) %>%
  select(-Semantic.change_Hie, -Grammaticality_Hie, 
          -first_letter) %>%
  mutate(
    ID = case_when(
    is.na(ID) ~ paste0(Amino_Acid,X,Amino_Acid),
    TRUE ~ ID),
    mutant = case_when(substr(ID, nchar(ID), nchar(ID)) == substring(ID, 1, 1) ~ "wt",
                   TRUE ~ "mutant")) %>%
  select(-X, -Amino_Acid, -partial_id)
df_ESM2_hiv_mask
```

Read in and process the 
#MutComputeX data

```{r}
fname = "hiv/bg505-env-translated_EnsResNet_0222.csv"
df_mutcomp = read.delim(file = fname, header = TRUE, sep=",",quote = "\"",
                 dec = ".", fill = TRUE, comment.char = "#")
names(df_mutcomp)
head(df_mutcomp)
df_mutcomp <- df_mutcomp %>%
  rename(prAA = "mut_pred",
         pred_prob = "mut_pred_prob") %>%
  pivot_longer(cols = starts_with("pr"),
               names_to = "Amino_Acid",
               values_to = "prob_MutCompX") %>%
  select(chain_id, pos, wtAA, mut_pred, Amino_Acid, prob_MutCompX) %>%
  mutate(
    Amino_Acid = gsub("pr", "", Amino_Acid),
    Amino_Acid = sapply(Amino_Acid, three_to_single),
    wtAA = sapply(wtAA, three_to_single),
    mut_pred = sapply(mut_pred, three_to_single),
    Mutations = paste(wtAA, pos, Amino_Acid, sep = "")
  ) %>% 
  filter(chain_id == "A") %>%
  select(Mutations,prob_MutCompX)
#high probability means more like the original sequence, so we want to invert^ this to match semantic change
df_mutcomp
```

Read in and process the
#MutRank data

```{r}
fname = "hiv/bg505-env_mutrank_0524_uncertainty.csv"
df_mutrank = read.delim(file = fname, header = TRUE, sep=",",quote = "\"",
                 dec = ".", fill = TRUE, comment.char = "#")
names(df_mutrank)
head(df_mutrank)
df_mutrank <- df_mutrank %>%
  # rename(prAA = "mut_pred",
  #        pred_prob = "mut_pred_prob") %>%
  # rename(mut_pred = "pred_AA",
  #        mut_pred_prob = "pred_prob") %>%
  pivot_longer(cols = starts_with("mean"),
               names_to = "Amino_Acid",
               values_to = "prob_MutRank") %>%
  select(chain_id, res_seq_num, wt_AA, Amino_Acid, prob_MutRank) %>%
  mutate(
    Amino_Acid = gsub("mean_", "", Amino_Acid),
    Amino_Acid = sapply(Amino_Acid, three_to_single),
    wt_AA = sapply(wt_AA, three_to_single),
    #mut_pred = sapply(mut_pred, three_to_single),
    Mutations = paste(wt_AA, res_seq_num, Amino_Acid, sep = "")
  ) %>% 
  filter(chain_id == "A") %>%
  select(Mutations,prob_MutRank)
#high probability means more like the original sequence, so we want to invert^ this to match semantic change
df_mutrank
```

Read in and process the 
#StabilityOracle data

```{r}
fname = "hiv/bg505-env-translated_stability_oracle_0423_uncertainty.csv"
df_stab_ora = read.delim(file = fname, header = TRUE, sep=",",quote = "\"",
                 dec = ".", fill = TRUE, comment.char = "#")

df_stab_ora <- df_stab_ora %>%
  pivot_longer(cols = starts_with("mean_"),
               names_to = "Amino_Acid",
               values_to = "ddG_Stab_Ora") %>%
  select(chain_id, res_seq_num, wt_AA, pred_AA, Amino_Acid, ddG_Stab_Ora) %>%
  mutate(
    Amino_Acid = gsub("mean_", "", Amino_Acid),
    Amino_Acid = sapply(Amino_Acid, three_to_single),
    wt_AA = sapply(wt_AA, three_to_single),
    pred_AA = sapply(pred_AA, three_to_single),
    Mutations = paste(wt_AA, res_seq_num, Amino_Acid, sep = ""),
    wt_idx = paste0(wt_AA, res_seq_num, sep = "")
  ) %>% 
  filter(chain_id == "A") %>%
  #filter(wt_AA != Amino_Acid) %>%
 select(Mutations,ddG_Stab_Ora, wt_idx, res_seq_num)
df_stab_ora

# unique_values_df1 <- anti_join(df_stab_ora, df_Hie_hiv, by = c("Mutations" = "ID"))
anti_join(df_Hie_hiv, df_stab_ora, by = c("ID" = "Mutations"))
# anti_join(df_stab_ora, hiv_DMS_comb, by = c("Mutations" = "ID_HXB2"))
anti_join(hiv_DMS_comb, df_stab_ora, by = c("ID_HXB2" = "Mutations"))

# df_Hie_hiv
# df_stab_ora
```


# EVEscape

```{r}
setwd("/Users/beallma/Documents/GitHub/NLP_viral_escape/data/hiv/evescape")
#fname = "Wuhan(A)_single_mutation_predictions.csv"
fname = "hiv_evescape_predictions.csv"

df_eves = read.delim(file = fname, header = TRUE, sep=",",quote = "\"",
                 dec = ".", fill = TRUE, comment.char = "#") %>%
  arrange(i) %>%
  mutate(ID = paste0(wt, i, mut)) %>%
  select(ID, evescape, eve)
  #select(mutations, EVEscape)
df_eves 
```

```{r}
ggplot(df_eves, aes(x = eve, y = (evescape))) +
  #geom_line() +
  geom_point() +
  labs(x = "EVE", y = "EVEscape", title = "") + 
  theme_classic()
```


# Combine 

Note that mutations that were not tested for escape have a value of NA for the any_ab_ESC column. Mutations that were not tested for viability have values of NA in the viab_thresh001 column. These sites 
```{r}
hiv_DMS_comb_temp = hiv_DMS_comb %>% mutate(wt_idx_DMS = paste0(wildtype,HXB2_site)) # dataframe describing the DMS escape behavior of each mutation. Map between H1 and H3 site indices.
df_Hie_hiv # contains the grammaticality and semantic change measures from the Hie model
           # large gramm -> closer to wt -> large rank
           # small sem change -> close to wt -> supposed less likely escape -> smaller rank
env_prefs_scaled_long_brief = env_prefs_scaled_long %>% ungroup() %>% select(-Amino_Acid, -site, -wildtype) 
env_prefs_scaled_long_brief
                        # DMS fitness data. Large pref -> more fit -> larger rank 
                        # (here the ranks within site have been reversed so small rank is more fit. ignore)
df_stab_ora # ddG from stability Oracle. More negative -> more stable -> smaller rank, more positive -> less stable -> larger rank
df_mutcomp # large probability -> more stable -> larger rank
df_ESM2_hiv_mask # probability from masked ESM model -> larger probability -> more grammatical -> larger rank
df_ESM2_hiv
#anti_join(df_stab_ora, flu_DMS_comb, by = c('ID' = 'ID_H1'))

hiv_comb_data = hiv_DMS_comb_temp %>% select(-site) %>%
  full_join(env_prefs_scaled_long_brief %>% select(-ID,-HXB2_site), by = c('ID_HXB2' = 'ID_HXB2'))%>%
  full_join(df_Hie_hiv, by = c('ID_HXB2' = 'ID')) %>%
  full_join(df_ESM2_hiv_mask, by = c('ID_HXB2' = 'ID')) %>%
  full_join(df_stab_ora, by = c('ID_HXB2' = 'Mutations')) %>%
  full_join(df_mutcomp, by = c('ID_HXB2' = 'Mutations'))%>%
  full_join(df_mutrank, by = c('ID_HXB2' = 'Mutations'))%>%
  full_join(df_eves, by = c('ID_HXB2' = 'ID'))%>%
  full_join(df_ESM2_hiv %>% select(c(Semantic.change_ESM2, Grammaticality_ESM2_log, ID)), by = c('ID_HXB2' = 'ID')) %>%
  select(-starts_with("ab_")) %>%
  select(-starts_with("mutfracsurvive")) %>%
  select(-wt_idx, -wt_idx_DMS)

my_measures = c("Semantic.change_Hie","Grammaticality_Hie", 
                "prob_ESM2_masked", "ddG_Stab_Ora", "prob_MutCompX","prob_MutRank",
                "Semantic.change_ESM2", "Grammaticality_ESM2_log", "evescape", "eve")
# my_measures = colnames(flu_comb_data %>% 
#                          select(-c(ID_H1, H1_site, mutation,
#                                 ID_H1, wildtype, any_ab_ESC)))

hiv_comb_data
subset(hiv_comb_data, is.na(viab_thresh001)) %>% arrange(ID_HXB2)#nest(data = -ID_HXB2)
subset(hiv_comb_data, is.na(any_ab_ESC)) %>% arrange(ID_HXB2)

rev_rank_measures = c("ddG_Stab_Ora")

comb_data_norm = hiv_comb_data %>% 
  mutate_at(my_measures, function(x) rank(x, na.last = "keep")) %>%
  mutate_at(rev_rank_measures, ~max(rank(., na.last = "keep"), na.rm=T) + 1 - rank(.,na.last = "keep") ) %>%
  mutate_at(my_measures, function(x) x/max(x, na.rm = TRUE))


comb_data_long = comb_data_norm %>% 
  select(-wildtype, -mutation, -HXB2_site, -pref_scaled) %>%
  pivot_longer(cols = my_measures ,
               names_to = "Measure",
               values_to = "rank") %>%
  mutate(Measure = 
           case_when(
             Measure == "ddG_Stab_Ora" ~ "Stability Oracle",
             Measure == "prob_MutCompX" ~ "MutComputeX",
             Measure == "prob_MutRank" ~ "MutRank",
             Measure == "prob_ESM2_masked" ~ "Grammaticality (ESM2, masked)",
             Measure == "Semantic.change_Hie" ~ "Semantic change (Hie)",
             Measure == "Grammaticality_Hie" ~ "Grammaticality (Hie)",
             Measure == "Semantic.change_ESM2" ~ "Semantic change (ESM2)",
             Measure == "Grammaticality_ESM2_log" ~ "Grammaticality (ESM2, unmasked)",
             Measure == "evescape" ~ "EVEscape",
             Measure == "eve" ~ "EVE"
             )
         )

comb_data_norm
comb_data_long
subset(comb_data_norm, (any_ab_ESC))
subset(comb_data_norm, (!any_ab_ESC) & (viab_thresh001 == "viable"))
subset(comb_data_norm, (any_ab_ESC) & (viab_thresh001 == "not viable"))
subset(comb_data_norm, (any_ab_ESC) & (viab_thresh001 == "viable"))

subset(comb_data_norm, (viab_thresh001 == "not viable") & (mutant == "wt"))
subset(comb_data_norm, (viab_thresh001 == "not viable") & (mutant == "mutant"))
subset(comb_data_norm, (viab_thresh001 == "viable") & (mutant == "mutant"))
subset(comb_data_norm, (viab_thresh1 == "not viable") & (mutant == "wt"))

```

##Remove WT
```{r}
comb_data_norm_no_wt = hiv_comb_data %>% subset(mutant == "mutant") %>%
  mutate_at(my_measures, function(x) rank(x, na.last = "keep")) %>%
  mutate_at(rev_rank_measures, ~max(rank(., na.last = "keep"), na.rm=T) + 1 - rank(.,na.last = "keep") ) %>%
  mutate_at(my_measures, function(x) x/max(x, na.rm = TRUE))


comb_data_long_no_wt = comb_data_norm %>% 
  select(-wildtype, -mutation, -HXB2_site, -pref_scaled) %>%
  pivot_longer(cols = my_measures ,
               names_to = "Measure",
               values_to = "rank") %>%
  mutate(Measure = 
           case_when(
             Measure == "ddG_Stab_Ora" ~ "Stability Oracle",
             Measure == "prob_MutCompX" ~ "MutComputeX",
             Measure == "prob_MutRank" ~ "MutRank",
             Measure == "prob_ESM2_masked" ~ "Grammaticality (ESM2, masked)",
             Measure == "Semantic.change_Hie" ~ "Semantic change (Hie)",
             Measure == "Grammaticality_Hie" ~ "Grammaticality (Hie)",
             Measure == "Semantic.change_ESM2" ~ "Semantic change (ESM2)",
             Measure == "Grammaticality_ESM2_log" ~ "Grammaticality (ESM2, unmasked)",
             Measure == "evescape" ~ "EVEscape",
             Measure == "eve" ~ "EVE"
             )
         )
```

# Boxplots viability

Now I want to recreate the cov bar charts. The issue is choosing which measure of viability to use

Let's start with the coloring the plot by viability

prepare the data

```{r}
comb_data_norm
comb_data_long

comb_data_long_gramm = comb_data_long

x_order = rev(c("Grammaticality (Hie)",
            "Grammaticality (ESM2, unmasked)",
            "Grammaticality (ESM2, masked)", 
            "MutComputeX",
            "MutRank",
            "Stability Oracle",
            "EVEscape",
            "EVE",
            "Semantic change (Hie)",
            "Semantic change (ESM2)"
         ))

count_comb_hiv_esc_via_eff_tol = comb_data_long %>% 
  mutate(tested_via = case_when(
    (viab_thresh001 == "viable") ~ "viable",
    (viab_thresh001 == "not viable") ~ "not viable",
    TRUE ~ "not tested"
  )) %>%
  mutate(tested_esc = case_when(
    (any_ab_ESC) ~ "escape",
    (!any_ab_ESC) ~ "non-escape",
    TRUE ~ "not tested"
  ))
nested_comb_data_long_esc_eff_tol = count_comb_hiv_esc_via_eff_tol %>% 
  subset(tested_via != "not tested") %>% 
  subset(Measure != "FoldX") %>% 
  subset(Measure != "PyRosetta") |>
  pivot_wider(names_from = "viab_thresh001", values_from = "rank") |>
  select(-c(mutant))  %>% 
  arrange(factor(Measure, levels = x_order))
fname = paste0('hiv_boxplots_viability_tested_.001thresh_', Sys.Date(), '.png')

nested_comb_data_long_esc_temp = nested_comb_data_long_esc_eff_tol %>% nest(data = -Measure)

nested_comb_data_long_esc_eff_tol
# count_comb_hiv_esc_via_eff_tol %>% subset(ESC_via_thresh001 == "non-escape, not viable")
# count_comb_hiv_esc_via_eff_tol %>% subset(ESC_via_thresh001 == "escape, not viable")
count_comb_hiv_esc_via_eff_tol %>% subset(tested_via == "not tested")
count_comb_hiv_esc_via_eff_tol %>% subset(tested_esc == "not tested")
count_comb_hiv_esc_via_eff_tol %>% subset((tested_esc == "not tested") & (tested_via == "not tested"))
count_comb_hiv_esc_via_eff_tol %>% subset((tested_esc != "not tested") & (tested_via == "not tested"))
count_comb_hiv_esc_via_eff_tol %>% subset((tested_esc == "not tested") & (tested_via != "not tested"))
# count_comb_hiv_esc_via_eff_tol %>% subset(ESC_via_thresh001 == "escape")
# count_comb_hiv_esc_via_eff_tol %>% subset(ESC_via_thresh001 == "non-escape")

```


```{r}
data_stats_gramm <- nested_comb_data_long_esc_eff_tol |>
  nest(data = -Measure) %>%
  mutate(
    via_non_via = map(data, ~broom::tidy(wilcox.test(.x$viable, .x$`not viable`)))
    
  ) |>
  pivot_longer(
    cols = c(via_non_via),
    names_to = "test",
    values_to = "stats"
  ) |>
  mutate(
    column1 = case_when(
      test == "via_non_via" ~ "viable"
    ),
    column2 = case_when(
      test == "via_non_via" ~ "not viable"
    )
  ) |>
  unnest_wider(stats) |>
  select(Measure, column1, column2, p.value)

y_delta = 0#.25 # distance between adjacent bloxplots (.8/number of boxplots)
x_delta = 0#.185 # distance between line segments
data_comparisons_gramm <- data_stats_gramm |>
  mutate(
    # apply overall false-discovery rate correction
    p.value = p.adjust(p.value, method = "fdr"),
    # set stars
    label = case_when(
      p.value < 0.001 ~ "***",
      p.value < 0.01 ~ "**",
      p.value < 0.05 ~ "*",
      TRUE ~ "NS"
    ),
    # offset to correct for stars not being vertically centered
    label_off = case_when(
      label == "NS" ~ 0,
      TRUE ~ -.04,
    ),
    # calculate x and y offsets for line segments
    yoff1 = case_when(
      column1 == "viable" ~ y_delta,
      column1 == "not viable" ~ -y_delta,
      TRUE ~ 0
    ),
    yoff2 = case_when(
      column2 == "viable" ~ y_delta,
      column2 == "not viable" ~ -y_delta,
      TRUE ~ 0
    ),
    xoff = case_when(
      column1 == "viable" & column2 == "not viable" ~ 0,
      column1 == "not viable" & column2 == "viable" ~ x_delta,
      TRUE ~ 2 * x_delta
    )
  )

data_comparisons_via_gramm = data_comparisons_gramm %>% 
  subset(!(Measure %in% c("Semantic change (Hie)", 
                          "Semantic change (ESM2)",
                          "EVEscape",
                          "PyRosetta",
                          "FoldX"))) %>% arrange(factor(Measure, levels = (x_order)))
data_comparisons_via_sem = data_comparisons_gramm %>%
  subset(Measure %in% c("Semantic change (Hie)", 
                          "Semantic change (ESM2)",
                        "EVEscape"))

nested_comb_data_long_via_gramm = nested_comb_data_long_esc_temp %>%
  subset(!(Measure %in% c("Semantic change (Hie)", 
                          "Semantic change (ESM2)",
                          "EVEscape",
                          "PyRosetta",
                          "FoldX"))) %>% arrange(factor(Measure, levels = (x_order)))
nested_comb_data_long_via_sem = nested_comb_data_long_esc_temp %>%
  subset(Measure %in% c("Semantic change (Hie)", 
                          "Semantic change (ESM2)",
                        "EVEscape"))

custom_order_via = c("viable", "not viable")

x_order_gramm = rev(c("Grammaticality (Hie)",
            "Grammaticality (ESM2, unmasked)",
            "Grammaticality (ESM2, masked)", 
            "MutComputeX",
            "MutRank",
            "Stability Oracle",
            "EVE"
         ))

nested_comb_data_long_esc_temp
nested_comb_data_long_esc_eff_tol

via_gramm_plot = nested_comb_data_long_esc_eff_tol %>% select(c(Measure, viable, `not viable`)) %>%#select(-c(ID_HXB2,ID, any_ab_ESC, res_seq_num, ESC_via_thresh0005)) %>%
  subset(!(Measure %in% c("Semantic change (Hie)", 
                          "Semantic change (ESM2)",
                          "EVEscape",
                          "PyRosetta",
                          "FoldX"))) %>% 
  arrange(factor(Measure, levels = (x_order_gramm))) %>%
  pivot_longer(-Measure, names_to = "type", values_to = "value") %>% 
  drop_na(value) |>
  ggplot(aes(x = value, y = factor(Measure, level = x_order_gramm), fill = factor(type, rev(custom_order_via)))) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_manual(values = c("viable" = "#399c52",
                                 "not viable" = '#AF596E' #"#cd6a83", 
                                 ),
                      breaks = custom_order_via,
                      name = "env\nmutation type"
                    ) +
  geom_segment( # line segments for significance
    data = data_comparisons_via_gramm,
    aes(
      x = 1.0 + xoff + .03, 
      xend = 1.0 + xoff + .03,
      y = as.numeric(factor(Measure, level = x_order_gramm)) + yoff1, 
      yend = as.numeric(factor(Measure, level = x_order_gramm)) + yoff2
      #y = as.numeric(factor(Measure)) + y_delta, yend = as.numeric(factor(Measure)) - y_delta #midpoint +/- y_delta
    ),
    inherit.aes = FALSE
  ) +
  geom_text(
    data = data_comparisons_via_gramm,
    aes(
      x = 1 + xoff +.04,
      y = as.numeric(factor(Measure, level = x_order_gramm)) + (yoff1 + yoff2) / 2 + label_off,
      label = label,
    ),
    size = 10 / .pt, # 10 point font size
    vjust = 0.5,
    hjust = 0,
    inherit.aes = FALSE
  ) +
  xlim(-.02, 1.55) +
  scale_x_continuous(name = "relative rank", breaks =c(0,.5,1)) +
  scale_y_discrete(labels = c("Grammaticality (ESM2, unmasked)" = "Grammaticality\n(ESM2, unmasked)",
                              "Grammaticality (ESM2, masked)" = "Grammaticality\n(ESM2, masked)",
                              "Grammaticality (Hie)" = "Grammaticality\n(Hie)")) +
  # scale_y_discrete(labels = c("Grammaticality (ESM2)" = "Grammaticality\n(ESM2, unmasked)",
  #                             "Grammaticality (Hie)" = "Grammaticality\n(Hie)")) +
  theme_classic() + ylab("") + theme(legend.position = "top")
via_gramm_plot
data_stats_gramm
data_comparisons_via_gramm
layer_data(via_gramm_plot)
```

Semantic change subplot for viability

```{r}
x_order_sem = rev(c("Semantic change (Hie)", "Semantic change (ESM2)", "EVEscape"))

via_sem_plot = nested_comb_data_long_esc_eff_tol %>% select(c(Measure, viable, `not viable`)) %>% 
  subset(Measure %in% c("Semantic change (Hie)", 
                          "Semantic change (ESM2)",
                        "EVEscape"
                        )) %>% 
  pivot_longer(-Measure, names_to = "type", values_to = "value") %>% drop_na(value) |> 
  arrange(factor(Measure, levels = (x_order_sem))) %>%
  ggplot(aes(x = value, y = factor(Measure, level = x_order_sem), fill = factor(type, rev(custom_order_via)))) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_manual(values = c("viable" = "#399c52",
                                 "not viable" = '#AF596E', #"#cd6a83", 
                                 "not tested" = "#C5C5C5"),
                      breaks = custom_order_via,
                      name = "env mutation type"
                    ) +
  geom_segment( # line segments for significance
    data = data_comparisons_via_sem,
    aes(
      x = 1.0 + xoff + .03, 
      xend = 1.0 + xoff + .03,
      y = as.numeric(factor(Measure, level = x_order_sem)) + yoff1, 
      yend = as.numeric(factor(Measure, level = x_order_sem)) + yoff2
    ),
    inherit.aes = FALSE
  ) +
  geom_text(
    data = data_comparisons_via_sem,
    aes(
      x = 1 + xoff +.04,
      y = as.numeric(factor(Measure, level = x_order_sem)) + (yoff1 + yoff2) / 2 + label_off,
      label = label,
    ),
    size = 10 / .pt, # 10 point font size
    vjust = 0.5,
    hjust = 0,
    inherit.aes = FALSE
  ) +
  xlim(-.04, 1.55) +
  scale_x_continuous(name = "relative rank", breaks =c(0,.5,1)) +
  scale_y_discrete(labels = c("Semantic change (ESM2)" = "Semantic\nchange\n(ESM2)",
                              "Semantic change (Hie)" = "Semantic\nchange\n(Hie)")
                   ) +
  theme_classic() + ylab("") + theme(legend.position = "none")
via_sem_plot
```
```{r}
via_sem_plot_new_labs = via_sem_plot + 
  theme(legend.position="none")+ # plot.margin = margin(r = 0, l = 0, t= 0, b = 0.1, unit = "inches"))
  scale_y_discrete(labels = c("Semantic change (ESM2)" = "Semantic\nchange\n(ESM2)",
                              "Semantic change (Hie)" = "Semantic\nchange\n(Hie)")) + 
  scale_x_continuous(name = "relative rank", breaks =c(0,.5,1), limits = c(NA, 1.2))
via_sem_plot_new_labs

via_gramm_plot_new = via_gramm_plot + 
  theme(legend.position="none", plot.margin = margin(l = 0,t = 0.1, b = 0.1, r = 0.1, unit = "inches"))+#unit(c(0, 0, 0, 0), "inches"))  + 
  #xlim(-.02,2.5) +
  scale_x_continuous(name = "relative rank", breaks =c(0,.5,1), limits = c(NA, 1.2))
via_gramm_plot_new

legend_b <- get_legend(
  via_gramm_plot_new + theme(legend.position="right",
                             legend.key.width = unit(.6, "in"),
                             legend.key.height = unit(.1, "in")) +
    guides(color = guide_legend(nrow = 3,
                                title = "env\nmutation type"))
)

aligned <- align_plots(via_gramm_plot_new+ xlab("relative rank"), 
                       via_sem_plot_new_labs+ xlab("relative rank"), 
                       align = 'h', axis = 'l')

sem_col <- plot_grid(
  aligned[[2]],#via_sem_plot_new_labs+ xlab("relative rank"),
  legend_b,
  NULL,
  labels = c("", "", ""),
  nrow = 3,
  rel_heights = c(1, .5, .35)
)
sem_col

gram_col = plot_grid(
  aligned[[1]] ,
  NULL,
  labels = c("", ""),
  nrow = 2,
  rel_heights = c(1, .05)
)

via_all_measures = plot_grid(
  gram_col,#aligned[[1]], 
  #NULL,
  sem_col, 
  ncol = 2,
  align = 'h',
  labels = c("A", "B"),
  label_x = 0.05, label_y = 0.95,
  hjust = -0.5, vjust = -0.5,
  rel_widths = c(1, .75))

via_all_measures

#fname = paste0('hiv_boxplots_viability_.0008thresh_', Sys.Date(), '.png')
ggsave(
  file = fname, 
  plot = via_all_measures + theme(plot.background = element_rect(fill = "white", linetype = "blank"),
                             panel.border = element_blank()),
  width = 6.5, height = 5.3, units = "in"
)
```

# Boxplots Escape

```{r}
nested_comb_data_long_esc_eff_tol = count_comb_hiv_esc_via_eff_tol %>% 
  subset(tested_esc != "not tested") %>% 
  subset(Measure != "FoldX") %>% 
  subset(Measure != "PyRosetta") |>
  #subset(mutant != "wt") %>% 
  # mutate(any_ab_ESC = case_when(any_ab_ESC ~ "escape",
  #                                     FALSE ~ "non-escape",
  #                               TRUE ~ "not tested")
  #        ) %>%
  pivot_wider(names_from = "tested_esc", values_from = "rank") |>
  select(-c(mutant))  %>% 
  arrange(factor(Measure, levels = x_order))
fname = paste0('hiv_boxplots_escape_tested_dms_tools_', Sys.Date(), '.png')



nested_comb_data_long_esc_temp = nested_comb_data_long_esc_eff_tol %>% nest(data = -Measure)
nested_comb_data_long_esc_eff_tol
```


```{r}
# nested_comb_data_long_gramm = comb_data_long_gramm |>
#   pivot_wider(names_from = "viability", values_from = "rank") |>
#   select(-c(mutant, Starr_esc, Baum_esc)) %>%
#   subset(!(Measure %in% c(
#                           "PyRosetta",
#                           "FoldX")))

data_stats_gramm <- nested_comb_data_long_esc_eff_tol |>
  nest(data = -Measure) %>%
  mutate(
    via_non_via = map(data, ~broom::tidy(wilcox.test(.x$escape, .x$`non-escape`)))
    
  ) |>
  pivot_longer(
    cols = c(via_non_via),
    names_to = "test",
    values_to = "stats"
  ) |>
  mutate(
    column1 = case_when(
      test == "via_non_via" ~ "escape"
    ),
    column2 = case_when(
      test == "via_non_via" ~ "non-escape"
    )
  ) |>
  unnest_wider(stats) |>
  select(Measure, column1, column2, p.value)

y_delta = 0#.25 # distance between adjacent bloxplots (.8/number of boxplots)
x_delta = 0#.185 # distance between line segments
data_comparisons_gramm <- data_stats_gramm |>
  mutate(
    # apply overall false-discovery rate correction
    p.value = p.adjust(p.value, method = "fdr"),
    # set stars
    label = case_when(
      p.value < 0.001 ~ "***",
      p.value < 0.01 ~ "**",
      p.value < 0.05 ~ "*",
      TRUE ~ "NS"
    ),
    # offset to correct for stars not being vertically centered
    label_off = case_when(
      label == "NS" ~ 0,
      TRUE ~ -.04,
    ),
    # calculate x and y offsets for line segments
    yoff1 = case_when(
      column1 == "viable" ~ y_delta,
      column1 == "not viable" ~ -y_delta,
      TRUE ~ 0
    ),
    yoff2 = case_when(
      column2 == "viable" ~ y_delta,
      column2 == "not viable" ~ -y_delta,
      TRUE ~ 0
    ),
    xoff = case_when(
      column1 == "escape" & column2 == "non-escape" ~ 0,
      column1 == "non-escape" & column2 == "escape" ~ x_delta,
      TRUE ~ 2 * x_delta
    )
  )

data_comparisons_via_gramm = data_comparisons_gramm %>% 
  subset(!(Measure %in% c("Semantic change (Hie)", 
                          "Semantic change (ESM2)",
                          "EVEscape",
                          "PyRosetta",
                          "FoldX"))) %>% arrange(factor(Measure, levels = (x_order)))
data_comparisons_via_sem = data_comparisons_gramm %>%
  subset(Measure %in% c("Semantic change (Hie)", 
                          "Semantic change (ESM2)",
                        "EVEscape"))

nested_comb_data_long_via_gramm = nested_comb_data_long_esc_temp %>%
  subset(!(Measure %in% c("Semantic change (Hie)", 
                          "Semantic change (ESM2)",
                          "EVEscape",
                          "PyRosetta",
                          "FoldX"))) %>% arrange(factor(Measure, levels = (x_order)))
nested_comb_data_long_via_sem = nested_comb_data_long_esc_temp %>%
  subset(Measure %in% c("Semantic change (Hie)", 
                          "Semantic change (ESM2)",
                        "EVEscape"))

custom_order_via = c("escape", "non-escape")

x_order_gramm = rev(c("Grammaticality (Hie)",
            "Grammaticality (ESM2, unmasked)",
            "Grammaticality (ESM2, masked)", 
            "MutComputeX",
            "MutRank",
            "Stability Oracle",
            "EVE"
         ))

nested_comb_data_long_esc_temp
nested_comb_data_long_esc_eff_tol

via_gramm_plot = nested_comb_data_long_esc_eff_tol %>% select(c(Measure, escape, `non-escape`)) %>%#select(-c(ID_HXB2,ID, any_ab_ESC, res_seq_num, ESC_via_thresh0005)) %>%
  subset(!(Measure %in% c("Semantic change (Hie)", 
                          "Semantic change (ESM2)",
                          "EVEscape",
                          "PyRosetta",
                          "FoldX"))) %>% 
  arrange(factor(Measure, levels = (x_order_gramm))) %>%
  pivot_longer(-Measure, names_to = "type", values_to = "value") %>% 
  drop_na(value) |>
  ggplot(aes(x = value, y = factor(Measure, level = x_order_gramm), fill = factor(type, rev(custom_order_via)))) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_manual(values = c("escape" = "#5D3A9B",
                                 "non-escape" = "#E66100" 
                                 ),
                      breaks = custom_order_via,
                      name = "env\nmutation type"
                    ) +
  geom_segment( # line segments for significance
    data = data_comparisons_via_gramm,
    aes(
      x = 1.0 + xoff + .03, 
      xend = 1.0 + xoff + .03,
      y = as.numeric(factor(Measure, level = x_order_gramm)) + yoff1, 
      yend = as.numeric(factor(Measure, level = x_order_gramm)) + yoff2
      #y = as.numeric(factor(Measure)) + y_delta, yend = as.numeric(factor(Measure)) - y_delta #midpoint +/- y_delta
    ),
    inherit.aes = FALSE
  ) +
  geom_text(
    data = data_comparisons_via_gramm,
    aes(
      x = 1 + xoff +.04,
      y = as.numeric(factor(Measure, level = x_order_gramm)) + (yoff1 + yoff2) / 2 + label_off,
      label = label,
    ),
    size = 10 / .pt, # 10 point font size
    vjust = 0.5,
    hjust = 0,
    inherit.aes = FALSE
  ) +
  xlim(-.02, 1.55) +
  scale_x_continuous(name = "relative rank", breaks =c(0,.5,1)) +
  scale_y_discrete(labels = c("Grammaticality (ESM2, unmasked)" = "Grammaticality\n(ESM2, unmasked)",
                              "Grammaticality (ESM2, masked)" = "Grammaticality\n(ESM2, masked)",
                              "Grammaticality (Hie)" = "Grammaticality\n(Hie)")) +
  # scale_y_discrete(labels = c("Grammaticality (ESM2)" = "Grammaticality\n(ESM2, unmasked)",
  #                             "Grammaticality (Hie)" = "Grammaticality\n(Hie)")) +
  theme_classic() + ylab("") + theme(legend.position = "top")
via_gramm_plot
data_stats_gramm
data_comparisons_via_gramm
layer_data(via_gramm_plot)
```

Semantic change subplot for escape

```{r}
x_order_sem = rev(c("Semantic change (Hie)", "Semantic change (ESM2)", "EVEscape"))

via_sem_plot = nested_comb_data_long_esc_eff_tol %>% select(c(Measure, escape, `non-escape`)) %>% 
  subset(Measure %in% c("Semantic change (Hie)", 
                          "Semantic change (ESM2)",
                        "EVEscape"
                        )) %>% 
  pivot_longer(-Measure, names_to = "type", values_to = "value") %>% drop_na(value) |> 
  arrange(factor(Measure, levels = (x_order_sem))) %>%
  ggplot(aes(x = value, y = factor(Measure, level = x_order_sem), fill = factor(type, rev(custom_order_via)))) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_manual(values = c("escape" = "#5D3A9B",
                                 "non-escape" = "#E66100"),
                      breaks = custom_order_via,
                      name = "env mutation type"
                    ) +
  geom_segment( # line segments for significance
    data = data_comparisons_via_sem,
    aes(
      x = 1.0 + xoff + .03, 
      xend = 1.0 + xoff + .03,
      y = as.numeric(factor(Measure, level = x_order_sem)) + yoff1, 
      yend = as.numeric(factor(Measure, level = x_order_sem)) + yoff2
    ),
    inherit.aes = FALSE
  ) +
  geom_text(
    data = data_comparisons_via_sem,
    aes(
      x = 1 + xoff +.04,
      y = as.numeric(factor(Measure, level = x_order_sem)) + (yoff1 + yoff2) / 2 + label_off,
      label = label,
    ),
    size = 10 / .pt, # 10 point font size
    vjust = 0.5,
    hjust = 0,
    inherit.aes = FALSE
  ) +
  xlim(-.04, 1.55) +
  scale_x_continuous(name = "relative rank", breaks =c(0,.5,1)) +
  scale_y_discrete(labels = c("Semantic change (ESM2)" = "Semantic\nchange\n(ESM2)",
                              "Semantic change (Hie)" = "Semantic\nchange\n(Hie)")
                   ) +
  theme_classic() + ylab("") + theme(legend.position = "none")
via_sem_plot
```

```{r}
via_sem_plot_new_labs = via_sem_plot + 
  theme(legend.position="none")+ # plot.margin = margin(r = 0, l = 0, t= 0, b = 0.1, unit = "inches"))
  scale_y_discrete(labels = c("Semantic change (ESM2)" = "Semantic\nchange\n(ESM2)",
                              "Semantic change (Hie)" = "Semantic\nchange\n(Hie)")) + 
  scale_x_continuous(name = "relative rank", breaks =c(0,.5,1), limits = c(NA, 1.2))
via_sem_plot_new_labs

via_gramm_plot_new = via_gramm_plot + 
  theme(legend.position="none", plot.margin = margin(l = 0,t = 0.1, b = 0.1, r = 0.1, unit = "inches"))+#unit(c(0, 0, 0, 0), "inches"))  + 
  #xlim(-.02,2.5) +
  scale_x_continuous(name = "relative rank", breaks =c(0,.5,1), limits = c(NA, 1.2))
via_gramm_plot_new

legend_b <- get_legend(
  via_gramm_plot_new + theme(legend.position="right",
                             legend.key.width = unit(.6, "in"),
                             legend.key.height = unit(.1, "in")) +
    guides(color = guide_legend(nrow = 3,
                                title = "env\nmutation type"))
)

aligned <- align_plots(via_gramm_plot_new+ xlab("relative rank"), 
                       via_sem_plot_new_labs+ xlab("relative rank"), 
                       align = 'h', axis = 'l')

sem_col <- plot_grid(
  aligned[[2]],#via_sem_plot_new_labs+ xlab("relative rank"),
  legend_b,
  NULL,
  labels = c("", "", ""),
  nrow = 3,
  rel_heights = c(1, .5, .35)
)
sem_col

gram_col = plot_grid(
  aligned[[1]] ,
  NULL,
  labels = c("", ""),
  nrow = 2,
  rel_heights = c(1, .05)
)

via_all_measures = plot_grid(
  sem_col,#aligned[[1]], 
  #NULL,
  gram_col, 
  ncol = 2,
  align = 'h',
  labels = c("A", "B"),
  label_x = 0.05, label_y = 0.95,
  hjust = -0.5, vjust = -0.5,
  rel_widths = c(.75, 1))

via_all_measures

#fname = paste0('hiv_boxplots_viability_.0008thresh_', Sys.Date(), '.png')
ggsave(
  file = fname, 
  plot = via_all_measures + theme(plot.background = element_rect(fill = "white", linetype = "blank"),
                             panel.border = element_blank()),
  width = 6.5, height = 5.3, units = "in"
)
```
#Pie Chart

```{r}
library(ggforce)

count_comb_hiv = comb_data_norm %>% 
  mutate(tested_via = case_when(
    (viab_thresh001 == "viable") ~ "viable",
    (viab_thresh001 == "not viable") ~ "not viable",
    TRUE ~ "not tested"
  )) %>%
  mutate(tested_esc = case_when(
    (any_ab_ESC) ~ "escape",
    (!any_ab_ESC) ~ "non-escape",
    TRUE ~ "not tested"
  )) %>%
  mutate(tested_esc_via = case_when(
    (any_ab_ESC) & (tested_via == "viable") ~ "escape",
    (!any_ab_ESC) & (tested_via == "viable") ~ "non-escape",
    (any_ab_ESC) & (tested_via == "not viable") ~ "not viable",
    (!any_ab_ESC) & (tested_via == "not viable") ~ "not viable",
    TRUE ~ "not tested"
  )) %>% 
  group_by(tested_esc) %>%
  summarize(
    count = n()
  )

mut_order = c("escape", "non-escape", 
              "not viable","not tested")

pie_data_hiv <- count_comb_hiv %>%
  arrange(match(tested_esc, mut_order)) %>% # sort so pie slices end up sorted
  mutate(
    end_angle = 2*pi*cumsum(count)/sum(count),   # ending angle for each pie slice
    start_angle = lag(end_angle, default = 0),   # starting angle for each pie slice
    mid_angle = 0.5*(start_angle + end_angle),   # middle of each pie slice, for text labels
    # horizontal and vertical justifications for outer labels
    hjust = ifelse(mid_angle > pi, 1, 0),
    vjust = ifelse(mid_angle < pi/2 | mid_angle > 3*pi/2, 0, 1)
  )
pie_data_hiv
pie_data_hiv$hjust[2] = 0 # to get the labels to not overlap since escape makes up such a small fraction of the circle
# pie_data_hiv$vjust[4] = 1
pie_data_hiv$mid_angle[2] = pie_data_hiv$end_angle[3] / 2.5
pie_data_hiv$mid_angle[3] = pie_data_hiv$end_angle[3] / 1.05

pie_data_hiv

hiv_esc_pie = ggplot(pie_data_hiv) +
    aes(
      x0 = 0, y0 = 0,
      r0 = 0.4, r = 1,
      amount = count,
      start = start_angle, end = end_angle,
      fill = tested_esc
    ) +
  geom_arc_bar(alpha = 0.7) +
  geom_text(
    aes(x = 1.1 * sin(mid_angle), 
        y = 1.1 * cos(mid_angle), 
        label = count,
        hjust = hjust,
        vjust = vjust)#vjust)
  ) +
  scale_fill_manual(values = c("escape" = "#5D3A9B",
                                 "non-escape" = "#E66100", 
                                 "not viable" = "#AF596E",
                                 "not tested" = "#C5C5C5"),
                    name = "env\nmutation type",
                    breaks = mut_order) +
  coord_fixed() +
  xlim(-1.0, 1.0) +
  ylim(-1.1, 1.4) +
  theme_void() +
  theme(
    strip.text = element_blank(),
    strip.background = element_blank(),
    legend.key=element_rect(color="white")
  ) + 
  geom_segment(aes(
      x = 1 * sin(mid_angle), 
      xend = 1.1 * sin(mid_angle),
      y = 1 * cos(mid_angle), 
      yend = 1.1 * cos(mid_angle)
      #y = as.numeric(factor(Measure)) + y_delta, yend = as.numeric(factor(Measure)) - y_delta #midpoint +/- y_delta
    ))
hiv_esc_pie

fname = paste0('hiv_DMS_esc_piechart_', Sys.Date(), '.png')
ggsave(
  file = fname, 
  plot = hiv_esc_pie,
  width = 4, height = 3.5, units = "in"
)

mut_order = c("escape", "non-escape", 
              "not viable","not tested")
#layer_data(plot = cov_esc_pie)
```

#Scatter

##Hie Gramm and SC
```{r}
custom_esc_order = (c("escape","non-escape", "not viable","not tested"))
comb_data_scatter = hiv_comb_data %>% 
  mutate(tested_via = case_when(
    (viab_thresh001 == "viable") ~ "viable",
    (viab_thresh001 == "not viable") ~ "not viable",
    TRUE ~ "not tested"
  )) %>%
  mutate(tested_esc = case_when(
    (any_ab_ESC) ~ "escape",
    (!any_ab_ESC) ~ "non-escape",
    TRUE ~ "not tested"
  )) %>%
  mutate(tested_esc_via = case_when(
    (any_ab_ESC) & (tested_via == "viable") ~ "escape",
    (!any_ab_ESC) & (tested_via == "viable") ~ "non-escape",
    (any_ab_ESC) & (tested_via == "not viable") ~ "not viable",
    (!any_ab_ESC) & (tested_via == "not viable") ~ "not viable",
    TRUE ~ "not tested"
  )) %>%
  mutate(tested_esc_via = factor(tested_esc_via, rev(custom_esc_order))) %>%
  arrange(tested_esc_via)
  #group_by(Starr_esc) %>%
  #arrange(match(Starr_esc, rev(custom_esc_order)))

# p_hiv_scatter= ggplot() + #ggplot(subset(comb_data, Starr_esc != "not viable")) +
#   geom_point( data =comb_data_scatter %>% filter(tested_esc_via == "not tested"), 
#               aes(x = Grammaticality_Hie, y = Semantic.change_Hie, 
#                  color = tested_esc_via),
#              alpha = 0.5, size = 1.5) +
#   geom_point( data =comb_data_scatter %>% filter(tested_esc_via == "not viable"), 
#               aes(x = Grammaticality_Hie, y = Semantic.change_Hie, 
#                  color = tested_esc_via),
#              alpha = 0.5, size = 1.5) +
#   geom_point( data =comb_data_scatter %>% filter(tested_esc_via == "non-escape"), 
#               aes(x = Grammaticality_Hie, y = Semantic.change_Hie, 
#                  color = tested_esc_via),
#              alpha = 0.5, size = 1.5) +
#   geom_point( data =comb_data_scatter %>% filter(tested_esc_via == "escape"),
#               aes(x = Grammaticality_Hie, y = Semantic.change_Hie,
#                  color = tested_esc_via),
#              alpha = 0.5, size = 1.5) +
#   scale_color_manual(values = c("escape" = "#5D3A9B",
#                                  "non-escape" = "#E66100", 
#                                  "not viable" = "#AF596E",
#                                  "not tested" = "#C5C5C5")) +
#   # geom_point(data = subset(comb_data_scatter, tested_esc_via == "escape"), 
#   #            aes(x = Grammaticality_Hie, y = Semantic.change_Hie),
#   #                shape = 4, color = "black", size = 1.5, stroke = .75, alpha = .7)  +
#   theme_bw() + theme(legend.position = "none",#"top",
#                      #aspect.ratio = 1,
#                      legend.title = element_blank()
#                      ) + 
#   ylab("Semantic Change") +
#   xlab("Grammaticality")

p_hiv_scatter= ggplot() + #ggplot(subset(comb_data, Starr_esc != "not viable")) +
  geom_point( data =comb_data_scatter, 
              aes(x = Grammaticality_Hie, y = Semantic.change_Hie, 
                 color = tested_esc_via),
             alpha = 0.5, size = 1.5) +
  scale_color_manual(values = c("escape" = "#5D3A9B",
                                 "non-escape" = "#E66100", 
                                 "not viable" = "#AF596E",
                                 "not tested" = "#C5C5C5")) +
  theme_bw() + theme(legend.position = "none",#"top",
                     #aspect.ratio = 1,
                     legend.title = element_blank()
                     ) + 
  ylab("Semantic Change") +
  xlab("Grammaticality") 
p_hiv_scatter = p_hiv_scatter %>%
  ggMarginal( 
           groupColour = TRUE, groupFill = TRUE,
           alpha = 0.0)
fname = paste0('hiv_hie_sem_gramm_scatter_escape_', Sys.Date(), '.png')
ggsave(
  file = fname, 
  plot = p_hiv_scatter,
  width = 3.5, height = 3.5, units = "in",
  dpi = 600
)
  
```
##Gramm and AA preferences
```{r}
p_hiv_scatter_pref= ggplot() + #ggplot(subset(comb_data, Starr_esc != "not viable")) +
  geom_point( data =comb_data_scatter %>% filter(tested_esc == "non-escape"), 
              aes(x = Grammaticality_Hie, y = log10(pref_scaled), 
                 color = tested_esc),
             alpha = 0.5, size = 1.5) +
  geom_point( data =comb_data_scatter %>% filter(tested_esc == "escape"),
              aes(x = Grammaticality_Hie, y = log10(pref_scaled),
                 color = tested_esc),
             alpha = 0.5, size = 1.5) +
  scale_color_manual(values = c("escape" = "#5D3A9B",
                                 "non-escape" = "#E66100", 
                                 "not viable" = "#AF596E",
                                 "not tested" = "#C5C5C5")) +
  # geom_point(data = subset(comb_data_scatter, tested_esc_via == "escape"), 
  #            aes(x = Grammaticality_Hie, y = Semantic.change_Hie),
  #                shape = 4, color = "black", size = 1.5, stroke = .75, alpha = .7)  +
  theme_bw() + theme(legend.position = "none",#"top",
                     #aspect.ratio = 1,
                     legend.title = element_blank()
                     ) + 
  xlab("Grammaticality") +
  ylab("Amino Acid Preference")
p_hiv_scatter_pref
fname = paste0('hiv_gramm_aapref_scatter_escape_', Sys.Date(), '.png')
ggsave(
  file = fname, 
  plot = p_hiv_scatter_pref,
  width = 3.5, height = 3.5, units = "in",
  dpi = 600
)

```
##ESM2 unmasked and SC
```{r}
p_hiv_scatter_ESM2= ggplot() + #ggplot(subset(comb_data, Starr_esc != "not viable")) +
  geom_point( data =comb_data_scatter %>% filter(tested_esc == "not tested"), 
              aes(x = Grammaticality_ESM2_log, y = Semantic.change_ESM2, 
                 color = tested_esc),
             alpha = 0.5, size = 1.5) +
  geom_point( data =comb_data_scatter %>% filter(tested_esc == "non-escape"), 
              aes(x = Grammaticality_ESM2_log, y = Semantic.change_ESM2, 
                 color = tested_esc),
             alpha = 0.5, size = 1.5) +
  geom_point( data =comb_data_scatter %>% filter(tested_esc == "escape"),
              aes(x = Grammaticality_ESM2_log, y = Semantic.change_ESM2,
                 color = tested_esc),
             alpha = 0.5, size = 1.5) +
  scale_color_manual(values = c("escape" = "#5D3A9B",
                                 "non-escape" = "#E66100", 
                                 "not viable" = "#AF596E",
                                 "not tested" = "#C5C5C5")) +
  # geom_point(data = subset(comb_data_scatter, tested_esc_via == "escape"), 
  #            aes(x = Grammaticality_Hie, y = Semantic.change_Hie),
  #                shape = 4, color = "black", size = 1.5, stroke = .75, alpha = .7)  +
  theme_bw() + theme(legend.position = "none",#"top",
                     #aspect.ratio = 1,
                     legend.title = element_blank()
                     ) + 
  xlab("Grammaticality (ESM2, unmasked)") +
  ylab("Semantic Change")
p_hiv_scatter_ESM2
fname = paste0('hiv_gramm_ESM2unmasked_scatter_escape_', Sys.Date(), '.png')
ggsave(
  file = fname, 
  plot = p_hiv_scatter_ESM2,
  width = 3.5, height = 3.5, units = "in",
  dpi = 600
)

```

##ESM2 masked and SC
```{r}
p_hiv_scatter_ESM2masked=ggplot() + #ggplot(subset(comb_data, Starr_esc != "not viable")) +
  geom_point( data =comb_data_scatter %>% filter(tested_esc == "not tested"), 
              aes(x = log10(prob_ESM2_masked), y = Semantic.change_ESM2, 
                 color = tested_esc),
             alpha = 0.5, size = 1.5) +
  geom_point( data =comb_data_scatter %>% filter(tested_esc == "non-escape"), 
              aes(x = log10(prob_ESM2_masked), y = Semantic.change_ESM2, 
                 color = tested_esc),
             alpha = 0.5, size = 1.5) +
  geom_point( data =comb_data_scatter %>% filter(tested_esc == "escape"),
              aes(x = log10(prob_ESM2_masked), y = Semantic.change_ESM2,
                 color = tested_esc),
             alpha = 0.5, size = 1.5) +
  scale_color_manual(values = c("escape" = "#5D3A9B",
                                 "non-escape" = "#E66100", 
                                 "not viable" = "#AF596E",
                                 "not tested" = "#C5C5C5")) +
  # geom_point(data = subset(comb_data_scatter, tested_esc_via == "escape"), 
  #            aes(x = Grammaticality_Hie, y = Semantic.change_Hie),
  #                shape = 4, color = "black", size = 1.5, stroke = .75, alpha = .7)  +
  theme_bw() + theme(legend.position = "none",#"top",
                     #aspect.ratio = 1,
                     legend.title = element_blank()
                     ) + 
  xlab("Grammaticality (ESM2, masked)") +
  ylab("Semantic Change")
p_hiv_scatter_ESM2masked
fname = paste0('hiv_gramm_ESM2masked_scatter_escape_', Sys.Date(), '.png')
ggsave(
  file = fname, 
  plot = p_hiv_scatter_ESM2masked,
  width = 3.5, height = 3.5, units = "in",
  dpi = 600
)

```
##ESM2 masked and AA preferences
```{r}
p_hiv_scatter_ESM2pref=ggplot() + #ggplot(subset(comb_data, Starr_esc != "not viable")) +
  geom_point( data =comb_data_scatter %>% filter(tested_esc == "not tested"), 
              aes(x = log10(prob_ESM2_masked), y = log10(pref_scaled), 
                 color = tested_esc),
             alpha = 0.5, size = 1.5) +
  geom_point( data =comb_data_scatter %>% filter(tested_esc == "non-escape"), 
              aes(x = log10(prob_ESM2_masked), y = log10(pref_scaled), 
                 color = tested_esc),
             alpha = 0.5, size = 1.5) +
  geom_point( data =comb_data_scatter %>% filter(tested_esc == "escape"),
              aes(x = log10(prob_ESM2_masked), y = log10(pref_scaled),
                 color = tested_esc),
             alpha = 0.5, size = 1.5) +
  scale_color_manual(values = c("escape" = "#5D3A9B",
                                 "non-escape" = "#E66100", 
                                 "not viable" = "#AF596E",
                                 "not tested" = "#C5C5C5")) +
  # geom_point(data = subset(comb_data_scatter, tested_esc_via == "escape"), 
  #            aes(x = Grammaticality_Hie, y = Semantic.change_Hie),
  #                shape = 4, color = "black", size = 1.5, stroke = .75, alpha = .7)  +
  theme_bw() + theme(legend.position = "none",#"top",
                     #aspect.ratio = 1,
                     legend.title = element_blank()
                     ) + 
  xlab("Grammaticality (ESM2, masked)") +
  ylab("Amino Acid Preference")
p_hiv_scatter_ESM2pref
fname = paste0('hiv_pref_ESM2masked_scatter_escape_', Sys.Date(), '.png')
ggsave(
  file = fname, 
  plot = p_hiv_scatter_ESM2pref,
  width = 3.5, height = 3.5, units = "in",
  dpi = 600
)

```

#Pie with Scatter
```{r}
library(cowplot)
setwd("/Users/beallma/OneDrive - Emory University/Documents/PIPP/SARS_NLP/aggregate_data_vizualization")
#renv::init()
```

```{r}
#mywd = "/Users/beallma/OneDrive - Emory University/Documents/PIPP/SARS_NLP/aggregate_data_vizualization"
setwd("/Users/beallma/OneDrive - Emory University/Documents/PIPP/SARS_NLP/aggregate_data_vizualization")
hiv_esc_pie
p_hiv_scatter


prow <- plot_grid(
  p_hiv_scatter,
  hiv_esc_pie + theme(legend.position="none"),
  align = 'h',
  labels = c("A", "B"),
  hjust = -1,
  nrow = 1,
  rel_widths = c(1, 1)
)
prow

legend_b <- get_legend(
  hiv_esc_pie + 
    guides(color = guide_legend(nrow = 1)) +
    theme(legend.position = "bottom")
)
scatter_pie = plot_grid(prow, legend_b, 
          ncol = 1,
          rel_heights = c(1, .15))
scatter_pie
fname = paste0('hiv_scatter_escape_piechart_', Sys.Date(), '.png')
ggsave(
  file = fname,
  plot = scatter_pie + theme(plot.background = element_rect(fill = "white", linetype = "blank"),
                             panel.border = element_blank()),
  width = 5.3, height = 3.8, units = "in"
)
```